### 2015-09-17 outline/outline.tex
First commit - standard latex ignores and an outline of what I'm thinking

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{bm, bbm}
%\usepackage{mathtools}

\author{Matt Simpson}
\title{Outline of Ideas for BTC Transaction Frequency}
\begin{document}
\maketitle

\abstract{As Bitcoin's popularity rises, so does the number of transactions per block. Existing Bitcoin rules limit the number of transactions per block to 1 megabytes’ worth. Recent average transaction rates have been about 40 percent of that limit. But because Bitcoin blocks are discovered according to a stochastic Poisson process, some blocks may reach the limit even if the average is well below the limit. This paper models the percentage of blocks that reach the 1MB limit under various average transaction rates. In our most basic model, we derive a formula for that percentage as a function of constant transaction rates. We also estimate how that percentage changes when transaction rates exhibit minute-to-minute variability along specified, empirically defensible distributions.}

\section{The Basic Problem}
Bitcoin blocks are by design distributed according a Poisson process with a rate of discovery of one block every 10 minutes. Let $t$ index time in minutes, and $N(t)$ denote the number of blocks discovered by time $t$. Then for $t>s$, the number of blocks discovered between time $s$ and time $t$ is Poisson distributed, that is $N(t) - N(s) \sim Poi((t-s)/10)$. Let $k=1,2,\dots$ denote each of the blocks, in order. Then the amount of time between blocks is exponentially distributed. Let $\tau_k$ denote the the time when block $k$ arrives. Then $\tau_k - \tau_{k-1} \sim exp(1/10)$. This part of the problem is determined for us, but determining the how the number of transactions is distributed or, more precisely, how the amount of transaction {\it data} is distributed requires more work. Suppose the amount of transaction data in MBs is distributed according to some stochastic process $dx_t$ so that the total amount of data between time $s$ and time $t$, $t>s$, is the stochastic integral $x_t - x_s = \int_s^tdx_t$. Then the amount of transaction data that arrives in $k$'th block is given by $\int_{\tau_k}^{\tau_{k+1}}dx_t$. The quantity we are interested in is thus
\[
P\left(\int_{\tau_k}^{\tau_{k+1}}dx_t > 1\right)
\]
for some choice of $k$. If we discretize the processes so that $t=1,2,\dots$ denotes the minute, this becomes
\[
P\left(\sum_{t=\tau_{k}}^{\tau_{k+1}}dx_t > 1\right).
\]
For some distributional forms for the process $dx_t$ it will be possible to compute these quantities analytically, but in general it may be difficult. Using a Monte Carlo simulation of $N$ draws, we can approximate these probabilities by
\[
\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{\sum_{t=\tau_{k}^{(i)}}^{\tau_{k+1}^{(i)}}dx_{t}^{(i)} > 1\right\}
\]
where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function.

The only piece we need to do this is the distribution of the $dx_t$'s. We have two major options here --- we can choose prespecified distributions in order to see how different distributional assumptions affect these probabilities, or we can attempt to estimate the distribution of the transaction data. The former is easy and worth doing just to gather some idea of how different forms of uncertainty affect these probabilities. In the easiest case, we can assume that the $dx_t$'s are fixed at a mean rate of data arrival and analytically compute the probability that a block fills up, showing how this changes depending on the mean data rate. In the most complicated case, we can specify a dynamic model which includes seasonality and other forms of temporal dependence with, again, some fixed mean rate.

Ideally, we would like to fit this sort of model to the BTC transaction data. Doing so will give us more a more empirically plausible distribution with which to compute the probability that a block fills up. Depending on the model, we may be able to predict the growth of BTC transaction data as a side effect, and this may be interesting in its own right.

\section{Simulation Exercises}

Each simulation exercise starts with specifying a distribution for $dx_t$. There are three basic categories here: deterministic, i.e. fixed at the mean rate, independent, and temporally dependent. There is actually some interesting room in the deterministic class --- for example, $dx_t$ could be increasing in $t$, or at least not constant. This behavior can be captured in a more realistic stochastic environment, so we will focus on that.

There are several possibilities in the independent class. Transaction data amounts are nonnegative, so it is natural to assume that $dx_t \stackrel{iid}{\sim}\mathcal{G}(\alpha,\beta)$ where $\mathcal{G}(\alpha,\beta)$ denotes the gamma distribution with shape and rate parameters $\alpha$ and $\beta$ respectively. Similarly a log-normal distribution is natural, e.g. $\log(dx_t) \stackrel{iid}{\sim}\mathcal{N}(\mu,\sigma^2)$. The log-$T$ distribution is a useful extension of the log-normal. However, all of these distributions encode some degree of right-skewness which may not match the BTC transaction data. A useful generalization to account for all possible sorts of skew is the skew-normal or more generally skew-$T$ distribution. The skew-normal distribution has the pdf
\begin{align*}
p(x) = \frac{2}{\sigma}\phi\left(\frac{x-\mu}{\sigma}\right)\Phi\left(\frac{\alpha(x-\mu)}{\sigma}\right)
\end{align*}
where $\phi(.)$ is the standard normal pdf and $\Phi(.)$ is the standard normal cdf. The parameter $\alpha$ controls the degree and direction of skewness, while $\mu$ and $\sigma$ are respectively location and scale parameters. The definition of the skew-$T$ distribution is analogous. This is only one class of options that could be used to account for possible skewness in the distribution of BTC transaction data.

With temporally dependent data, things start to get more interesting. I think a crucial feature here is that the mean structure be easily separable from the structure of the temporal dependence, that way we can easily ask questions like ``if the mean transaction rate increases to X given this dependence structure, what happens to the probability that a given block fills up?'' So the classic decomposition of a time series into trend, seasonal, and error components is exactly what we are looking for. To wit:
\begin{align*}
log(dx_t) = T_t + S_t + E_t
\end{align*}
where $T_t$ is a deterministic trend component, $S_t$ is a seasonal component which may be stochastic, and $E_t$ is a temporally dependent error component. In a particularly simple case $T_t = \mu$ and $E_t = \rho E_{t-1} + \varepsilon_t$ where $\varepsilon_t \stackrel{iid}{\sim} \mathcal{N}(0,\sigma^2)$, i.e. an AR(1) model. Any mean zero ARMA model can be used for $E_t$, or a plethora of other things, and the error term can also be changed to something skewed skewed if necessary. I suspect the most interesting part for us is the seasonal component --- seasonal variation over the time of day, week, month, or even year may cause certain time periods to have a very high probability of filling up the block. I have a less firm grasp on what I want to try here mostly because I have less experience with modeling seasonal effects, but this does seem like payoff portion of modeling the temporal dependence.

Another payoff of seasonal effects term is, combined with the trend term, we can predict when in the future BTC will start to have trouble with blocks filling up under a given regime. Also note that we can make these predictions with {\it any} block size substituted in for the actual block size. This is probably not a fruitful exercise unless we build a model specifically with forecasting in mind though.

\section{Fitting Models}
Instead of choosing parameter values and doing the simulation exercise using the models listed in the previous section, we can estimate the models and use simulations form the posterior predictive distribution in order to do the same exercises while not just using more plausible parameter values, but also taking into account parameter uncertainty. A little more formally, suppose our parameter vector is $\bm{\theta}$ and $\bm{y}$ denotes our data. Let $\bm{y}^{pred}$ denote vector predicted observations. The posterior preidctive density for $\bm{y}^{pred}$ is given by
\begin{align*}
p(\bm{y}^{pred}|\bm{y}) = \int p(\bm{y}^{pred}|\bm{\theta})p(\bm{\theta}|\bm{y}) d\bm{\theta}
\end{align*}
where $p(y|x)$ denotes the conditional density of the random variable $y$ given the random variable $x$. Given a simulation from $p(\bm{\theta}|\bm{y})$, the posterior distribution of the model parameters, using e.g. MCMC, this density is easy to approximate simply by simulating draws of the predicted $\bm{y}^{pred}$ for each parameter value simulated. More generally, we can fix some of the parameter values and let others vary to, e.g., isolate the effect of increasing the mean of the time series.

Fitting the models should be straightforward, at least in theory. STAN is probably great for what we are trying to do, which means way less work on our end. The only wrinkles here are 1) now we have to pick prior distributions, which may be controversial, but we can easily do some robustness checks with different priors and 2) potential big data concerns.
\end{document}

### 2015-09-22 outline/outline.tex
Ran some short simulations using Brownian Motion for the transaction data process, results reported in the outline file

@@
 The only wrinkles here are 1) now we have to pick prior distributions, which may be controversial, but we can easily do some robustness checks with different priors and 2) potential big data concerns.
+
+\section{A Simple Example}
+
+In this example simulation we assume that the transaction data process $x_t$ follows a continuous time random walk with drift with parameters $\mu$ and $\sigma$, that is
+\[
+x_t - x_s \sim N((t-s)\mu, (t-s)\sigma^2)
+\]
+for $t > s$.
+So $dx_t = \mu dt + \sigma d B_t$ where $B_t$ is Brownian motion.
+Using a simulation, Table \ref{tab:bmprob} contains Monte Carlo estimates of the probability that a block fills up under a variety of choices of $\mu$ and $\sigma$.
+The current average transaction data rate per minute  is $\mu=0.04$ MBs.
+As $\sigma$ increases the probability of a block filling increases, though when $\sigma$ is close to zero an increase in $\sigma$ may cause the probability to decrease slightly.
+For most of the range, however, increasing $\sigma$ increases the probability that a block fills up.
+
+% latex table generated in R 3.2.2 by xtable 1.7-4 package
+% Tue Sep 22 16:26:38 2015
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrrrrrrr}
+  \hline
+ & $\sigma = 0$ & $\sigma = 0.001$ & $\sigma = 0.01$ & $\sigma = 0.1$ & $\sigma = 1$ & $\sigma = 10$ & $\sigma = 100$ & $\sigma = 1000$ \\ 
+  \hline
+$\mu = 0.035$ & 0.0594 & 0.0594 & 0.0597 & 0.0927 & 0.3574 & 0.4822 & 0.4979 & 0.4994 \\ 
+  $\mu = 0.04$ & 0.0836 & 0.0835 & 0.0839 & 0.1149 & 0.3624 & 0.4824 & 0.4978 & 0.5005 \\ 
+  $\mu = 0.045$ & 0.1098 & 0.1098 & 0.1103 & 0.1367 & 0.3680 & 0.4829 & 0.4977 & 0.4995 \\ 
+  $\mu = 0.05$ & 0.1366 & 0.1366 & 0.1370 & 0.1596 & 0.3739 & 0.4837 & 0.4983 & 0.5003 \\ 
+   \hline
+\end{tabular}
+\caption{Approximate probability of filling of a block under various choices of $\mu$ and $\sigma$.}
+\label{tab:bmprob}
+\end{table}
+
 \end{document}

### 2015-09-22 sims/sims.R

@@ -1,0 +1,27 @@
+library(xtable)
+
+## sample size
+n <- 1000000
+
+## rate of block discovery
+lambda <- 1/10
+
+## means and variances of lognormal distribution
+mus <- c(0.035, 0.04, 0.045, 0.05)
+sigs <- c(0,10^(1:7-4))
+
+out <- matrix(0,nrow=length(mus),ncol=length(sigs))
+colnames(out) <- paste("$\\sigma = ", sigs, "$", sep="")
+rownames(out) <- paste("$\\mu = ", mus, "$", sep="")
+for(i in 1:length(mus)){
+  for(j in 1:length(sigs)){
+    mu <- mus[i]
+    sig <- sigs[j]
+    mbs <- mu*arrive + sig*sqrt(arrive)*rnorm(n)
+    out[i,j] <- mean(mbs > 1)
+  }
+}
+
+
+xout <- xtable(out, digits=4)
+print.xtable(xout, digits=4, sanitize.text.function=function(x){x})

### 2015-09-23 postgres.R
Script to access our postgres db
Simple R script that connects to our instance of postgres

@@ -1,0 +1,37 @@
+# install.packages("devtools")
+# devtools::install_github("RcppCore/Rcpp")
+# devtools::install_github("rstats-db/DBI")
+# devtools::install_github("rstats-db/RPostgres")
+
+library(DBI)
+# Connect to our postgres database with a readonly user
+con <- dbConnect(RPostgres::Postgres(),dbname = 'toshi', 
+                 host = 'toshi.cn6zzwcfsto5.us-east-1.rds.amazonaws.com',
+                 port = 5432,
+                 user = 'readonly',
+                 password = 'password')
+dbListTables(con)
+
+#construct a query
+res <- dbSendQuery(con, "SELECT * FROM blocks WHERE height = 0")
+while(!dbHasCompleted(res)){
+  chunk <- dbFetch(res, n = 5)
+  print(nrow(chunk))
+}
+
+#fetch the results of the query
+dbFetch(res)
+dbClearResult(res)
+
+
+#let's count how many transactions are in our db
+res <- dbSendQuery(con, "select count(*) from transactions")
+while(!dbHasCompleted(res)){
+  chunk <- dbFetch(res, n = 5)
+  print(nrow(chunk))
+}
+dbFetch(res)
+dbClearResult(res)
+
+
+dbDisconnect(con)

### 2015-09-23 postgres.R
A few useful postgres functions

@@ -34,4 +34,25 @@ dbFetch(res)
 dbClearResult(res)
 
 
+#let's count how many blocks are in our db
+res <- dbSendQuery(con, "select count(*) from blocks")
+while(!dbHasCompleted(res)){
+  chunk <- dbFetch(res, n = 5)
+  print(nrow(chunk))
+}
+dbFetch(res)
+dbClearResult(res)
+
+
+#let's see the timestamp of the most recent block in our database
+res <- dbSendQuery(con, "select time from blocks order by height desc limit 1")
+while(!dbHasCompleted(res)){
+  chunk <- dbFetch(res, n = 5)
+  print(nrow(chunk))
+}
+time <- dbFetch(res)
+as.POSIXct(time$time[1], origin="1970-01-01")
+dbClearResult(res)
+
+
 dbDisconnect(con)

### 2015-09-24 postgres.R
Some blockchain stats

@@ -55,4 +55,26 @@ as.POSIXct(time$time[1], origin="1970-01-01")
 dbClearResult(res)
 
 
+#let's see the timestamp of the most recent block in our database
+res <- dbSendQuery(con, "select height, time, size, transactions_count from blocks")
+while(!dbHasCompleted(res)){
+  chunk <- dbFetch(res, n = 5)
+  print(nrow(chunk))
+}
+blockstats <- dbFetch(res)
+dbClearResult(res)
+
+# I'm sure this is horrible code...
+blockstats$elapsed_time <- blockstats$time - c(NA,blockstats$time[-length(blockstats$time)])
+# for some reason this results in some negative values for elapsed time. not sure why that might be
+# maybe some of the computers that discovered blocks had clocks that were off? by up to 2 hours?
+
+blockstats$txpermin <- blockstats$transactions_count*60/blockstats$elapsed_time
+blockstats$bytespermin <- blockstats$size*60/blockstats$elapsed_time
+
+library(Hmisc)
+describe(blockstats) 
+
+
+
 dbDisconnect(con)

### 2015-09-24 postgres.R
fixed comment

@@ -55,7 +55,7 @@ as.POSIXct(time$time[1], origin="1970-01-01")
 dbClearResult(res)
 
 
-#let's see the timestamp of the most recent block in our database
+#let's calculate some block stats
 res <- dbSendQuery(con, "select height, time, size, transactions_count from blocks")
 while(!dbHasCompleted(res)){
   chunk <- dbFetch(res, n = 5)

### 2015-09-24 postgres.R
more summary stats

@@ -68,12 +68,31 @@ dbClearResult(res)
 blockstats$elapsed_time <- blockstats$time - c(NA,blockstats$time[-length(blockstats$time)])
 # for some reason this results in some negative values for elapsed time. not sure why that might be
 # maybe some of the computers that discovered blocks had clocks that were off? by up to 2 hours?
+# might be a problem for only early in the chain, though
 
 blockstats$txpermin <- blockstats$transactions_count*60/blockstats$elapsed_time
 blockstats$bytespermin <- blockstats$size*60/blockstats$elapsed_time
 
 library(Hmisc)
-describe(blockstats) 
+describe(blockstats)
+
+
+#let's calculate some block stats for the last 1000 blocks in our db
+res <- dbSendQuery(con, "select height, time, size, transactions_count from blocks order by height desc limit 1000")
+while(!dbHasCompleted(res)){
+  chunk <- dbFetch(res, n = 5)
+  print(nrow(chunk))
+}
+blockstats1k <- dbFetch(res)
+dbClearResult(res)
+blockstats1k <-blockstats1k[order(blockstats1k$height),] # re-sorting because we selected blocks in reverse order
+blockstats1k$elapsed_time <- blockstats1k$time - c(NA,blockstats1k$time[-length(blockstats1k$time)])
+blockstats1k$txpermin <- blockstats1k$transactions_count*60/blockstats1k$elapsed_time
+blockstats1k$bytespermin <- blockstats1k$size*60/blockstats1k$elapsed_time
+
+library(Hmisc)
+describe(blockstats1k)
+plot(blockstats1k$elapsed_time, blockstats1k$size, main="Block size v. time between blocks", xlab="Seconds since prior block discovery", ylab="bytes", pch=19)
 
 
 

### 2015-09-25 postgres.R
Added a comment

@@ -64,8 +64,9 @@ while(!dbHasCompleted(res)){
 blockstats <- dbFetch(res)
 dbClearResult(res)
 
-# I'm sure this is horrible code...
-blockstats$elapsed_time <- blockstats$time - c(NA,blockstats$time[-length(blockstats$time)])
+# I'm sure this is horrible code...  
+blockstats$elapsed_time <-  blockstats$time - c(NA,blockstats$time[-length(blockstats$time)])
+## cleaner but slower: c(NA, diff(blockstats$time))
 # for some reason this results in some negative values for elapsed time. not sure why that might be
 # maybe some of the computers that discovered blocks had clocks that were off? by up to 2 hours?
 # might be a problem for only early in the chain, though

### 2015-09-29 time/time.R
Corrected mean struct of the model and ran it for the last 1000 observations

@@ -20,7 +20,7 @@ timedata$elapsed_time <- timedata$time - c(NA,timedata$time[-length(timedata$tim
 timedata$min <- timedata$time/60 ## convert to minutes
 
 
-nobs <- 500 
+nobs <- 1000 
 timedatashort <- timedata[(1000 - nobs + 1):1011,]
 lbs <- rep(0,nobs)
 for(i in 1:nobs){
@@ -37,18 +37,34 @@ timelist <- list(nobs = nobs, t = times, lb = lbs, lambda = 1/10)
 ## lambda = 1 / mean block discovery time = 1/10 minutes
 
 fit <- stan(file = 'time.stan', data = timelist, iter=1, chains=1)
-system.time(fit2 <- stan(fit = fit, data = timelist, iter=1000, chains=2))
+## initializes the model and does some basic checks
+
+system.time(fit2 <- stan(fit = fit, data = timelist, iter=2000, chains=4))
+## fits the model using Hamiltonian Monto Carlo. Took about 45 minute son my laptop.
+## note: easy to parallilize this across chains.
 
 
-fit2out <- extract(fit2)
+
+## traceplots to check for convergence, looks good
 library(MCMCpack)
-
+fit2out <- extract(fit2)
 par(mfrow=c(4,4))
 traceplot(mcmc(fit2out$sigma))
 traceplot(mcmc(fit2out$tau[,1:15]))
 
-summary(mcmc(fit2out$tau))
+## summarizes all estimates and gives some diagnostics
+## neff = effective sample size. Closer to actual samples size is better
+## Rhat is a measure of convergence, closer to 1 is better
+print(fit2, pars = c('sigma', 'tau'))
 
-#### lognormal seems to work fine, need to fit bigger model, also investigate other error
-#### distributions, e.g. gamma
+## collect and report tau estimates for blocks near negative elapsed times
+taumeans <- summary(mcmc(fit2out$tau))[[1]][,1]
+tauquants <- summary(mcmc(fit2out$tau))[[2]][,c(1,5)]
+idxs <- which(diff(times)<0)
+idxs <- c(idxs, idxs-1, idxs+1)
+idxs <- idxs[order(idxs)]
+negativetime <- cbind(times[idxs], taumeans[idxs], tauquants[idxs,])
+rownames(negativetime) <- idxs
+colnames(negativetime) <- c('t', 'tau', '2.5%', '97.5%')
 
+round(negativetime, 2)

### 2015-10-01 sims/sims.R
Added an accidentally deleted line

@@ -13,6 +13,8 @@ sigs <- c(0,10^(1:7-4))
 out <- matrix(0,nrow=length(mus),ncol=length(sigs))
 colnames(out) <- paste("$\\sigma = ", sigs, "$", sep="")
 rownames(out) <- paste("$\\mu = ", mus, "$", sep="")
+
+arrive <- rexp(n, lambda)
 for(i in 1:length(mus)){
   for(j in 1:length(sigs)){
     mu <- mus[i]

### 2015-10-02 model/modelfits.R
Stan files for two models that include arrival times with measurement error and a block size process

@@ -1,0 +1,65 @@
+## this is mostly a scratchpad right now
+
+
+library(DBI)
+library(rstan)
+
+con <- dbConnect(RPostgres::Postgres(),dbname = 'toshi', 
+                 host = 'toshi.cn6zzwcfsto5.us-east-1.rds.amazonaws.com',
+                 port = 5432,
+                 user = 'readonly',
+                 password = 'password')
+
+res <- dbSendQuery(con, "select height, time, size, transactions_count from blocks order by height desc limit 1011")
+while(!dbHasCompleted(res)){
+  chunk <- dbFetch(res, n = 5)
+  print(nrow(chunk))
+}
+
+timedata <- dbFetch(res)
+dbClearResult(res)
+timedata <- timedata[order(timedata$height),]
+timedata$elapsed_time <- timedata$time - c(NA,timedata$time[-length(timedata$time)])
+timedata$min <- timedata$time/60 ## convert to minutes
+
+
+nobs <- 1000
+timedatashort <- timedata[(1000 - nobs + 1):1011,]
+lbs <- rep(0,nobs)
+for(i in 1:nobs){
+  lbs[i] <- median(timedatashort$min[i:(i+10)])
+}
+times <- timedatashort$min[-c(1:11)]
+
+which(lbs > times) ## check that each lower bound is below what it's bounding
+
+times <- times - timedatashort$min[11]
+lbs <- lbs - timedatashort$min[11]
+mbs <- timedatashort$size[-c(1:11)]/1000000  ## convert to megabytes
+
+gamlist <- list(nobs = nobs, t = times, lb = lbs, lambda = 1/10, mb = mbs, sigpars = c(1, 1), apars=c(1, 1), bpars=c(1,1))
+## lambda = 1 / mean block discovery time = 1/10 minutes
+
+fit <- stan(file = 'lognormalgamma.stan', data = gamlist, iter=1, chains=1)
+## initializes the model and does some basic checks
+
+system.time(fit2 <- stan(fit = fit, data = gamlist, iter=200, chains=1, control = list(adapt_delta = 0.9, max_treedepth = 13)))
+
+
+gamgamlist <- list(nobs = nobs, t = times/60/24/7/52, lb = lbs/60/24/7/52, lambda = 1/10/60/24/7/52, mb = mbs, betapars = c(1, 1), apars=c(1, 1), bpars=c(1,1))
+
[…]

### 2015-10-02 outline/outline.tex
Sketched a class of models including measurement error on arrival times and block size

@@
 
-\section{The Basic Problem}
+\section{The Basic Model}
 Bitcoin blocks are by design distributed according a Poisson process with a rate of discovery of one block every 10 minutes.
-Let $t$ index time in minutes, and $N(t)$ denote the number of blocks discovered by time $t$.
-Then for $t>s$, the number of blocks discovered between time $s$ and time $t$ is Poisson distributed, that is $N(t) - N(s) \sim Poi((t-s)/10)$.
+Let $\tau$ index time in minutes, and $N(\tau)$ denote the number of blocks discovered by time $\tau$.
+Then for $\delta > 0$, the number of blocks discovered between time $\tau$ and time $\tau + \delta$ is Poisson distributed, that is $N(\tau + \delta) - N(\tau) \sim Poi(\delta/10)$.
 Let $k=1,2,\dots$ denote each of the blocks, in order.
@@
 Let $\tau_k$ denote the the time when block $k$ arrives.
-Then $\tau_k - \tau_{k-1} \sim exp(1/10)$.
-This part of the problem is determined for us, but determining the how the number of transactions is distributed or, more precisely, how the amount of transaction {\it data} is distributed requires more work.
-Suppose the amount of transaction data in MBs is distributed according to some stochastic process $dx_t$ so that the total amount of data between time $s$ and time $t$, $t>s$, is the stochastic integral $x_t - x_s = \int_s^tdx_t$.
-Then the amount of transaction data that arrives in $k$'th block is given by $\int_{\tau_k}^{\tau_{k+1}}dx_t$.
+Then $\tau_{k+1} - \tau_{k} \stackrel{iid}{\sim} exp(1/10)$ for $k=1,2,\dots,K$.
+This part of the model is determined for us, but determining the how the number of transactions is distributed or, more precisely, how the amount of transaction {\it data} is distributed requires more work.
+
+Crucial constraints on the models we can fit come from the data obtained from the block chain.
+Since Bitcoin is a distributed network without a home node, there is no official timestamp for when each block arrives on the chain.
+Instead, the block miner's local time is recorded.
+Two constraints on the reported time that the block was mined are built into the Bitcoin protocol.
+First, any block submitted to the chain with a local time less than the median local times of the last 11 blocks is rejected.
+Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node, with a maximum offset of 70 minutes.
+Because of the nature of the blocks' timestamps, occasionally a block has a timestamp which is earlier than one or more previous blocks in the chain.
+Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid timestamps define lower and upper bounds for the measurement error distribution.
+Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the upper bound is at our disposal.
+
+The second major constraint that the block chain imposes on us is that transaction data is only measured once a block is completed.
+So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error.
+These two constraints motivate a class of models for modeling Bitcoin transaction data.
+Let $\tau_k$ denote the actual time of the arriaval of the $k$'th block, $\delta_k$ the amount of time it is the current block so that $\tau_{k+1}=\tau_k + \delta_k$, and $t_k$ the observed time in the blockchain.
+Further, let $x_\tau$ denote the amount data in the chain at time $\tau$ so that $x_{\tau_{k+1}} - x_{\tau_k}$ denotes the amount of data in block $k$.
+Now suppose the amount of transaction data in MBs is distributed according to some stochastic process $dx_\tau$ so that $x_{\tau + \delta} - x_\tau = \int_\tau^{\tau + \delta}dx_\tau$, a stochastic integral.
+Then the amount of transaction data that arrives in $k$'th block is given by $\int_{\tau_k}^{\tau_{k}+\delta_k}dx_\tau$.
+
+Most of the interesting modeling choices come from modeling $dx_\tau$.
+Conventional Brownian motion is inappropriate since the increments should be nonnegative.
+On the other hand, geometric Brownian motion is a poor choice since it cannot start from zero.
+Instead, we consider the classes of non-negative infinite activity L\'{e}vy processes with closed form solutions for $x_{\tau}$.
+The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$.
+For example, the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$.
+We use the shape-rate parameterization of the gamma distribution so that if $x\sim G(alpha,\beta)$, it has the density
+\[
+p(x) = \frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-x\beta}
+\]
+for $x>0$ with mean $\alpha/\beta$ and variance $\alpha/\beta^2$.
+
+While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form.
+Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$.
+The inverse Gaussian process is another L\'{e}vy process which has a closed form solution for the distribution of $x_\tau$.
+The inverse Gaussian density is
+\[
+p(x) = \left[\frac{\lambda}{2\pi x^3}\right]^{1/2}\exp\left[\frac{-\lambda(x - \mu)^2}{2\mu^2x}\right]
+\]
+for $x>0$ with mean parameter $\mu>0$, shape parameter $\lambda>0$, and variance $\mu^3/\lambda$.
+Then increments of the inverse Gaussian process are inverse Gaussian distributed, viz.
+$x_{\tau + \delta} - x_\tau \sim IG(\mu\delta,\lambda\delta^2)$.
+
+It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error since this simplifies Markov chain Monte Carlo methods that may be used to fit the model.
+The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g.
+seasonal variation or stochastic volatility, but most reasonable models on these parameters will not allow for closed form distributions of the parameter increments.
+
+On the data level we have measurement error with a lower bound, motivating a class of models of the form 
+\[
+t_{k+1} |t_{1:k}, \tau_{1:K} \stackrel{ind}{\sim} p(t|\tau_{k+1}, \phi)\mathbbm{1}\left\{t > m_{k+1})\right\}
+\]
+where $p(t|\tau,\phi)$ is a class of probability densities on $t>0$ with centrality parameter $\tau$ and dispersion parameter $\phi$, and $m_k = \mathrm{median}(t_{k - 0:10})$.
+For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision.
+This interpretation of these parameter applies for the {\it untruncated} dsitribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$.
+
+For example, consider the following parameterized modeling using the structure discussed above. For block $k=1,2,\dots,K$
+\begin{align*}
+t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
+x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\alpha\delta_k,\beta)\\
+\delta_{k-1} = \tau_k - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
+\end{align*}
+where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$.
+Note that $x_{\tau_{k+1}} - x_{\tau_k}$ depends on $\delta_k$ because the $k$'th block arrives at $\tau_k$ and departs at $\tau_{k+1}=\tau_k + \delta_k$ --- the block has $\delta_k$ minutes to accumulate transaction data.
+
+\section{Put this where it needs to go}
+
 The quantity we are interested in is thus
 \[
-P\left(\int_{\tau_k}^{\tau_{k+1}}dx_t > 1\right)
+P\left(\int_{\tau_k}^{\tau_{k}+\delta_k}dx_\tau > 1\right)
 \]
-for some choice of $k$. If we discretize the processes so that $t=1,2,\dots$ denotes the minute, this becomes
-\[
-P\left(\sum_{t=\tau_{k}}^{\tau_{k+1}}dx_t > 1\right).
-\]
-For some distributional forms for the process $dx_t$ it will be possible to compute these quantities analytically, but in general it may be difficult.
+for some choice of $k$.
+
+For some distributional forms for the process $dx_\tau$ it may be possible to compute these quantities analytically, but in general it may be difficult.
 Using a Monte Carlo simulation of $N$ draws, we can approximate these probabilities by
 \[
-\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{\sum_{t=\tau_{k}^{(i)}}^{\tau_{k+1}^{(i)}}dx_{t}^{(i)} > 1\right\}
+\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{\int_{\tau_k^{(i)}}^{\tau_{k}^{(i)}+\delta_k^{(i)}}dx_\tau^{(i)} > 1\right\}
 \]
 where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function.
-
-The only piece we need to do this is the distribution of the $dx_t$'s.
-We have two major options here --- we can choose prespecified distributions in order to see how different distributional assumptions affect these probabilities, or we can attempt to estimate the distribution of the transaction data.
-The former is easy and worth doing just to gather some idea of how different forms of uncertainty affect these probabilities.
-In the easiest case, we can assume that the $dx_t$'s are fixed at a mean rate of data arrival and analytically compute the probability that a block fills up, showing how this changes depending on the mean data rate.
-In the most complicated case, we can specify a dynamic model which includes seasonality and other forms of temporal dependence with, again, some fixed mean rate.
-
-Ideally, we would like to fit this sort of model to the BTC transaction data.
-Doing so will give us more a more empirically plausible distribution with which to compute the probability that a block fills up.
-Depending on the model, we may be able to predict the growth of BTC transaction data as a side effect, and this may be interesting in its own right.
+This formula assumes that we can perfectly simulate the continous sample paths of $dx_\tau$, which may require further approximation.
 

### 2015-10-05 outline/outline.tex
Added some detail about simulations

@@
 Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid timestamps define lower and upper bounds for the measurement error distribution.
-Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the upper bound is at our disposal.
+Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the lower bound is at our disposal.
 
@@
 These two constraints motivate a class of models for modeling Bitcoin transaction data.
-Let $\tau_k$ denote the actual time of the arriaval of the $k$'th block, $\delta_k$ the amount of time it is the current block so that $\tau_{k+1}=\tau_k + \delta_k$, and $t_k$ the observed time in the blockchain.
+Let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time it is the current block so that $\tau_{k+1}=\tau_k + \delta_k$, and $t_k$ the observed arrival time in the blockchain.
 Further, let $x_\tau$ denote the amount data in the chain at time $\tau$ so that $x_{\tau_{k+1}} - x_{\tau_k}$ denotes the amount of data in block $k$.
@@
 The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$.
-For example, the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$.
+Brownian motion is the most common example.
+A better example for our purposes is the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$.
 We use the shape-rate parameterization of the gamma distribution so that if $x\sim G(alpha,\beta)$, it has the density
@@
 
-While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form.
+While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x\sim G(\alpha_x,\beta)$ independent of $y\sim G(alpha_y,\beta)$, then $x+y \sim G(\alpha_x + \alpha_y, \beta)$.
 Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$.
@@
 
-It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error since this simplifies Markov chain Monte Carlo methods that may be used to fit the model.
+It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error.
+This simplifies Markov chain Monte Carlo methods that may be used to fit the model by allowing us to exactly simulate $x_{\tau_k + \delta_k}$ without resorting to approximation.
 The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g.
-seasonal variation or stochastic volatility, but most reasonable models on these parameters will not allow for closed form distributions of the parameter increments.
+seasonal variation or stochastic volatility, but many reasonable models on these parameters will not allow for closed form distributions of the parameter increments.
 
@@
 For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision.
-This interpretation of these parameter applies for the {\it untruncated} dsitribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$.
+This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$.
+A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative.
+However the gamma causes computational problems since it forces the latent measurement times, $\tau_k$ to enter the gamma function.
+A similar model that avoids this problem is the lognormal distribution.
 
-For example, consider the following parameterized modeling using the structure discussed above. For block $k=1,2,\dots,K$
+To start, consider the following parameterized model using the structure discussed above. For block $k=1,2,\dots,K$
 \begin{align*}
 t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
-x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\alpha\delta_k,\beta)\\
+x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\mu\delta_k,\beta)\\
 \delta_{k-1} = \tau_k - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
@@
 
-\section{Put this where it needs to go}
-
-The quantity we are interested in is thus
+\section{The key question}
+As Bitcoin becomes more popular, the transaction data rate increases.
+Since existing Bitcoin rules limit the number of transactions per block to one MB, at some point the blocks will begin filling up.
+The purpose of modeling Bitcoin's transaction data rate is to estimate the probability of seeing a full block under a variety of assumptions.
+Starting from the base model above, the quantity we are interested in is thus
 \[
@@
 where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function.
-This formula assumes that we can perfectly simulate the continous sample paths of $dx_\tau$, which may require further approximation.
+The key parameter we will control in these simulations is the mean transaction data rate, $\mu$, or the distribution of this rate.
+Starting with the basic model fit to, say, the last 1000 blocks on the blockchain (approximately one week of elapsed time), we can control this parameter in several ways.
+
+The simplest way to compute parameter estimates based on fitting model, e.g.
+maximum likelihood estimates, posterior means, or posterior medians.
+Then estimate the probability of a block filling up by simulating the model conditional on these parameter values.
+In order to take into account increased Bitcoin usage, we can manually increase the mean transaction data rate parameter, $\mu$, while holding other parameters constant.
+
+This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\mu$ increases.
+We can mitigate 1) while making 2) if not worse, different.
+Specifically, suppose we fit the model and have access to the full (approximate) posterior distribution of the model parameters.
+Then we can simulate from the posterior of the parameters and for each simulated parameter, simulate the model in order to estimate the probability of a block filling up.
+This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\mu$'s distribution were different, e.g.
+its mean $\bar{\mu}$.
+We estimate the probability of a block filling using both approaches, but because of the required assumptions the estimates become less credible for values of $\mu$ or $\bar{\mu}$ further away from the estimated values.
 

### 2015-10-05 outline/outline.tex
Added some preliminary results

@@
 
+Table \ref{tab:probsims} contains estimates of the probability of a filled block under a variety of assumptions, using parameter estimates of the posterior distribution of the model using 1000 contiguous blocks ending with block $253,221$ from August 20, 2013 arriving at about 11:39 a.m.
+% latex table generated in R 3.2.2 by xtable 1.7-4 package
+% Mon Oct  5 14:56:54 2015
+\begin{table}[ht]
+\centering
+\begin{tabular}{llll}
+  \hline
+$\mu - \hat{\mu}$ & $P_{\bar{\mu}}$ & $P_{\mu^{(0.5)}}$ & $P_{\mathrm{post}}$ \\ 
+  \hline
+0 & 0.01 & 0.01 & 0.00 \\ 
+  0.01 & 0.03 & 0.03 & 0.03 \\ 
+  0.05 & 0.23 & 0.23 & 0.22 \\ 
+  0.1 & 0.42 & 0.42 & 0.42 \\ 
+  0.5 & 0.82 & 0.82 & 0.83 \\ 
+   \hline
+\end{tabular}
+\caption{Estimated probability of a filled block under a variety of assumptions.
+From the posterior distribution the mean is $\bar{\mu} = 0.0179$ and the median is $\mu^{(0.5)} = 0.0174$.
+The value of $\mu - \hat{\mu}$ is the amount added to the estimated value of $\mu$ for the purposes of the simulation, $P_{\bar{\mu}}$ is the probability of a filled block using posterior means for parameter estimates, $P_{\mu^{(0.5)}}$ is the probability of a filled block using posterior medians, and $P_{\mathrm{post}}$ is the probability of a filled block using the full posterior distribution.}
+\label{tab:probsims}
+\end{table}
+
 \section{Simulation Exercises}
@@
 \]
-for $t > s$.
-So $dx_t = \mu dt + \sigma d B_t$ where $B_t$ is Brownian motion.
-Using a simulation, Table \ref{tab:bmprob} contains Monte Carlo estimates of the probability that a block fills up under a variety of choices of $\mu$ and $\sigma$.
+for $t > s$. So $dx_t = \mu dt + \sigma d B_t$ where $B_t$ is Brownian motion. Using a simulation, Table %\ref{tab:bmprob}
+ contains Monte Carlo estimates of the probability that a block fills up under a variety of choices of $\mu$ and $\sigma$.
 The current average transaction data rate per minute  is $\mu=0.04$ MBs.
@@
 % Tue Sep 22 16:26:38 2015
-\begin{table}[ht]
-\centering
-\begin{tabular}{lrrrrrrrr}
-  \hline
- & $\sigma = 0$ & $\sigma = 0.001$ & $\sigma = 0.01$ & $\sigma = 0.1$ & $\sigma = 1$ & $\sigma = 10$ & $\sigma = 100$ & $\sigma = 1000$ \\ 
-  \hline
-$\mu = 0.035$ & 0.0594 & 0.0594 & 0.0597 & 0.0927 & 0.3574 & 0.4822 & 0.4979 & 0.4994 \\ 
-  $\mu = 0.04$ & 0.0836 & 0.0835 & 0.0839 & 0.1149 & 0.3624 & 0.4824 & 0.4978 & 0.5005 \\ 
-  $\mu = 0.045$ & 0.1098 & 0.1098 & 0.1103 & 0.1367 & 0.3680 & 0.4829 & 0.4977 & 0.4995 \\ 
-  $\mu = 0.05$ & 0.1366 & 0.1366 & 0.1370 & 0.1596 & 0.3739 & 0.4837 & 0.4983 & 0.5003 \\ 
-   \hline
-\end{tabular}
-\caption{Approximate probability of filling of a block under various choices of $\mu$ and $\sigma$.}
-\label{tab:bmprob}
-\end{table}
+%\begin{table}[ht]
+%\centering
+%\begin{tabular}{lrrrrrrrr}
+%  \hline
+% & $\sigma = 0$ & $\sigma = 0.001$ & $\sigma = 0.01$ & $\sigma = 0.1$ & $\sigma = 1$ & $\sigma = 10$ & $\sigma = 100$ & $\sigma = 1000$ \\ 
+%  \hline
+%$\mu = 0.035$ & 0.0594 & 0.0594 & 0.0597 & 0.0927 & 0.3574 & 0.4822 & 0.4979 & 0.4994 \\ 
+%  $\mu = 0.04$ & 0.0836 & 0.0835 & 0.0839 & 0.1149 & 0.3624 & 0.4824 & 0.4978 & 0.5005 \\ 
+%  $\mu = 0.045$ & 0.1098 & 0.1098 & 0.1103 & 0.1367 & 0.3680 & 0.4829 & 0.4977 & 0.4995 \\ 
+%  $\mu = 0.05$ & 0.1366 & 0.1366 & 0.1370 & 0.1596 & 0.3739 & 0.4837 & 0.4983 & 0.5003 \\ 
+%   \hline
+%\end{tabular}
+%\caption{Approximate probability of filling of a block under various choices of $\mu$ and $\sigma$%.}
+%\label{tab:bmprob}
+%\end{table}
 

### 2015-10-05 model/modelfits.R

@@ -1,6 +1,4 @@
 ## this is mostly a scratchpad right now
-
-
 library(DBI)
 library(rstan)
 
@@ -9,6 +7,8 @@ con <- dbConnect(RPostgres::Postgres(),dbname = 'toshi',
                  port = 5432,
                  user = 'readonly',
                  password = 'password')
+
+dbListTables(con)
 
 res <- dbSendQuery(con, "select height, time, size, transactions_count from blocks order by height desc limit 1011")
 while(!dbHasCompleted(res)){
@@ -22,6 +22,8 @@ timedata <- timedata[order(timedata$height),]
 timedata$elapsed_time <- timedata$time - c(NA,timedata$time[-length(timedata$time)])
 timedata$min <- timedata$time/60 ## convert to minutes
 
+save(timedata, file = "timedata.RData")
+load("timedata.RData")
 
 nobs <- 1000
 timedatashort <- timedata[(1000 - nobs + 1):1011,]
@@ -43,23 +45,62 @@ gamlist <- list(nobs = nobs, t = times, lb = lbs, lambda = 1/10, mb = mbs, sigpa
 fit <- stan(file = 'lognormalgamma.stan', data = gamlist, iter=1, chains=1)
 ## initializes the model and does some basic checks
 
-system.time(fit2 <- stan(fit = fit, data = gamlist, iter=200, chains=1, control = list(adapt_delta = 0.9, max_treedepth = 13)))
+system.time(fit2 <- stan(fit = fit, data = gamlist, iter=2000, chains=4, control = list(adapt_delta = 0.9, max_treedepth = 13)))
 
+save(fit2, file="loggammafit.RData")
 
-gamgamlist <- list(nobs = nobs, t = times/60/24/7/52, lb = lbs/60/24/7/52, lambda = 1/10/60/24/7/52, mb = mbs, betapars = c(1, 1), apars=c(1, 1), bpars=c(1,1))
+load("loggammafit.RData")
 
-initlist <- list(list(x = rep(1, length(times)), a = 1, b = 1, beta = 1))
+rstan::traceplot(fit2, pars=c('a', 'b', 'mu', 'sigma', 'lp__'), ncol=1)
 
[…]

### 2015-10-07 outline/outline.tex
Added some description of the seasonal effects & trend components of the model

@@
 
-While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x\sim G(\alpha_x,\beta)$ independent of $y\sim G(alpha_y,\beta)$, then $x+y \sim G(\alpha_x + \alpha_y, \beta)$.
+While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x\sim G(\alpha_x,\beta)$ independent of $y\sim G(\alpha_y,\beta)$, then $x+y \sim G(\alpha_x + \alpha_y, \beta)$.
 Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$.
@@
 t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
-x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\mu\delta_k,\beta)\\
-\delta_{k-1} = \tau_k - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
+x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\gamma\delta_k,\beta)\\
+\delta_{k} = \tau_{k} - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
 \end{align*}
@@
 Note that $x_{\tau_{k+1}} - x_{\tau_k}$ depends on $\delta_k$ because the $k$'th block arrives at $\tau_k$ and departs at $\tau_{k+1}=\tau_k + \delta_k$ --- the block has $\delta_k$ minutes to accumulate transaction data.
+
+\subsection{Stochastic transaction rates --- seasonality and autocorrelation}
+It is unlikely that the transaction rate is constant in time.
+The transaction rate is clearly growing as Bitcoin becomes more popular, and seasonal variation is expected throughout the data.
+Further, autocorrelation in transaction rates is likely since periods of high activity will sometimes be caused by, e.g., new information about the long term prospects of Bitcoin.
+
+Let $\gamma_\tau$ denote the transaction rate at time $\tau$.
+We model the evolution of this rate as a variant of the exponential Ornstein–Uhlenbeck (OU) process.
+Specifically let $\varepsilon_t = \log(\gamma_t) - \mu_t$ , and let $\mu_t$ define a deterministic function of time containing seasonal and trend components.
+Then $\varepsilon_{\tau}$ is an OU process:
+\begin{align*}
+d\varepsilon_{\tau} = -\phi \varepsilon_{\tau}d\tau + \sigma_{\varepsilon} dB_{\tau}
+\end{align*}
+where $B_\tau$ is standard Brownian motion. It\^{o}'s lemma gives the stochastic differential equation for $\gamma_{\tau} = e^{\varepsilon_{\tau} + \mu_{\tau}}$:
+\begin{align*}
+d\gamma_\tau = -\phi(\log\gamma_\tau - \mu_\tau)\gamma_\tau + \sigma_{\varepsilon}^2/2\gamma_\tau +d\mu_\tau \gamma_\tau + \sigma_{\varepsilon}\gamma_\tau dB_\tau.
+\end{align*}
+The solution to the OU process is
+\begin{align*}
+\varepsilon_{\tau} = e^{-\phi\tau}\left(\varepsilon_0 + \int_0^\tau \sigma_{\varepsilon}e^{\phi s}dB_s\right)
+\end{align*}
+so that 
+\begin{align*}
+\varepsilon_{\tau + \delta} | \varepsilon_{\tau} \sim N\left(e^{-\phi\delta}\varepsilon_{\tau}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right).
+\end{align*}
+and thus
+\begin{align*}
+\log \gamma_{\tau + \delta}|\log\gamma_\tau \sim N\left(e^{-\phi\delta}(\log\gamma_{\tau} - \mu_\tau) + \mu_{\tau + \delta}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right).
+\end{align*}
+In other words, $\gamma_{\tau + \delta}$ has a lognormal distribution.
+
+To stitch this piece onto the larger model, we exploit the inifnite divisibility property of the gamma distribution. Suppose that
+\begin{align*}
+x_{\tau + \delta} - x_\tau &\sim G(\beta \int_{\tau}^{\tau + \delta} d\gamma_s, \beta)
+\end{align*}
+with $d\gamma_\tau$ defined as above.
+Since both $x_{\tau}$ and $\gamma_{\tau}$ are continuous time processes, it should be possible to break the interval $[\tau, \tau + \delta]$ into a set of subintervals $[\tau + \delta_k, \tau + \delta_{k+1}]$, $0=\delta_0<\delta_1<\dots<\delta_K=\delta$ such that $x_{\tau + \delta} - x_{\tau} = \sum_{k=1}^K (x_{\tau + \delta_k} - x_{\tau + \delta_{k-1}})$.
+Since $\int_{\tau}^{\tau + \delta}d\gamma_s = \gamma_{\tau + \delta} - \gamma_{\tau}$ we have
+\begin{align*}
+x_{\tau + \delta_k} - x_{\tau + \delta_{k-1}} \stackrel{ind}{\sim} G(\beta(\gamma_{\tau + \delta_k} - \gamma_{\tau + \delta_{k-1}}), \beta)
+\end{align*}
+so that by the gamma distribution's infinite divisibility property
+\begin{align*}
+\sum_{k=1}^Kx_{\tau + \delta_k} - x_{\tau + \delta_{k-1}} \sim G\left(\beta\sum_{k=1}^K(\gamma_{\tau + \delta_k} - \gamma_{\tau + \delta_{k-1}}), \beta\right) = G\left(\beta(\gamma_{\tau + \delta} - \gamma_\tau), \beta\right).
+\end{align*}
+In other words, this is consistent with the assumed distribution on $x_{\tau + \delta} - x_{\tau}$.
+
+The full model can then be written as
+\begin{align*}
+t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
+x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G\left(\beta(\gamma_{\tau_k + \delta_k} - \gamma_{\tau_k}),\beta\right)\\
+\gamma_{\tau_k + \delta_k}|\gamma_{\tau_k} & \sim LN(\left(e^{-\phi\delta}(\log\gamma_{\tau_k} - \mu_{\tau_k}) + \mu_{\tau_k + \delta_k}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right)\\
+\delta_{k} = \tau_{k+1} - \tau_{k} &\stackrel{iid}{\sim} Exp(1/10)
+\end{align*}
+for blocks $k=1,2,\dots,N$ where $\mu_{\tau}$ is a parameterized deterministic function of $\tau$.
+For example, a trend with seasonal variability can be obtained via $\mu_{\tau} = \alpha_0 + \alpha_1\tau + \alpha_2\sin(\omega \tau)$ where $alpha_{0:2}$ is an uknown parameter and $\omega$ is the known frequency of the seasonal effect.
 
@@
 where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function.
-The key parameter we will control in these simulations is the mean transaction data rate, $\mu$, or the distribution of this rate.
+The key parameter we will control in these simulations is the mean transaction data rate, $\gamma$, or the distribution of this rate.
 Starting with the basic model fit to, say, the last 1000 blocks on the blockchain (approximately one week of elapsed time), we can control this parameter in several ways.
@@
 Then estimate the probability of a block filling up by simulating the model conditional on these parameter values.
-In order to take into account increased Bitcoin usage, we can manually increase the mean transaction data rate parameter, $\mu$, while holding other parameters constant.
+In order to take into account increased Bitcoin usage, we can manually increase the mean transaction data rate parameter, $\gamma$, while holding other parameters constant.
 
-This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\mu$ increases.
+This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\gamma$ increases.
 We can mitigate 1) while making 2) if not worse, different.
@@
 Then we can simulate from the posterior of the parameters and for each simulated parameter, simulate the model in order to estimate the probability of a block filling up.
-This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\mu$'s distribution were different, e.g.
-its mean $\bar{\mu}$.
-We estimate the probability of a block filling using both approaches, but because of the required assumptions the estimates become less credible for values of $\mu$ or $\bar{\mu}$ further away from the estimated values.
+This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different, e.g.
+its mean $\bar{\gamma}$.
+We estimate the probability of a block filling using both approaches, but because of the required assumptions the estimates become less credible for values of $\gamma$ or $\bar{\gamma}$ further away from the estimated values.
 
@@
   \hline
-$\mu - \hat{\mu}$ & $P_{\bar{\mu}}$ & $P_{\mu^{(0.5)}}$ & $P_{\mathrm{post}}$ \\ 
+$\gamma - \hat{\gamma}$ & $P_{\bar{\gamma}}$ & $P_{\gamma^{(0.5)}}$ & $P_{\mathrm{post}}$ \\ 
   \hline
@@
 \caption{Estimated probability of a filled block under a variety of assumptions.
-From the posterior distribution the mean is $\bar{\mu} = 0.0179$ and the median is $\mu^{(0.5)} = 0.0174$.
-The value of $\mu - \hat{\mu}$ is the amount added to the estimated value of $\mu$ for the purposes of the simulation, $P_{\bar{\mu}}$ is the probability of a filled block using posterior means for parameter estimates, $P_{\mu^{(0.5)}}$ is the probability of a filled block using posterior medians, and $P_{\mathrm{post}}$ is the probability of a filled block using the full posterior distribution.}
+From the posterior distribution the mean is $\bar{\gamma} = 0.0179$ and the median is $\gamma^{(0.5)} = 0.0174$.
+The value of $\gamma - \hat{\gamma}$ is the amount added to the estimated value of $\gamma$ for the purposes of the simulation, $P_{\bar{\gamma}}$ is the probability of a filled block using posterior means for parameter estimates, $P_{\gamma^{(0.5)}}$ is the probability of a filled block using posterior medians, and $P_{\mathrm{post}}$ is the probability of a filled block using the full posterior distribution.}
 \label{tab:probsims}

### 2015-10-08 outline/outline.tex
tweaked the seasonal effects model

@@
 for blocks $k=1,2,\dots,N$ where $\mu_{\tau}$ is a parameterized deterministic function of $\tau$.
-For example, a trend with seasonal variability can be obtained via $\mu_{\tau} = \alpha_0 + \alpha_1\tau + \alpha_2\sin(\omega \tau)$ where $alpha_{0:2}$ is an uknown parameter and $\omega$ is the known frequency of the seasonal effect.
+For example, a trend with seasonal variability can be obtained via $\mu_{\tau} = \alpha_0 + \alpha_1\tau + \alpha_2\sin(2\pi f \tau + \lambda)$ where $alpha_{0}$, $\alpha_1$, $\alpha_2$, and $\lambda$ are uknown parameters and $f$ is the known frequency of the seasonal effect.
 

### 2015-10-08 outline/outline.tex
Attempted to fit seasonal model: doesn't work because of a mathematical mistake

--- 2015-10-07 outline/outline.tex
@@
 x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G\left(\beta(\gamma_{\tau_k + \delta_k} - \gamma_{\tau_k}),\beta\right)\\
-\gamma_{\tau_k + \delta_k}|\gamma_{\tau_k} & \sim LN(\left(e^{-\phi\delta}(\log\gamma_{\tau_k} - \mu_{\tau_k}) + \mu_{\tau_k + \delta_k}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right)\\
+\gamma_{\tau_k + \delta_k}|\gamma_{\tau_k} & \sim LN\left(e^{-\phi\delta}(\log\gamma_{\tau_k} - \mu_{\tau_k}) + \mu_{\tau_k + \delta_k}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right)\\
 \delta_{k} = \tau_{k+1} - \tau_{k} &\stackrel{iid}{\sim} Exp(1/10)
@@
 for blocks $k=1,2,\dots,N$ where $\mu_{\tau}$ is a parameterized deterministic function of $\tau$.
-For example, a trend with seasonal variability can be obtained via $\mu_{\tau} = \alpha_0 + \alpha_1\tau + \alpha_2\sin(\omega \tau)$ where $alpha_{0:2}$ is an uknown parameter and $\omega$ is the known frequency of the seasonal effect.
+For example, a trend with seasonal variability can be obtained via $\mu_{\tau} = \alpha_0 + \alpha_1\tau + \alpha_2\sin(2\pi f \tau + \psi)$ where $alpha_{0}$, $\alpha_1$, $\alpha_2$, and $\psi$ are uknown parameters and $f$ is the known frequency of the seasonal effect.
 

### 2015-10-08 model/modelfits.R

@@ -25,7 +25,7 @@ timedata$min <- timedata$time/60 ## convert to minutes
 save(timedata, file = "timedata.RData")
 load("timedata.RData")
 
-nobs <- 1000
+nobs <- 100
 timedatashort <- timedata[(1000 - nobs + 1):1011,]
 lbs <- rep(0,nobs)
 for(i in 1:nobs){
@@ -104,3 +104,22 @@ rownames(out) <- mus
 library(xtable)
 
 xtable(out)
+
+
+
+
+
+
+omega <- 2*pi/(60*24)
+
+gamlist <- list(nobs = nobs, t = times, lb = lbs, lambda = 1/10, mb = mbs, sigpars = c(1, 1), sigepspars = c(1, 1), bpars=c(1,1), omega=omega, alphamns=rep(0,3), alphasds=rep(1,3), rhopars=c(1,1), phipar = 2*pi, lgam0pars = c(0, 1))
+## lambda = 1 / mean block discovery time = 1/10 minutes
+
+fit <- stan(file = 'seasonallognormalgamma2.stan', data = gamlist, iter=1, chains=1)
+
+system.time(fit2 <- stan(fit = fit, data = gamlist, iter=100, chains=1, control = list(adapt_delta = 0.98, max_treedepth = 12)))
+
+traceplot(fit2, pars=c('b', 'sigma', 'phi', 'sigeps', 'alpha0', 'alpha1', 'alpha2', 'rho'))
+
+##, control = list(adapt_delta = 0.98, max_treedepth = 14)))
+

### 2015-10-12 model/modelfits.R
Added code for fitting the gamma model. Note: wanted GIG, but Stan can't do it in our case

@@ -1,6 +1,7 @@
-## this is mostly a scratchpad right now
 library(DBI)
 library(rstan)
+library(MCMCpack)
+library(xtable)
 
 con <- dbConnect(RPostgres::Postgres(),dbname = 'toshi', 
                  host = 'toshi.cn6zzwcfsto5.us-east-1.rds.amazonaws.com',
@@ -10,42 +11,84 @@ con <- dbConnect(RPostgres::Postgres(),dbname = 'toshi',
 
 dbListTables(con)
 
-res <- dbSendQuery(con, "select height, time, size, transactions_count from blocks order by height desc limit 1011")
+res <- dbSendQuery(con, "select height, time, size, transactions_count from blocks order by height desc limit 5011")
 while(!dbHasCompleted(res)){
   chunk <- dbFetch(res, n = 5)
   print(nrow(chunk))
 }
 
-timedata <- dbFetch(res)
+blockdata <- dbFetch(res)
 dbClearResult(res)
-timedata <- timedata[order(timedata$height),]
-timedata$elapsed_time <- timedata$time - c(NA,timedata$time[-length(timedata$time)])
-timedata$min <- timedata$time/60 ## convert to minutes
+blockdata <- blockdata[order(blockdata$height),]
+blockdata$elapsed_time <- blockdata$time - c(NA,blockdata$time[-length(blockdata$time)])
+blockdata$min <- blockdata$time/60 ## convert to minutes
 
-save(timedata, file = "timedata.RData")
-load("timedata.RData")
+save(blockdata, file = "blockdata.RData")
 
-nobs <- 100
-timedatashort <- timedata[(1000 - nobs + 1):1011,]
+load("blockdata.RData")
+
+
+############# Gamma fits
+
+## fit on all 5000
+mbs <- blockdata$size[-c(1:11)]/1000000  ## convert to megabytes
+gamma5000list <- list(mb = mbs, logalphapars = c(0,10), betapars = c(1, 1), nobs = length(mbs))
+g5000fit <- stan(file = 'gamma.stan', data = gamma5000list, chains = 4, iter = 4000)
+g5000out <- extract(g5000fit)
[…]

### 2015-10-12 outline/outline.tex
Major update, still in progress. Flows much better and includes results from gamma fit showing a nonconstant rate is necessary

--- 2015-10-07 outline/outline.tex
@@
 
-\section{The Basic Model}
+\section{Motivation}
+As Bitcoin becomes more popular, the transaction rate increases.
+Since existing Bitcoin rules limit the amount of transactions per block to one MB at some point the blocks will begin filling up.
 Bitcoin blocks are by design distributed according a Poisson process with a rate of discovery of one block every 10 minutes.
@@
 Let $\tau_k$ denote the the time when block $k$ arrives.
-Then $\tau_{k+1} - \tau_{k} \stackrel{iid}{\sim} exp(1/10)$ for $k=1,2,\dots,K$.
-This part of the model is determined for us, but determining the how the number of transactions is distributed or, more precisely, how the amount of transaction {\it data} is distributed requires more work.
+Then $\delta_k \equiv \tau_{k+1} - \tau_{k} \stackrel{iid}{\sim} Exp(1/10)$ for $k=1,2,\dots,K$, where if $x\sim Exp(\lambda)$ it has the density $p(x) = \lambda e^{-x\lambda}$ for $x>0$ and $\lambda>0$ with mean $1/\lambda$.
 
+There is always some chance that a given block will fill up since arrival times are stochastic. Specifically, for the $k$'th block this probability is
+\[
+P\left(x_{\tau_k + \delta_k} - x_{\tau_k} > 1\right).
+\]
+This probability depends crucially on the process that $x_\tau$ follows. If we assume that $x_{\tau}$ is determinstic with a constant transaction rate of $\gamma$ MB/min, then 
+\[
+x_{\tau_k + \delta_k} - x_{\tau_k} = \gamma\delta_k \stackrel{ind}{\sim} Exp(1/10\gamma)
+\]
+and the probability of a full block is available in closed form as
+\[
+P\left(x_{\tau_k + \delta_k} - x_{\tau_k} > 1\right) = \int_1^\infty 10\gamma e^{-10\gamma x}dx = e^{-10\gamma}.
+\]
+
+Clearly a constant transaction rate does not capture Bitcoin's behavior since it has been growing in popularity.
+With a non-constant and potentially stochastic transaction rate, the probability of a block filling up is no longer necessarily available in closed form.
+Using a Monte Carlo simulation of $N$ draws of the block sizes from the process $x_{\tau}$, we can approximate the probability of block $k$ filling up with
+\[
+\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{x^{(i)}_{\tau_{k}^{(i)}+\delta_k^{(i)}} - x^{(i)}_{\tau_k^{(i)}} > 1\right\}
+\]
+where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function.
+If we let $\gamma$ denote the mean transaction rate, which is the same as the actual transaction rate in the constant, deterministic case, we can alter $\gamma$ to see how the probability of a full block changes under a variety of regimes.
+
+A key question is whether the assumption of a constant transaction rate is approximately correct on a short enough time-scale.
+In order to test this assumption we fit a the gamma distribution to the block sizes of various pieces of the block chain.
+We use the shape-rate parameterization of the gamma distribution so that if $x\sim G(\alpha,\beta)$, it has the density
+\[
+p(x) = \frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-x\beta}
+\]
+for $x>0$ with mean $\alpha/\beta$ and variance $\alpha/\beta^2$.
+Since the $G(1,\beta)$ and $Exp(\beta)$ distributions are identical, if we find that $\alpha\neq 1$ then transaction rates cannot be constant.
+We focus on the most recent 5000 blocks in our database, which is in the process of downloading the entire chain, ending with block 303151.
+For priors we will assume $\log \alpha \sim N(0, 10)$ and $\beta \sim G(1, 1)$.
+This prior on $\alpha$ implies that the prior median is $\alpha=1$ so that the null hypothesis is favored relatively strongly by the prior.
+We fit the model to all 5000 blocks, the first 1000 blocks, the second 1000 blocks, etc., and the first 500 blocks, the second 500 blocks, etc.
+The results are in Table \ref{tab:gammafit}.
+
+% latex table generated in R 3.2.2 by xtable 1.7-4 package
+% Mon Oct 12 10:44:18 2015
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrr}
+  \hline
+ & Mean & 2.5\% & 97.5\% \\ 
+  \hline
+       $n=5000$ & 1.34 & 1.29 & 1.39 \\ 
+$n=1000$, $g=1$ & 1.26 & 1.17 & 1.37 \\ 
+$n=1000$, $g=2$ & 1.36 & 1.26 & 1.47 \\ 
+$n=1000$, $g=3$ & 1.37 & 1.26 & 1.47 \\ 
+$n=1000$, $g=4$ & 1.27 & 1.17 & 1.37 \\ 
+$n=1000$, $g=5$ & 1.41 & 1.30 & 1.52 \\ 
+ $n=500$, $g=1$ & 1.29 & 1.15 & 1.45 \\ 
+ $n=500$, $g=2$ & 1.23 & 1.10 & 1.37 \\ 
+ $n=500$, $g=3$ & 1.46 & 1.31 & 1.63 \\ 
+ $n=500$, $g=4$ & 1.28 & 1.14 & 1.42 \\ 
+ $n=500$, $g=5$ & 1.55 & 1.38 & 1.74 \\ 
+ $n=500$, $g=6$ & 1.21 & 1.08 & 1.34 \\ 
+ $n=500$, $g=7$ & 1.24 & 1.11 & 1.38 \\ 
+ $n=500$, $g=8$ & 1.29 & 1.15 & 1.44 \\ 
+ $n=500$, $g=9$ & 1.36 & 1.21 & 1.52 \\ 
+ $n=500$, $g=10$ & 1.46 & 1.30 & 1.63 \\ 
+ \hline
+\end{tabular}
+\caption{Posterior mean and 95\% credible intervals for $\alpha$ from fitting a gamma distribution to block sizes of the last 5000 blocks in our database.
+For sample sizes less than 5000, blocks were broken into groups of equal size and the model was fit separately to each group.
+Sample size is indicated by $n$, group id by $g$.
+Each model was fit in RStan\protect\footnotemark using 4 chains, each with 4000 iterations, 2000 of which used for warmup/burn in.}
+\label{tab:gammafit}
+\end{table}
+
+For\footnotetext{Cite RStan} the full 5000 blocks, the posterior mean of $\alpha$ is 1.34 and the lower bound of the 95\% credible interval is 1.29, so we have strong evidence that $\alpha$ is nonzero and that a constant transaction rate is not appropriate.
+This is not surprising since Bitcoin usage has gone up over time, but on a short enough time scale usage may be approximately constant.
+A block is around 10 minutes on average, so 5000 blocks takes a little over a month on average.
+It is possible that on a short enough time-scale, constant transaction rates are more appropriate.
+So we fit the gamma distribution to 1000 blocks at a time --- representing about a week --- by breaking the original 5000 blocks into five groups.
+For each of these groups the 95\% credible interval does not contain one --- the lowest lower bound is 1.17.
+Similarly for groups of 500 blocks, each representing about 3.5 days, the lowest lower bound is about 1.08.
+So we can safely conclude that the transaction rate is not constant.
+A likely reason for this is seasonality in transaction rates --- Bitcoin trades are probably more likely to happen during certain times of day than others, based on when relevant markets are open and when the bulk of Bitcoin users are awake.
+The {\it mean} transaction rate averaging over e.g.
+seasonality is probably approximately constant for time-scales this small, however.
+In any case, transaction rates are nonconstant even on short time-scales, which motivates modeling them as such.
+
+\section{The Model}
+
+The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how how the amount of transaction data is distributed requires more work.
 Crucial constraints on the models we can fit come from the data obtained from the block chain.
@@
 Two constraints on the reported time that the block was mined are built into the Bitcoin protocol.
-First, any block submitted to the chain with a local time less than the median local times of the last 11 blocks is rejected.
-Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node, with a maximum offset of 70 minutes.
+First, any block submitted to the chain with a local time less than the median of local times of the last 11 blocks is rejected.
+Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node with a maximum offset of 70 minutes.
 Because of the nature of the blocks' timestamps, occasionally a block has a timestamp which is earlier than one or more previous blocks in the chain.
@@
 These two constraints motivate a class of models for modeling Bitcoin transaction data.
-Let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time it is the current block so that $\tau_{k+1}=\tau_k + \delta_k$, and $t_k$ the observed arrival time in the blockchain.
-Further, let $x_\tau$ denote the amount data in the chain at time $\tau$ so that $x_{\tau_{k+1}} - x_{\tau_k}$ denotes the amount of data in block $k$.
-Now suppose the amount of transaction data in MBs is distributed according to some stochastic process $dx_\tau$ so that $x_{\tau + \delta} - x_\tau = \int_\tau^{\tau + \delta}dx_\tau$, a stochastic integral.
-Then the amount of transaction data that arrives in $k$'th block is given by $\int_{\tau_k}^{\tau_{k}+\delta_k}dx_\tau$.
-
-Most of the interesting modeling choices come from modeling $dx_\tau$.
-Conventional Brownian motion is inappropriate since the increments should be nonnegative.
-On the other hand, geometric Brownian motion is a poor choice since it cannot start from zero.
-Instead, we consider the classes of non-negative infinite activity L\'{e}vy processes with closed form solutions for $x_{\tau}$.
+Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time it is the current block so that $\tau_{k+1}=\tau_k + \delta_k$, and now $t_k$ the observed arrival time in the blockchain.
+Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k} + \delta_k} - x_{\tau_k}$ denotes the amount of data in block $k$.
+Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing, infinitely divisible\footnote{Should we define?} stochastic process.
+Most of the interesting modeling choices come from modeling this stochastic process.
+Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero.
+Instead, we consider the classes of non-negative infinite activity\footnote{Should we define?} L\'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$.\footnote{Cite the Levy process chapter here}.
 The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$.
 Brownian motion is the most common example.
-A better example for our purposes is the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$.
-We use the shape-rate parameterization of the gamma distribution so that if $x\sim G(alpha,\beta)$, it has the density
-\[
-p(x) = \frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-x\beta}
-\]
-for $x>0$ with mean $\alpha/\beta$ and variance $\alpha/\beta^2$.
+A nondecreasing example is the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$.
 
-While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x\sim G(\alpha_x,\beta)$ independent of $y\sim G(\alpha_y,\beta)$, then $x+y \sim G(\alpha_x + \alpha_y, \beta)$.
+While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$.
 Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$.
@@
 It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error.
-This simplifies Markov chain Monte Carlo methods that may be used to fit the model by allowing us to exactly simulate $x_{\tau_k + \delta_k}$ without resorting to approximation.
+This simplifies Markov chain Monte Carlo methods that may be used to fit the model by allowing us to exactly simulate $x_{\tau_k + \delta_k}$ without resorting to approximation methods.
 The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g.
-seasonal variation or stochastic volatility, but many reasonable models on these parameters will not allow for closed form distributions of the parameter increments.
+seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the parameter increments.
+We will use the gamma process since it is widely available in statistical packages, unlike the main alternative, the inverse Gaussian.
 
@@
 A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative.
-However the gamma causes computational problems since it forces the latent measurement times, $\tau_k$ to enter the gamma function.
-A similar model that avoids this problem is the lognormal distribution.
+However the gamma causes numerical problems at times since it forces the latent measurement times, $\tau_k$ to enter the gamma function.
+A similar model that avoids this problem is the lognormal distribution, which we will use.
 
-To start, consider the following parameterized model using the structure discussed above. For block $k=1,2,\dots,K$
+Combining these pieces we obtain the following model. For block $k=1,2,\dots,K$
 \begin{align*}
@@
 x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\gamma\delta_k,\beta)\\
-\delta_{k} = \tau_{k} - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
+\delta_{k} = \tau_{k+1} - \tau_{k} &\stackrel{iid}{\sim} Exp(1/10)
 \end{align*}
 where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$.
-Note that $x_{\tau_{k+1}} - x_{\tau_k}$ depends on $\delta_k$ because the $k$'th block arrives at $\tau_k$ and departs at $\tau_{k+1}=\tau_k + \delta_k$ --- the block has $\delta_k$ minutes to accumulate transaction data.
+Note that $x_{\tau_{k+1}} - x_{\tau_k}$ depends on $\delta_k$ because the $k$'th block arrives at $\tau_k$ and departs at $\tau_{k+1}=\tau_k + \delta_k$ so that block $k$ has $\delta_k$ minutes to accumulate transaction data.
+To complete the model we assume $\sigma$, $\beta$, and $\gamma$ are independent in the prior with $\sigma \sim Cauchy(0, 2.5)$, $\beta \sim G(1, 1)$, and $\gamma \sim LN(0, 10)$.
+Table \ref{tab:modelfit} contains the results of fitting the model.
 
-\subsection{Stochastic transaction rates --- seasonality and autocorrelation}
-It is unlikely that the transaction rate is constant in time.
+% latex table generated in R 3.2.2 by xtable 1.7-4 package
+% Mon Oct 12 15:04:44 2015
+\begin{table}[ht]
+  \centering
+  \begin{tabular}{lrrrrrrrr}
+    \hline
+            & Mean    & SD     & SE(Mean) & 2.5\%   & 25\%    & 50\%    & 75\%    & 97.5\% \\ 
+    \hline
+   $\gamma$ & 0.0240  & 0.0004 & 0.0000   & 0.0232  & 0.0237  & 0.0240  & 0.0243  & 0.0248 \\ 
+   $\beta$  & 15.1590 & 0.9432 & 0.0105   & 13.4132 & 14.5042 & 15.1251 & 15.7791 & 17.0910 \\ 
+   $\sigma$ & 0.0024  & 0.0001 & 0.0000   & 0.0023  & 0.0024  & 0.0024  & 0.0024  & 0.0025 \\ 
+    \hline
+ \end{tabular}
+ \caption{Results of fitting the measurement error model with stochastic measurement times.
+We use RStan to fit the model, obtaining four chains each of 4000 iterations, 2000 of which used for warmup/burn in.\protect\footnotemark (Note: outdated numbers, updated ones coming) }
+ \label{tab:modelfit}
+ \end{table}
+
+%% \subsection{Stochastic transaction rates --- seasonality and autocorrelation}
+%% It is unlikely that the transaction rate is constant in time.
 The transaction rate is clearly growing as Bitcoin becomes more popular, and seasonal variation is expected throughout the data.
@@
 
-Let $\gamma_\tau$ denote the transaction rate at time $\tau$.
+%% Let $\gamma_\tau$ denote the transaction rate at time $\tau$.
 We model the evolution of this rate as a variant of the exponential Ornstein–Uhlenbeck (OU) process.
@@
 Then $\varepsilon_{\tau}$ is an OU process:
-\begin{align*}
-d\varepsilon_{\tau} = -\phi \varepsilon_{\tau}d\tau + \sigma_{\varepsilon} dB_{\tau}
-\end{align*}
-where $B_\tau$ is standard Brownian motion. It\^{o}'s lemma gives the stochastic differential equation for $\gamma_{\tau} = e^{\varepsilon_{\tau} + \mu_{\tau}}$:
-\begin{align*}
-d\gamma_\tau = -\phi(\log\gamma_\tau - \mu_\tau)\gamma_\tau + \sigma_{\varepsilon}^2/2\gamma_\tau +d\mu_\tau \gamma_\tau + \sigma_{\varepsilon}\gamma_\tau dB_\tau.
-\end{align*}
-The solution to the OU process is
-\begin{align*}
-\varepsilon_{\tau} = e^{-\phi\tau}\left(\varepsilon_0 + \int_0^\tau \sigma_{\varepsilon}e^{\phi s}dB_s\right)
-\end{align*}
-so that 
-\begin{align*}
-\varepsilon_{\tau + \delta} | \varepsilon_{\tau} \sim N\left(e^{-\phi\delta}\varepsilon_{\tau}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right).
-\end{align*}
-and thus
-\begin{align*}
-\log \gamma_{\tau + \delta}|\log\gamma_\tau \sim N\left(e^{-\phi\delta}(\log\gamma_{\tau} - \mu_\tau) + \mu_{\tau + \delta}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right).
-\end{align*}
-In other words, $\gamma_{\tau + \delta}$ has a lognormal distribution.
+%% \begin{align*}
+%% d\varepsilon_{\tau} = -\phi \varepsilon_{\tau}d\tau + \sigma_{\varepsilon} dB_{\tau}
+%% \end{align*}
+%% where $B_\tau$ is standard Brownian motion. It\^{o}'s lemma gives the stochastic differential equation for $\gamma_{\tau} = e^{\varepsilon_{\tau} + \mu_{\tau}}$:
+%% \begin{align*}
+%% d\gamma_\tau = -\phi(\log\gamma_\tau - \mu_\tau)\gamma_\tau + \sigma_{\varepsilon}^2/2\gamma_\tau +d\mu_\tau \gamma_\tau + \sigma_{\varepsilon}\gamma_\tau dB_\tau.
+%% \end{align*}
+%% The solution to the OU process is
+%% \begin{align*}
+%% \varepsilon_{\tau} = e^{-\phi\tau}\left(\varepsilon_0 + \int_0^\tau \sigma_{\varepsilon}e^{\phi s}dB_s\right)
+%% \end{align*}
+%% so that 
+%% \begin{align*}
+%% \varepsilon_{\tau + \delta} | \varepsilon_{\tau} \sim N\left(e^{-\phi\delta}\varepsilon_{\tau}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right).
+%% \end{align*}
+%% and thus
+%% \begin{align*}
+%% \log \gamma_{\tau + \delta}|\log\gamma_\tau \sim N\left(e^{-\phi\delta}(\log\gamma_{\tau} - \mu_\tau) + \mu_{\tau + \delta}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right).
+%% \end{align*}
+%% In other words, $\gamma_{\tau + \delta}$ has a lognormal distribution.
 
-To stitch this piece onto the larger model, we exploit the inifnite divisibility property of the gamma distribution. Suppose that
-\begin{align*}
-x_{\tau + \delta} - x_\tau &\sim G(\beta \int_{\tau}^{\tau + \delta} d\gamma_s, \beta)
-\end{align*}
-with $d\gamma_\tau$ defined as above.
+%% To stitch this piece onto the larger model, we exploit the inifnite divisibility property of the gamma distribution. Suppose that
+%% \begin{align*}
+%% x_{\tau + \delta} - x_\tau &\sim G(\beta \int_{\tau}^{\tau + \delta} d\gamma_s, \beta)
+%% \end{align*}
+%% with $d\gamma_\tau$ defined as above.
 Since both $x_{\tau}$ and $\gamma_{\tau}$ are continuous time processes, it should be possible to break the interval $[\tau, \tau + \delta]$ into a set of subintervals $[\tau + \delta_k, \tau + \delta_{k+1}]$, $0=\delta_0<\delta_1<\dots<\delta_K=\delta$ such that $x_{\tau + \delta} - x_{\tau} = \sum_{k=1}^K (x_{\tau + \delta_k} - x_{\tau + \delta_{k-1}})$.
 Since $\int_{\tau}^{\tau + \delta}d\gamma_s = \gamma_{\tau + \delta} - \gamma_{\tau}$ we have
-\begin{align*}
-x_{\tau + \delta_k} - x_{\tau + \delta_{k-1}} \stackrel{ind}{\sim} G(\beta(\gamma_{\tau + \delta_k} - \gamma_{\tau + \delta_{k-1}}), \beta)
-\end{align*}
-so that by the gamma distribution's infinite divisibility property
-\begin{align*}
-\sum_{k=1}^Kx_{\tau + \delta_k} - x_{\tau + \delta_{k-1}} \sim G\left(\beta\sum_{k=1}^K(\gamma_{\tau + \delta_k} - \gamma_{\tau + \delta_{k-1}}), \beta\right) = G\left(\beta(\gamma_{\tau + \delta} - \gamma_\tau), \beta\right).
-\end{align*}
-In other words, this is consistent with the assumed distribution on $x_{\tau + \delta} - x_{\tau}$.
+%% \begin{align*}
+%% x_{\tau + \delta_k} - x_{\tau + \delta_{k-1}} \stackrel{ind}{\sim} G(\beta(\gamma_{\tau + \delta_k} - \gamma_{\tau + \delta_{k-1}}), \beta)
+%% \end{align*}
+%% so that by the gamma distribution's infinite divisibility property
+%% \begin{align*}
+%% \sum_{k=1}^Kx_{\tau + \delta_k} - x_{\tau + \delta_{k-1}} \sim G\left(\beta\sum_{k=1}^K(\gamma_{\tau + \delta_k} - \gamma_{\tau + \delta_{k-1}}), \beta\right) = G\left(\beta(\gamma_{\tau + \delta} - \gamma_\tau), \beta\right).
+%% \end{align*}
+%% In other words, this is consistent with the assumed distribution on $x_{\tau + \delta} - x_{\tau}$.
 
-The full model can then be written as
-\begin{align*}
-t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
-x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G\left(\beta(\gamma_{\tau_k + \delta_k} - \gamma_{\tau_k}),\beta\right)\\
-\gamma_{\tau_k + \delta_k}|\gamma_{\tau_k} & \sim LN(\left(e^{-\phi\delta}(\log\gamma_{\tau_k} - \mu_{\tau_k}) + \mu_{\tau_k + \delta_k}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right)\\
-\delta_{k} = \tau_{k+1} - \tau_{k} &\stackrel{iid}{\sim} Exp(1/10)
-\end{align*}
-for blocks $k=1,2,\dots,N$ where $\mu_{\tau}$ is a parameterized deterministic function of $\tau$.
-For example, a trend with seasonal variability can be obtained via $\mu_{\tau} = \alpha_0 + \alpha_1\tau + \alpha_2\sin(\omega \tau)$ where $alpha_{0:2}$ is an uknown parameter and $\omega$ is the known frequency of the seasonal effect.
+%% The full model can then be written as
+%% \begin{align*}
+%% t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
+%% x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G\left(\beta(\gamma_{\tau_k + \delta_k} - \gamma_{\tau_k}),\beta\right)\\
+%% \gamma_{\tau_k + \delta_k}|\gamma_{\tau_k} & \sim LN\left(e^{-\phi\delta}(\log\gamma_{\tau_k} - \mu_{\tau_k}) + \mu_{\tau_k + \delta_k}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right)\\
+%% \delta_{k} = \tau_{k+1} - \tau_{k} &\stackrel{iid}{\sim} Exp(1/10)
+%% \end{align*}
+%% for blocks $k=1,2,\dots,N$ where $\mu_{\tau}$ is a parameterized deterministic function of $\tau$.
+For example, a trend with seasonal variability can be obtained via $\mu_{\tau} = \alpha_0 + \alpha_1\tau + \alpha_2\sin(2\pi f \tau + \psi)$ where $alpha_{0}$, $\alpha_1$, $\alpha_2$, and $\psi$ are uknown parameters and $f$ is the known frequency of the seasonal effect.
 
-\section{The key question}
-As Bitcoin becomes more popular, the transaction data rate increases.
-Since existing Bitcoin rules limit the number of transactions per block to one MB, at some point the blocks will begin filling up.
-The purpose of modeling Bitcoin's transaction data rate is to estimate the probability of seeing a full block under a variety of assumptions.
-Starting from the base model above, the quantity we are interested in is thus
-\[
-P\left(\int_{\tau_k}^{\tau_{k}+\delta_k}dx_\tau > 1\right)
-\]
-for some choice of $k$.
-
-For some distributional forms for the process $dx_\tau$ it may be possible to compute these quantities analytically, but in general it may be difficult.
-Using a Monte Carlo simulation of $N$ draws, we can approximate these probabilities by
-\[
-\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{\int_{\tau_k^{(i)}}^{\tau_{k}^{(i)}+\delta_k^{(i)}}dx_\tau^{(i)} > 1\right\}
-\]
-where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function.
-The key parameter we will control in these simulations is the mean transaction data rate, $\gamma$, or the distribution of this rate.
-Starting with the basic model fit to, say, the last 1000 blocks on the blockchain (approximately one week of elapsed time), we can control this parameter in several ways.
-
-The simplest way to compute parameter estimates based on fitting model, e.g.
-maximum likelihood estimates, posterior means, or posterior medians.
-Then estimate the probability of a block filling up by simulating the model conditional on these parameter values.
+Given\footnotetext{Cite RStan} an approximate simulation of the posterior distribution of the model parameters, we can use Monte Carlo simulation to estimate the probability of a block filling up in a number of ways.
+The simplest way is to compute parameter estimates based on the posterior, e.g.
+posterior means or posterior medians, then estimate the probability of a block filling up by simulating the model conditional on these parameter values.
 In order to take into account increased Bitcoin usage, we can manually increase the mean transaction data rate parameter, $\gamma$, while holding other parameters constant.
+In order to take into account seasonality, we can make $\gamma$ change in a predictable, seasonal pattern.
 
@@
 We can mitigate 1) while making 2) if not worse, different.
-Specifically, suppose we fit the model and have access to the full (approximate) posterior distribution of the model parameters.
-Then we can simulate from the posterior of the parameters and for each simulated parameter, simulate the model in order to estimate the probability of a block filling up.
-This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different, e.g.
-its mean $\bar{\gamma}$.
-We estimate the probability of a block filling using both approaches, but because of the required assumptions the estimates become less credible for values of $\gamma$ or $\bar{\gamma}$ further away from the estimated values.
+Specifically, given access to an approximate simulation of the posterior of the model parameters, we can simulate the model for each draw from the posterior in order to estimate the probability of a block filling up --- this is the posterior predictive distribution.\footnote{Cite Gelman et al.
+text here} In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ by some amount $\psi$ so that the posterior mean of the shifted distribution of $\gamma$ is the desired amount.
+This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different.
+We estimate the probability of a block filling using all three approaches, but because of the required assumptions the estimates become less credible for values of $\gamma$ or $\bar{\gamma}$ further away from the estimated values.
 
-Table \ref{tab:probsims} contains estimates of the probability of a filled block under a variety of assumptions, using parameter estimates of the posterior distribution of the model using 1000 contiguous blocks ending with block $253,221$ from August 20, 2013 arriving at about 11:39 a.m.
+Let $P_{\gamma,\beta}^{(k)}$ denote the probability of the $k$'th block filling up given the specified values for $\gamma$ and $\beta$, where the superscript $(k)$ is left off when this probability is independent of the block.
+Further let $P_{\gamma, \infty}$ denote case of a constant transaction $\gamma$ since for $\beta=\infty$, the $G(\beta \gamma,\beta)$ distribution can be thought of as a point mass at $\gamma$.
+Further, in this notation let $\tilde{\gamma}$ denote the posterior distribution of $\gamma$, $\bar{\gamma}$ the posterior mean, and $\gamma^{(0.5)}$ denote the posterior median, and similarly for $\beta$, so that e.g.
+$P_{\tilde{\gamma},\tilde{\beta}}$ is the probability of a block filling up under the posterior distribution of $(\gamma,\beta)$, and $P_{\tilde{\gamma} + \psi,\tilde{\beta}}$ is the probability of a block filling up under the posterior distribution of $(\gamma,\beta)$ except $\gamma$'s distribution has been shifted by $\psi$ so that it has mean $\bar{\gamma} + \psi$.
+Table \ref{tab:probsims} contains estimates of the probability of a filled block under a variety of assumptions, using parameter estimates of the posterior distribution of the model using 1000 contiguous blocks ending with block 303151.
 % latex table generated in R 3.2.2 by xtable 1.7-4 package
@@
 \end{tabular}
-\caption{Estimated probability of a filled block under a variety of assumptions.
-From the posterior distribution the mean is $\bar{\gamma} = 0.0179$ and the median is $\gamma^{(0.5)} = 0.0174$.
-The value of $\gamma - \hat{\gamma}$ is the amount added to the estimated value of $\gamma$ for the purposes of the simulation, $P_{\bar{\gamma}}$ is the probability of a filled block using posterior means for parameter estimates, $P_{\gamma^{(0.5)}}$ is the probability of a filled block using posterior medians, and $P_{\mathrm{post}}$ is the probability of a filled block using the full posterior distribution.}
+\caption{This table is a placeholder. Important question: what values for $\gamma$ are interesting?}
 \label{tab:probsims}
@@
 
-\section{Simulation Exercises}
-
-Each simulation exercise starts with specifying a distribution for $dx_t$.
-There are three basic categories here: deterministic, i.e.
-fixed at the mean rate, independent, and temporally dependent.
-There is actually some interesting room in the deterministic class --- for example, $dx_t$ could be increasing in $t$, or at least not constant.
-This behavior can be captured in a more realistic stochastic environment, so we will focus on that.
-
-There are several possibilities in the independent class.
-Transaction data amounts are nonnegative, so it is natural to assume that $dx_t \stackrel{iid}{\sim}\mathcal{G}(\alpha,\beta)$ where $\mathcal{G}(\alpha,\beta)$ denotes the gamma distribution with shape and rate parameters $\alpha$ and $\beta$ respectively.
-Similarly a log-normal distribution is natural, e.g.
-$\log(dx_t) \stackrel{iid}{\sim}\mathcal{N}(\mu,\sigma^2)$.
-The log-$T$ distribution is a useful extension of the log-normal.
-However, all of these distributions encode some degree of right-skewness which may not match the BTC transaction data.
-A useful generalization to account for all possible sorts of skew is the skew-normal or more generally skew-$T$ distribution.
-The skew-normal distribution has the pdf
-\begin{align*}
-p(x) = \frac{2}{\sigma}\phi\left(\frac{x-\mu}{\sigma}\right)\Phi\left(\frac{\alpha(x-\mu)}{\sigma}\right)
-\end{align*}
-where $\phi(.)$ is the standard normal pdf and $\Phi(.)$ is the standard normal cdf.
-The parameter $\alpha$ controls the degree and direction of skewness, while $\mu$ and $\sigma$ are respectively location and scale parameters.
-The definition of the skew-$T$ distribution is analogous.
-This is only one class of options that could be used to account for possible skewness in the distribution of BTC transaction data.
-
-With temporally dependent data, things start to get more interesting.
-I think a crucial feature here is that the mean structure be easily separable from the structure of the temporal dependence, that way we can easily ask questions like ``if the mean transaction rate increases to X given this dependence structure, what happens to the probability that a given block fills up?'' So the classic decomposition of a time series into trend, seasonal, and error components is exactly what we are looking for.
-To wit:
-\begin{align*}
-log(dx_t) = T_t + S_t + E_t
-\end{align*}
-where $T_t$ is a deterministic trend component, $S_t$ is a seasonal component which may be stochastic, and $E_t$ is a temporally dependent error component.
-In a particularly simple case $T_t = \mu$ and $E_t = \rho E_{t-1} + \varepsilon_t$ where $\varepsilon_t \stackrel{iid}{\sim} \mathcal{N}(0,\sigma^2)$, i.e.
-an AR(1) model.
-Any mean zero ARMA model can be used for $E_t$, or a plethora of other things, and the error term can also be changed to something skewed skewed if necessary.
-I suspect the most interesting part for us is the seasonal component --- seasonal variation over the time of day, week, month, or even year may cause certain time periods to have a very high probability of filling up the block.
-I have a less firm grasp on what I want to try here mostly because I have less experience with modeling seasonal effects, but this does seem like payoff portion of modeling the temporal dependence.
-
-Another payoff of seasonal effects term is, combined with the trend term, we can predict when in the future BTC will start to have trouble with blocks filling up under a given regime.
-Also note that we can make these predictions with {\it any} block size substituted in for the actual block size.
-This is probably not a fruitful exercise unless we build a model specifically with forecasting in mind though.
-
-\section{Fitting Models}
-Instead of choosing parameter values and doing the simulation exercise using the models listed in the previous section, we can estimate the models and use simulations form the posterior predictive distribution in order to do the same exercises while not just using more plausible parameter values, but also taking into account parameter uncertainty.
-A little more formally, suppose our parameter vector is $\bm{\theta}$ and $\bm{y}$ denotes our data.
-Let $\bm{y}^{pred}$ denote vector predicted observations.
-The posterior preidctive density for $\bm{y}^{pred}$ is given by
-\begin{align*}
-p(\bm{y}^{pred}|\bm{y}) = \int p(\bm{y}^{pred}|\bm{\theta})p(\bm{\theta}|\bm{y}) d\bm{\theta}
-\end{align*}
-where $p(y|x)$ denotes the conditional density of the random variable $y$ given the random variable $x$.
-Given a simulation from $p(\bm{\theta}|\bm{y})$, the posterior distribution of the model parameters, using e.g.
-MCMC, this density is easy to approximate simply by simulating draws of the predicted $\bm{y}^{pred}$ for each parameter value simulated.
-More generally, we can fix some of the parameter values and let others vary to, e.g., isolate the effect of increasing the mean of the time series.
-
-Fitting the models should be straightforward, at least in theory.
-STAN is probably great for what we are trying to do, which means way less work on our end.
-The only wrinkles here are 1) now we have to pick prior distributions, which may be controversial, but we can easily do some robustness checks with different priors and 2) potential big data concerns.
-
-\section{A Simple Example}
-
-In this example simulation we assume that the transaction data process $x_t$ follows a continuous time random walk with drift with parameters $\mu$ and $\sigma$, that is
-\[
-x_t - x_s \sim N((t-s)\mu, (t-s)\sigma^2)
-\]
-for $t > s$. So $dx_t = \mu dt + \sigma d B_t$ where $B_t$ is Brownian motion. Using a simulation, Table %\ref{tab:bmprob}
- contains Monte Carlo estimates of the probability that a block fills up under a variety of choices of $\mu$ and $\sigma$.
-The current average transaction data rate per minute  is $\mu=0.04$ MBs.
-As $\sigma$ increases the probability of a block filling increases, though when $\sigma$ is close to zero an increase in $\sigma$ may cause the probability to decrease slightly.
-For most of the range, however, increasing $\sigma$ increases the probability that a block fills up.
-
-% latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Tue Sep 22 16:26:38 2015
-%\begin{table}[ht]
-%\centering
-%\begin{tabular}{lrrrrrrrr}
-%  \hline
-% & $\sigma = 0$ & $\sigma = 0.001$ & $\sigma = 0.01$ & $\sigma = 0.1$ & $\sigma = 1$ & $\sigma = 10$ & $\sigma = 100$ & $\sigma = 1000$ \\ 
-%  \hline
-%$\mu = 0.035$ & 0.0594 & 0.0594 & 0.0597 & 0.0927 & 0.3574 & 0.4822 & 0.4979 & 0.4994 \\ 
-%  $\mu = 0.04$ & 0.0836 & 0.0835 & 0.0839 & 0.1149 & 0.3624 & 0.4824 & 0.4978 & 0.5005 \\ 
-%  $\mu = 0.045$ & 0.1098 & 0.1098 & 0.1103 & 0.1367 & 0.3680 & 0.4829 & 0.4977 & 0.4995 \\ 
-%  $\mu = 0.05$ & 0.1366 & 0.1366 & 0.1370 & 0.1596 & 0.3739 & 0.4837 & 0.4983 & 0.5003 \\ 
-%   \hline
-%\end{tabular}
-%\caption{Approximate probability of filling of a block under various choices of $\mu$ and $\sigma$%.}
-%\label{tab:bmprob}
-%\end{table}
-
 \end{document}

### 2015-10-13 outline/outline.tex
Added probability estimates

@@
 \[
-x_{\tau_k + \delta_k} - x_{\tau_k} = \gamma\delta_k \stackrel{ind}{\sim} Exp(1/10\gamma)
+x_{\tau_k + \delta_k} - x_{\tau_k} = \gamma\delta_k \stackrel{ind}{\sim} Exp(1/\gamma 10)
 \]
@@
 \[
-P\left(x_{\tau_k + \delta_k} - x_{\tau_k} > 1\right) = \int_1^\infty 10\gamma e^{-10\gamma x}dx = e^{-10\gamma}.
+P\left(x_{\tau_k + \delta_k} - x_{\tau_k} > 1\right) = \int_1^\infty 1/10\gamma e^{-1/10\gamma x}dx = e^{-1/10\gamma}.
 \]
@@
 % latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Mon Oct 12 15:04:44 2015
+% Mon Oct 12 22:27:11 2015
 \begin{table}[ht]
 \centering
-  \begin{tabular}{lrrrrrrrr}
+\begin{tabular}{rrrrrrrrr}
   \hline
@@
 $\gamma$ & 0.0240 & 0.0004 & 0.0000 & 0.0232 & 0.0237 & 0.0240 & 0.0243 & 0.0248 \\ 
-   $\beta$  & 15.1590 & 0.9432 & 0.0105   & 13.4132 & 14.5042 & 15.1251 & 15.7791 & 17.0910 \\ 
-   $\sigma$ & 0.0024  & 0.0001 & 0.0000   & 0.0023  & 0.0024  & 0.0024  & 0.0024  & 0.0025 \\ 
+  $\beta$ & 15.0951 & 0.9228 & 0.0103 & 13.3725 & 14.4461 & 15.0760 & 15.6962 & 16.9591 \\ 
+  $\sigma$ & 0.0024 & 0.0001 & 0.0000 & 0.0023 & 0.0023 & 0.0024 & 0.0024 & 0.0025 \\ 
    \hline
@@
 
-Let $P_{\gamma,\beta}^{(k)}$ denote the probability of the $k$'th block filling up given the specified values for $\gamma$ and $\beta$, where the superscript $(k)$ is left off when this probability is independent of the block.
-Further let $P_{\gamma, \infty}$ denote case of a constant transaction $\gamma$ since for $\beta=\infty$, the $G(\beta \gamma,\beta)$ distribution can be thought of as a point mass at $\gamma$.
-Further, in this notation let $\tilde{\gamma}$ denote the posterior distribution of $\gamma$, $\bar{\gamma}$ the posterior mean, and $\gamma^{(0.5)}$ denote the posterior median, and similarly for $\beta$, so that e.g.
-$P_{\tilde{\gamma},\tilde{\beta}}$ is the probability of a block filling up under the posterior distribution of $(\gamma,\beta)$, and $P_{\tilde{\gamma} + \psi,\tilde{\beta}}$ is the probability of a block filling up under the posterior distribution of $(\gamma,\beta)$ except $\gamma$'s distribution has been shifted by $\psi$ so that it has mean $\bar{\gamma} + \psi$.
  Table \ref{tab:probsims} contains estimates of the probability of a filled block under a variety of assumptions, using parameter estimates of the posterior distribution of the model using 1000 contiguous blocks ending with block 303151.
+Probabilities in the first row were computed exactly using the cdf of the exponential distribution.
+Probabilities in the second and third rows were computed by simulating 800,000 blocks from the model at the specified values for $\gamma$ and $\beta$, where $\beta^{(0.5)}$ denotes the posterior median for $\beta$.
+Probabilities in the fourth row were computed by simulating 100 blocks from the model for each of the 8000 draws of from the posterior of $\beta$ and with $\gamma$ fixed at the specified values.
+Probabilities in the fifth and final row were computed by simulating 100 blocks for each of the 8000 draws from the posterior of $\beta$ and the shifted posterior of $\gamma$, where a draw from the shifted posterior of $\gamma$ is a draw from the posterior plus the difference between the specified $\mathrm{E}[\gamma]$ and $\gamma$'s posterior mean.
 % latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Mon Oct  5 14:56:54 2015
+% Tue Oct 13 00:53:54 2015
 \begin{table}[ht]
 \centering
-\begin{tabular}{llll}
+\begin{tabular}{lrrrrrrrr}
   \hline
-$\gamma - \hat{\gamma}$ & $P_{\bar{\gamma}}$ & $P_{\gamma^{(0.5)}}$ & $P_{\mathrm{post}}$ \\ 
+$E[\gamma] =$ & 0.02 & 0.03 & 0.04 & 0.05 & 0.06 & 0.07 & 0.08 & 0.09 \\ 
   \hline
-0 & 0.01 & 0.01 & 0.00 \\ 
-  0.01 & 0.03 & 0.03 & 0.03 \\ 
-  0.05 & 0.23 & 0.23 & 0.22 \\ 
-  0.1 & 0.42 & 0.42 & 0.42 \\ 
-  0.5 & 0.82 & 0.82 & 0.83 \\ 
+$\beta = \infty$ & 0.0067 & 0.0357 & 0.0821 & 0.1353 & 0.1889 & 0.2397 & 0.2865 & 0.3292 \\ 
+  $\beta = 3\times\beta^{(0.5)}$ & 0.0082 & 0.0387 & 0.0853 & 0.1383 & 0.1915 & 0.2420 & 0.2874 & 0.3293 \\ 
+  $\beta = \beta^{(0.5)}$ & 0.0117 & 0.0446 & 0.0917 & 0.1435 & 0.1958 & 0.2446 & 0.2881 & 0.3300 \\ 
+  $\beta \sim p(\beta|x,t)$ & 0.0119 & 0.0452 & 0.0919 & 0.1437 & 0.1954 & 0.2438 & 0.2892 & 0.3304 \\ 
+  $(\beta,\gamma) \sim p(\beta,\gamma|x,t)$ & 0.0118 & 0.0448 & 0.0916 & 0.1433 & 0.1955 & 0.2442 & 0.2896 & 0.3301 \\ 
    \hline
 \end{tabular}
-\caption{This table is a placeholder. Important question: what values for $\gamma$ are interesting?}
+\caption{Probability of a full block under a variety of parameter values.
+In the first row the transaction rate is constant, in the second through fourth rows the transaction rate is stochastic with fixed mean $\gamma$, and in the final row the transaction rate is stochastic and unknown with mean $\mathrm{E}[\gamma]$.
+Under all five regimes the expected mean transaction rate is the same --- it is the spread of the distribution of transaction rates and mean transaction rates that changes across regimes.}
 \label{tab:probsims}
@@
 
+Under a fixed mean transaction rate, the probability of a full block increases as as $\beta$ decreases, especially for smaller mean transaction rates.
+For larger mean transaction rates there is little effect on the probability the next block is full from decreasing $\beta$.
+This is expected since increasing the spread of transaction rates while holding the mean constant increases the variance of block sizes.
+When block sizes are on average much smaller than 1 MB, increasing the variance of block sizes puts significantly more probability mass above 1 MB, but when the mean is near but still below 1 MB it has a relatively small effect.
+What is unexpected is that taking into account the uncertainty in $\beta$ and then in both $\gamma$ and $\beta$ seems to have essentially no effect on the probability that the next block fills up no matter the expected value of $\gamma$.
+
+These estimates seem rather high considering no block have filled up yet.
+Transaction rates have been fluctuating around 0.05 for several months and even under constant transaction rates, the probability of a full block over 0.1.
+This may be explained by several things.
+First, possibly transaction rates are usually low and only occasionally spike so that only during a spike is there a real chance of a block filling up.
+Another possibility is that Bitcoin traders are cognizant of the size limit and when a block is near full they wait to make their trades until a new block arrives.
+Since they would only have to wait 10 minutes to complete their transaction, on average, waiting it not much of a nuisance so the costs are small.
+
 \end{document}

### 2015-10-13 model/modelfits.R

@@ -94,77 +94,52 @@ save(fit2, file="loggammafit.RData")
 
 load("loggammafit.RData")
 
-rstan::traceplot(fit2, pars=c('a', 'b', 'mu', 'sigma', 'lp__'), ncol=1)
+rstan::traceplot(fit2, pars=c('gamma', 'beta', 'sigma', 'lp__'), ncol=1)
 
-print(fit2, pars=c('a', 'b', 'mu', 'sigma', 'lp__'), digits = 4)
+print(fit2, pars=c('gamma', 'beta', 'sigma', 'lp__'), digits = 4)
 
 fitex <- extract(fit2)
 
 niter <- 8000
-parout <- cbind(mu=fitex$mu, b=fitex$b, sigma=fitex$sigma)
+parout <- cbind(gamma=fitex$gamma, beta=fitex$beta, sigma=fitex$sigma)
 parsum <- summary(mcmc(parout))
 parout <- cbind(parsum[[1]][,c(1,2,4)], parsum[[2]])
 rownames(parout) <- c("$\\gamma$", "$\\beta$", "$\\sigma$")
-colnames(parout)[3] <- "SE of Mean"
+colnames(parout)[3] <- "SE(Mean)"
 
-xtable(parout, digits = 4)
+print(xtable(parout, digits = 4), sanitize.rownames.function = function(x)gsub("\\\\", "\\", x, fixed=TRUE))
 
-
-mus <- c(0, .01, .05, .1, .5)
-K <- length(mus)
-out <- matrix(0, ncol=3, nrow=K)
-
+gams <- seq(0.02, 0.09, 0.01)
+betamed <- parout[2,6]
+betamn <- parout[2,1]
+gammn <- parout[1,1]
+K <- length(gams)
+out <- matrix(0, ncol=K, nrow=5)
+nrep <- 100
 
 for(i in 1:K){
-  ## use posterior means
-  pars <- parsum[[1]][,1]
-  muhat <- pars[1]
-  bhat <- pars[2]
-  sigmahat <- pars[3]
-  xs <- rexp(niter, 1/10)
-  taus <- cumsum(xs)
-  mbs <- rgamma(niter, (muhat + mus[i])*bhat*xs, bhat)
-  out[i,1] <- mean(mbs>1)
-
-  ## use posterior medians
-  pars <- parsum[[2]][,3]
-  muhat <- pars[1]
-  bhat <- pars[2]
-  sigmahat <- pars[3]
[…]

### 2015-10-13 outline/outline.tex
Better explanation of why I think strategic behavior is happening

@@
 These estimates seem rather high considering no block have filled up yet.
-Transaction rates have been fluctuating around 0.05 for several months and even under constant transaction rates, the probability of a full block over 0.1.
-This may be explained by several things.
-First, possibly transaction rates are usually low and only occasionally spike so that only during a spike is there a real chance of a block filling up.
-Another possibility is that Bitcoin traders are cognizant of the size limit and when a block is near full they wait to make their trades until a new block arrives.
-Since they would only have to wait 10 minutes to complete their transaction, on average, waiting it not much of a nuisance so the costs are small.
+Transaction rates have been fluctuating around 0.05 for several months and even under constant transaction rates, the probability of a full block is over 0.1.
+Even at a constant transaction rate of 0.02, the probability of a full block is 0.067.
+In 100 blocks, which is approximately 17 hours, the probability of at least one block filling up is $1-(1-0.067)^{100}\approx 1$.
+Since no block has filled up yet, clearly something is happening that prevents blocks from filling up.
+One possibility is that transaction rates are usually low and in no danger of causing a full block, but occasionally there are large spikes in the rate.
+Another possibility is strategic action by Bitcoin traders and miners to prevent blocks from filling up.
+Traders may be waiting to trade until a new block replaces a nearly full block, which is not much of a cost since the wait is only about 10 minutes on average.
+Similarly miners have an incentive to keep blocks small since that reduces the change of an orphan block --- a mined block which is not accepted by the blockchain --- so the miner of the current block may be rejecting some transactions in order to prevent the limit from being reached.
 

### 2015-10-13 sims/sims.R
Outdated

@@ -1,29 +1,0 @@
-library(xtable)
-
-## sample size
-n <- 1000000
-
-## rate of block discovery
-lambda <- 1/10
-
-## means and variances of lognormal distribution
-mus <- c(0.035, 0.04, 0.045, 0.05)
-sigs <- c(0,10^(1:7-4))
-
-out <- matrix(0,nrow=length(mus),ncol=length(sigs))
-colnames(out) <- paste("$\\sigma = ", sigs, "$", sep="")
-rownames(out) <- paste("$\\mu = ", mus, "$", sep="")
-
-arrive <- rexp(n, lambda)
-for(i in 1:length(mus)){
-  for(j in 1:length(sigs)){
-    mu <- mus[i]
-    sig <- sigs[j]
-    mbs <- mu*arrive + sig*sqrt(arrive)*rnorm(n)
-    out[i,j] <- mean(mbs > 1)
-  }
-}
-
-
-xout <- xtable(out, digits=4)
-print.xtable(xout, digits=4, sanitize.text.function=function(x){x})

### 2015-10-13 outline/outline.tex
Fixed some of the language and notation describing how a block arrives and what happens when it arrives

@@
 Let $\tau_k$ denote the the time when block $k$ arrives.
-Then $\delta_k \equiv \tau_{k+1} - \tau_{k} \stackrel{iid}{\sim} Exp(1/10)$ for $k=1,2,\dots,K$, where if $x\sim Exp(\lambda)$ it has the density $p(x) = \lambda e^{-x\lambda}$ for $x>0$ and $\lambda>0$ with mean $1/\lambda$.
+Then $\delta_k \equiv \tau_{k} - \tau_{k-1} \stackrel{iid}{\sim} Exp(1/10)$ for $k=1,2,\dots,K$, where if $x\sim Exp(\lambda)$ it has the density $p(x) = \lambda e^{-x\lambda}$ for $x>0$ and $\lambda>0$ with mean $1/\lambda$.
+The longer $\delta_k$ the more time for transactions to add up before being added to the block chain as part of block $k$.
 
@@
 \[
-P\left(x_{\tau_k + \delta_k} - x_{\tau_k} > 1\right).
+P\left(x_{\tau_k} - x_{\tau_{k-1}} > 1\right).
 \]
@@
 \[
-x_{\tau_k + \delta_k} - x_{\tau_k} = \gamma\delta_k \stackrel{ind}{\sim} Exp(1/\gamma 10)
+x_{\tau_k} - x_{\tau_{k-1}} = \gamma\delta_k \stackrel{ind}{\sim} Exp(1/\gamma 10)
 \]
@@
 \[
-P\left(x_{\tau_k + \delta_k} - x_{\tau_k} > 1\right) = \int_1^\infty 1/10\gamma e^{-1/10\gamma x}dx = e^{-1/10\gamma}.
+P\left(x_{\tau_k} - x_{\tau_{k-1}} > 1\right) = \int_1^\infty 1/10\gamma e^{-1/10\gamma x}dx = e^{-1/10\gamma}.
 \]
@@
 \[
-\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{x^{(i)}_{\tau_{k}^{(i)}+\delta_k^{(i)}} - x^{(i)}_{\tau_k^{(i)}} > 1\right\}
+\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{x^{(i)}_{\tau_{k}^{(i)}} - x^{(i)}_{\tau_{k-1}^{(i)}} > 1\right\}
 \]
@@
 A key question is whether the assumption of a constant transaction rate is approximately correct on a short enough time-scale.
-In order to test this assumption we fit a the gamma distribution to the block sizes of various pieces of the block chain.
+In order to test this assumption we fit a gamma distribution to the block sizes of various pieces of the block chain.
 We use the shape-rate parameterization of the gamma distribution so that if $x\sim G(\alpha,\beta)$, it has the density
@@
 Since the $G(1,\beta)$ and $Exp(\beta)$ distributions are identical, if we find that $\alpha\neq 1$ then transaction rates cannot be constant.
-We focus on the most recent 5000 blocks in our database, which is in the process of downloading the entire chain, ending with block 303151.
+We focus on the most recent 5000 blocks in our database ending with block 303151 --- we are in the process of downloading the entire chain.
 For priors we will assume $\log \alpha \sim N(0, 10)$ and $\beta \sim G(1, 1)$.
@@
 
-The second major constraint that the block chain imposes on us is that transaction data is only measured once a block is completed.
+The second major constraint that the block chain imposes on us is that transaction data is only measured once a block arrives.
 So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error.
 These two constraints motivate a class of models for modeling Bitcoin transaction data.
-Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time it is the current block so that $\tau_{k+1}=\tau_k + \delta_k$, and now $t_k$ the observed arrival time in the blockchain.
-Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k} + \delta_k} - x_{\tau_k}$ denotes the amount of data in block $k$.
+Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time between block $k$ and block $k-1$ so that $\tau_{k}=\tau_{k-1} + \delta_k$, and now let $t_k$ denote the observed arrival time in the blockchain.
+Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k}} - x_{\tau_{k-1}}$ denotes the amount of data in block $k$.
 Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing, infinitely divisible\footnote{Should we define?} stochastic process.
@@
 It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error.
-This simplifies Markov chain Monte Carlo methods that may be used to fit the model by allowing us to exactly simulate $x_{\tau_k + \delta_k}$ without resorting to approximation methods.
+This simplifies Markov chain Monte Carlo methods that may be used to fit the model by allowing us to exactly simulate $x_{\tau_{k-1} + \delta_k}$ without resorting to approximation methods.
 The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g.
@@
 However the gamma causes numerical problems at times since it forces the latent measurement times, $\tau_k$ to enter the gamma function.
-A similar model that avoids this problem is the lognormal distribution, which we will use.
+A similar model that avoids numerical problems is the lognormal distribution, which we will use.
 
@@
 t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
-x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\gamma\delta_k,\beta)\\
-\delta_{k} = \tau_{k+1} - \tau_{k} &\stackrel{iid}{\sim} Exp(1/10)
+x_{\tau_{k-1} + \delta_k} - x_{\tau_{k-1}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\gamma\delta_k,\beta)\\
+\delta_{k} = \tau_{k} - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
 \end{align*}
 where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$.
-Note that $x_{\tau_{k+1}} - x_{\tau_k}$ depends on $\delta_k$ because the $k$'th block arrives at $\tau_k$ and departs at $\tau_{k+1}=\tau_k + \delta_k$ so that block $k$ has $\delta_k$ minutes to accumulate transaction data.
 To complete the model we assume $\sigma$, $\beta$, and $\gamma$ are independent in the prior with $\sigma \sim Cauchy(0, 2.5)$, $\beta \sim G(1, 1)$, and $\gamma \sim LN(0, 10)$.
@@
 Specifically, given access to an approximate simulation of the posterior of the model parameters, we can simulate the model for each draw from the posterior in order to estimate the probability of a block filling up --- this is the posterior predictive distribution.\footnote{Cite Gelman et al.
-text here} In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ by some amount $\psi$ so that the posterior mean of the shifted distribution of $\gamma$ is the desired amount.
+text here} In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ so that the posterior mean of the shifted posterior is a desired value.
 This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different.
-We estimate the probability of a block filling using all three approaches, but because of the required assumptions the estimates become less credible for values of $\gamma$ or $\bar{\gamma}$ further away from the estimated values.
+We estimate the probability of a block filling using all three approaches, but because of the required assumptions the estimates become less credible for values of $\gamma$ or $\mathrm{E}[\gamma]$ further away from the estimated values.
 

### 2015-10-13 model/modelfits.R
Added a csv file of the posterior of just the model parameters, not the latent times, so that it's on github

@@ -90,15 +90,20 @@ fit <- stan(file = 'lognormalgamma.stan', data = gamlist, iter=1, chains=1)
 
 system.time(fit2 <- stan(fit = fit, data = gamlist, iter=4000, chains=4, control = list(adapt_delta = 0.9, max_treedepth = 13)))
 
+## not stored on github because it's large, csv file of just model pars below is.
 save(fit2, file="loggammafit.RData")
-
-load("loggammafit.RData")
+load("loggammafit.RData") 
 
 rstan::traceplot(fit2, pars=c('gamma', 'beta', 'sigma', 'lp__'), ncol=1)
 
 print(fit2, pars=c('gamma', 'beta', 'sigma', 'lp__'), digits = 4)
 
+## save just the model parameters so we don't have to keep fitting the model
 fitex <- extract(fit2)
+modelpar <- cbind(gamma=fitex$gamma, beta=fitex$beta, sigma=fitex$sigma, lp__=fitex$lp__)
+write.csv(modelpar, file="postpar.csv", row.names = FALSE)
+
+fitex <- read.csv("postpar.csv")
 
 niter <- 8000
 parout <- cbind(gamma=fitex$gamma, beta=fitex$beta, sigma=fitex$sigma)

### 2015-10-16 outline/outline.tex
Added a couple details about the quality of the probability estimates under various counterfactuals.

@@
 In order to take into account seasonality, we can make $\gamma$ change in a predictable, seasonal pattern.
+This approach directly specifies the mean transaction rate at $\gamma$, but because of the mean-variance relationship of the gamma distribution it also indirectly specifies the variance of the transaction rate at $\gamma/\hat{beta}$ where $\hat{\beta}$ is an estimate of $\beta$.
 
@@
 Specifically, given access to an approximate simulation of the posterior of the model parameters, we can simulate the model for each draw from the posterior in order to estimate the probability of a block filling up --- this is the posterior predictive distribution.\footnote{Cite Gelman et al.
-text here} In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ so that the posterior mean of the shifted posterior is a desired value.
+text here} In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ so that the posterior mean of the shifted posterior is the desired value.
 This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different.
-We estimate the probability of a block filling using all three approaches, but because of the required assumptions the estimates become less credible for values of $\gamma$ or $\mathrm{E}[\gamma]$ further away from the estimated values.
+Essentially this requires that if we were in a higher transaction rate regime, our priors would have been different such that the posterior distribution would have obtained would be the shifted posterior we construct.
+Explicitly constructing this prior would be involved and maybe intractable or even impossible.
+Instead we estimate the probability of a block filling using all three approaches, but acknowledge that the estimates become less credible for values of $\gamma$ or $\mathrm{E}[\gamma]$ further away from their estimated values.
 

### 2015-10-20 paper/paper.tex
Added intro, new file

--- 2015-10-16 outline/outline.tex
@@
 %\usepackage{mathtools}
+\usepackage[noblocks]{authblk}
 
-\author{Matt Simpson}
-\title{Outline of Ideas for BTC Transaction Frequency}
+\author[1]{Matt Simpson}
+\author[2]{Eli Dourado}
+\affil[1]{M. Simpson is a postdoctoral fellow in statistics at the University of Missouri}
+\affil[2]{E.
+Dourado is a research fellow at the Mercatus Center at George Mason University and director of its Technology Policy Program, as well as a doctoral candidate in economics at George Mason University}
+\title{Estimating the distribution of block sizes as a function of average transaction rates}
+\renewcommand\Authand{, }
 \begin{document}
@@
 We also estimate how that percentage changes when transaction rates exhibit minute-to-minute variability along specified, empirically defensible distributions.}
+
+\section{Introduction}
+The optimal maximum size of Bitcoin blocks is an area of active controversy.
+When the maximum size is set at a high level, mining centralization will increase due to higher resource demands, and the network becomes vulnerable to selfish mining attacks.\footnote{http://arxiv.org/abs/1311.0243} When the maximum size is too small, there is inadequate space to accommodate the volume of transactions that users demand, resulting in higher fees, wait times, and potentially lower rates of adoption, which affects security through the price of Bitcoin.
+The issue may be with us for some time; even assuming that current disagreements over raising the 1MB cap are resolved, if the Bitcoin network keeps growing and as computing power increases, the issue will need to be re-litigated from time to time.
+One proposal, BIP 100, would subject the decision to a continuous voting process by miners indefinitely into the future.
+
+A better understanding of the relationship between the rate of Bitcoin transactions and block sizes is needed to inform this debate.
+Bitcoin blocks are discovered according to a Poisson process, directly implying that times between block discoveries are distributed exponentially.
+If transaction rates (in bytes per second) were constant, this would imply that block sizes too would be distributed exponentially.
+As the exponential distribution is well understood and has a closed-form cumulative density function, it would be trivial to evaluate the probability of blocks filling up as a function of the average transaction rate.
+
+In fact, transaction rates are not constant.
+Block sizes are therefore not distributed exponentially, but according to some other distribution.
+In this paper, we empirically estimate the distribution of block sizes and calculate the probability that a block will reach the upper bound as a function of the average transaction rate.
+We find that the actual distribution of transaction data is more likely to result in a binding maximum than an exponential distribution would across the spectrum of average transaction rates, although the biggest effect is at the relatively low end.
 

### 2015-10-20 postgres.R
Minor tweaks

@@ -45,13 +45,14 @@ dbClearResult(res)
 
 
 #let's see the timestamp of the most recent block in our database
-res <- dbSendQuery(con, "select time from blocks order by height desc limit 1")
+res <- dbSendQuery(con, "select time, height from blocks order by height desc limit 1")
 while(!dbHasCompleted(res)){
   chunk <- dbFetch(res, n = 5)
   print(nrow(chunk))
 }
 time <- dbFetch(res)
-as.POSIXct(time$time[1], origin="1970-01-01")
+print(time$height[1])
+print(as.POSIXct(time$time[1], origin="1970-01-01"))
 dbClearResult(res)
 
 
@@ -94,6 +95,30 @@ blockstats1k$bytespermin <- blockstats1k$size*60/blockstats1k$elapsed_time
 library(Hmisc)
 describe(blockstats1k)
 plot(blockstats1k$elapsed_time, blockstats1k$size, main="Block size v. time between blocks", xlab="Seconds since prior block discovery", ylab="bytes", pch=19)
+
+
+#let's see the 100 largest blocks in our database
+res <- dbSendQuery(con, "select height, size from blocks order by size desc limit 100")
+while(!dbHasCompleted(res)){
+  chunk <- dbFetch(res, n = 5)
+  print(nrow(chunk))
+}
+size <- dbFetch(res)
+print(size)
+dbClearResult(res)
+
+
+#let's scatterplot all the blocks in our database by height and size
+res <- dbSendQuery(con, "select height, size from blocks")
+while(!dbHasCompleted(res)){
+  chunk <- dbFetch(res, n = 5)
+  print(nrow(chunk))
+}
+heightsize <- dbFetch(res)
+dbClearResult(res)
+plot(heightsize$height, heightsize$size, main="Block Sizes", 
+   xlab="Block Height ", ylab="Size in Bytes ", pch=19)
+
 
 
 

### 2015-10-21 paper/paper.tex
Some minor edits plus adding citations

@@
 \usepackage{bm, bbm}
-%\usepackage{mathtools}
+\usepackage[authoryear]{natbib} 
 \usepackage[noblocks]{authblk}
@@
 The optimal maximum size of Bitcoin blocks is an area of active controversy.
-When the maximum size is set at a high level, mining centralization will increase due to higher resource demands, and the network becomes vulnerable to selfish mining attacks.\footnote{http://arxiv.org/abs/1311.0243} When the maximum size is too small, there is inadequate space to accommodate the volume of transactions that users demand, resulting in higher fees, wait times, and potentially lower rates of adoption, which affects security through the price of Bitcoin.
+When the maximum size is set at a high level, mining centralization will increase due to higher resource demands, and the network becomes vulnerable to selfish mining attacks\citep{eyal2014majority}.%\footnote{http://arxiv.org/abs/1311.0243}
+ When the maximum size is too small, there is inadequate space to accommodate the volume of transactions that users demand, resulting in higher fees, wait times, and potentially lower rates of adoption, which affects security through the price of Bitcoin.
 The issue may be with us for some time; even assuming that current disagreements over raising the 1MB cap are resolved, if the Bitcoin network keeps growing and as computing power increases, the issue will need to be re-litigated from time to time.
@@
 Sample size is indicated by $n$, group id by $g$.
-Each model was fit in RStan\protect\footnotemark using 4 chains, each with 4000 iterations, 2000 of which used for warmup/burn in.}
+Each model was fit in RStan \citep{rstan-software:2015,stan-software:2015} using 4 chains, each with 4000 iterations, 2000 of which used for warmup/burn in.}
 \label{tab:gammafit}
@@
 
-For\footnotetext{Cite RStan} the full 5000 blocks, the posterior mean of $\alpha$ is 1.34 and the lower bound of the 95\% credible interval is 1.29, so we have strong evidence that $\alpha$ is nonzero and that a constant transaction rate is not appropriate.
+For the full 5000 blocks, the posterior mean of $\alpha$ is 1.34 and the lower bound of the 95\% credible interval is 1.29, so we have strong evidence that $\alpha$ is nonzero and that a constant transaction rate is not appropriate.
 This is not surprising since Bitcoin usage has gone up over time, but on a short enough time scale usage may be approximately constant.
@@
 Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k}} - x_{\tau_{k-1}}$ denotes the amount of data in block $k$.
-Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing, infinitely divisible\footnote{Should we define?} stochastic process.
+Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing stochastic process.
 Most of the interesting modeling choices come from modeling this stochastic process.
 Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero.
-Instead, we consider the classes of non-negative infinite activity\footnote{Should we define?} L\'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$.\footnote{Cite the Levy process chapter here}.
+Instead, we consider non-negative infinite activity L'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$.
+The following discussion is brief -- see e.g.
+\citet{barndorff2012basics} for a fuller treatment.
 The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$.
@@
 
-While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$.
+The gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$.
 Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$.
-The inverse Gaussian process is another L\'{e}vy process which has a closed form solution for the distribution of $x_\tau$.
+The inverse Gaussian process is another L\'{e}vy process which has a closed form solution for the distribution of $x_\tau$ and for which $x_\tau$ is inverse Gaussian distributed for all $\tau>0$.
 The inverse Gaussian density is
@@
 seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the parameter increments.
-We will use the gamma process since it is widely available in statistical packages, unlike the main alternative, the inverse Gaussian.
+We will use the gamma process since it is widely available in statistical packages and in particular is available in Stan \citep{stan-software:2015}, unlike the main alternative, the inverse Gaussian.
 
@@
  \caption{Results of fitting the measurement error model with stochastic measurement times.
-We use RStan to fit the model, obtaining four chains each of 4000 iterations, 2000 of which used for warmup/burn in.\protect\footnotemark (Note: outdated numbers, updated ones coming) }
+We use RStan \citep{rstan-software:2015,stan-software:2015} to fit the model, obtaining four chains each of 4000 iterations, 2000 of which used for warmup/burn in.}
  \label{tab:modelfit}
@@
 
-Given\footnotetext{Cite RStan} an approximate simulation of the posterior distribution of the model parameters, we can use Monte Carlo simulation to estimate the probability of a block filling up in a number of ways.
+Given an approximate simulation of the posterior distribution of the model parameters, we can use Monte Carlo simulation to estimate the probability of a block filling up in a number of ways.
 The simplest way is to compute parameter estimates based on the posterior, e.g.
@@
 We can mitigate 1) while making 2) if not worse, different.
-Specifically, given access to an approximate simulation of the posterior of the model parameters, we can simulate the model for each draw from the posterior in order to estimate the probability of a block filling up --- this is the posterior predictive distribution.\footnote{Cite Gelman et al.
-text here} In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ so that the posterior mean of the shifted posterior is the desired value.
+Specifically, given access to an approximate simulation of the posterior of the model parameters, we can simulate the model for each draw from the posterior in order to estimate the probability of a block filling up --- this is the posterior predictive distribution \citep{gelman2014bayesian}.
+In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ so that the posterior mean of the shifted posterior is the desired value.
 This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different.
@@
 
+\bibliographystyle{apalike} 
+\bibliography{btc}
+
 \end{document}

### 2015-10-21 paper/paper.tex
Renamed sections

@@
 
-\section{Motivation}
-As Bitcoin becomes more popular, the transaction rate increases.
-Since existing Bitcoin rules limit the amount of transactions per block to one MB at some point the blocks will begin filling up.
-Bitcoin blocks are by design distributed according a Poisson process with a rate of discovery of one block every 10 minutes.
+\section{Verifying the variability of transaction rates}
 Let $\tau$ index time in minutes, and $N(\tau)$ denote the number of blocks discovered by time $\tau$.
@@
 
-\section{The Model}
+\section{Modeling the distribution of block sizes}
 

### 2015-10-22 paper/paper.tex
Typos, edits, a few queries, discussion

@@
 
-The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how how the amount of transaction data is distributed requires more work.
+The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how transaction data is distributed requires more work.
 Crucial constraints on the models we can fit come from the data obtained from the block chain.
@@
 Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero.
-Instead, we consider non-negative infinite activity L'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$.
+Instead, we consider non-negative infinite activity L\'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$.
 The following discussion is brief -- see e.g.
@@
 
-The gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$.
+The gamma process is particularly convenient since the distribution of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$.
 Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$.
@@
 For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision.
-This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$.
+This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the conditional distribution of $t_k$.
 A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative.
@@
 In order to take into account seasonality, we can make $\gamma$ change in a predictable, seasonal pattern.
-This approach directly specifies the mean transaction rate at $\gamma$, but because of the mean-variance relationship of the gamma distribution it also indirectly specifies the variance of the transaction rate at $\gamma/\hat{beta}$ where $\hat{\beta}$ is an estimate of $\beta$.
+This approach directly specifies the mean transaction rate at $\gamma$, but because of the mean-variance relationship of the gamma distribution it also indirectly specifies the variance of the transaction rate at $\gamma/\hat{\beta}$ where $\hat{\beta}$ is an estimate of $\beta$.
 
@@
 This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different.
-Essentially this requires that if we were in a higher transaction rate regime, our priors would have been different such that the posterior distribution would have obtained would be the shifted posterior we construct.
+Essentially this requires that if we were in a higher transaction rate regime, our priors would have been different such that the posterior distribution would have obtained would be the shifted posterior we construct [IS THIS SENTENCE RIGHT OR IS THERE A GRAMMATICAL ERROR?].
 Explicitly constructing this prior would be involved and maybe intractable or even impossible.
@@
 
-Under a fixed mean transaction rate, the probability of a full block increases as as $\beta$ decreases, especially for smaller mean transaction rates.
+Under a fixed mean transaction rate, the probability of a full block increases as $\beta$ decreases, especially for smaller mean transaction rates.
 For larger mean transaction rates there is little effect on the probability the next block is full from decreasing $\beta$.
@@
 What is unexpected is that taking into account the uncertainty in $\beta$ and then in both $\gamma$ and $\beta$ seems to have essentially no effect on the probability that the next block fills up no matter the expected value of $\gamma$.
+[CAN YOU CLARIFY THIS LAST SENTENCE?]
 
-These estimates seem rather high considering no block have filled up yet.
+%% These estimates seem rather high considering no block have filled up yet.
 Transaction rates have been fluctuating around 0.05 for several months and even under constant transaction rates, the probability of a full block is over 0.1.
@@
 
+%% [A PLACEHOLDER FOR NEXT DRAFT IN WHICH WE DISCUSS SOFT LIMITS AND HOW THEY AFFECT OUR MODEL?]
+
+\section{Discussion}
+Our analysis does not resolve the contentious block size issue, but it is an important input into weighing the tradeoff between supporting higher transaction rates and the security issues created by larger blocks.
+For example, at the existing 1 MB maximum, as average transaction rates increase from 500 KB/block to 600 KB/block, an additional five percent or so of blocks will hit the limit.
+These full blocks signal that not all transactions are being included in the first block after their transmission.
+
+In addition, a greater understanding of the dynamics of full blocks versus the average transaction rates could be used to help clients set fees appropriately, as a function of the observed rate of transactions over the last several blocks and the existing maximum block size.
+Creating a more dynamic fee market is another way, besides simply increasing the maximum block size, to scale the network effectively.
+
 \bibliographystyle{apalike} 

### 2015-10-22 paper/paper.tex
Fixed some typos and improved some explanation

--- 2015-10-21 paper/paper.tex
@@
 The optimal maximum size of Bitcoin blocks is an area of active controversy.
-When the maximum size is set at a high level, mining centralization will increase due to higher resource demands, and the network becomes vulnerable to selfish mining attacks\citep{eyal2014majority}.%\footnote{http://arxiv.org/abs/1311.0243}
+When the maximum size is set at a high level, mining centralization will increase due to higher resource demands, and the network becomes vulnerable to selfish mining attacks \citep{eyal2014majority}.%\footnote{http://arxiv.org/abs/1311.0243}
  When the maximum size is too small, there is inadequate space to accommodate the volume of transactions that users demand, resulting in higher fees, wait times, and potentially lower rates of adoption, which affects security through the price of Bitcoin.
@@
 Let $\tau$ index time in minutes, and $N(\tau)$ denote the number of blocks discovered by time $\tau$.
-Then for $\delta > 0$, the number of blocks discovered between time $\tau$ and time $\tau + \delta$ is Poisson distributed, that is $N(\tau + \delta) - N(\tau) \sim Poi(\delta/10)$.
+Then for an increment $\delta > 0$, the number of blocks discovered between time $\tau$ and time $\tau + \delta$ is Poisson distributed, that is $N(\tau + \delta) - N(\tau) \sim Poi(\delta/10)$.
 Let $k=1,2,\dots$ denote each of the blocks, in order.
-Then the amount of time between blocks is exponentially distributed.
-Let $\tau_k$ denote the the time when block $k$ arrives.
-Then $\delta_k \equiv \tau_{k} - \tau_{k-1} \stackrel{iid}{\sim} Exp(1/10)$ for $k=1,2,\dots,K$, where if $x\sim Exp(\lambda)$ it has the density $p(x) = \lambda e^{-x\lambda}$ for $x>0$ and $\lambda>0$ with mean $1/\lambda$.
+Then the amount of time between blocks is exponentially distributed, i.e.
+$\delta_k \equiv \tau_{k} - \tau_{k-1} \stackrel{iid}{\sim} Exp(1/10)$ for $k=1,2,\dots,K$ where $\tau_k$ is the arrival time of the $k$'th block.
+We use the rate parameterization of the exponential distribution so that if $x\sim Exp(\lambda)$ it has the density $p(x) = \lambda e^{-x\lambda}$ for $x>0$ and $\lambda>0$ with mean $1/\lambda$.
 The longer $\delta_k$ the more time for transactions to add up before being added to the block chain as part of block $k$.
@@
 \]
-for $x>0$ with mean $\alpha/\beta$ and variance $\alpha/\beta^2$.
+for $x>0$, $\alpha>0$, and $\beta>0$ with mean $\alpha/\beta$ and variance $\alpha/\beta^2$.
+The gamma function in the denominator is defined as $\Gamma(x) = \int_0^\infty x^{t-1}e^{-x}dx$.
 Since the $G(1,\beta)$ and $Exp(\beta)$ distributions are identical, if we find that $\alpha\neq 1$ then transaction rates cannot be constant.
@@
 
-The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how how the amount of transaction data is distributed requires more work.
+The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how the amount of transaction data is distributed requires more work.
 Crucial constraints on the models we can fit come from the data obtained from the block chain.
@@
 Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero.
-Instead, we consider non-negative infinite activity L'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$.
+Instead, we consider nonnegative infinite activity L\'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$.
 The following discussion is brief -- see e.g.
@@
 Brownian motion is the most common example.
-A nondecreasing example is the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$.
+A nondecreasing example is the gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$.
+The gamma process is called an infinite activitive process since for any time $\tau$ and increment $\delta>0$, $x_{\tau + \delta} - x_{\tau} > 0$.
 
 The gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$.
+More generally the gamma distribution is infinitely divisible so that if $x\sim G(\alpha,\beta)$ then for any integer $n>0$ there exist iid random variables $y_1,y_2,\dots,y_n$ such that $y_1+y_2+\dots+y_n$ and $x$ have the same distribution.
 Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$.
-The inverse Gaussian process is another L\'{e}vy process which has a closed form solution for the distribution of $x_\tau$ and for which $x_\tau$ is inverse Gaussian distributed for all $\tau>0$.
+The inverse Gaussian process is another L\'{e}vy process for which the distribution of $x_\tau$ is in the same class as the distribution for $x_1$ for all $\tau>0$.
 The inverse Gaussian density is
@@
 
-It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error.
-This simplifies Markov chain Monte Carlo methods that may be used to fit the model by allowing us to exactly simulate $x_{\tau_{k-1} + \delta_k}$ without resorting to approximation methods.
+Since the observation times are observed with error, it is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form for any $tau$ and $\delta>0$.
+This simplifies fitting the model because it allows us to write down a relatively simple exact likelihood function, facilitating both maximum likelihood estimation and Markov chain Monte Carlo methods.
 The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g.
@@
 However the gamma causes numerical problems at times since it forces the latent measurement times, $\tau_k$ to enter the gamma function.
-A similar model that avoids numerical problems is the lognormal distribution, which we will use.
+A similar distribution that avoids numerical problems is the lognormal, which we will use.
 
@@
 In order to take into account seasonality, we can make $\gamma$ change in a predictable, seasonal pattern.
-This approach directly specifies the mean transaction rate at $\gamma$, but because of the mean-variance relationship of the gamma distribution it also indirectly specifies the variance of the transaction rate at $\gamma/\hat{beta}$ where $\hat{\beta}$ is an estimate of $\beta$.
+This approach directly specifies the mean transaction rate at $\gamma$, but because of the mean-variance relationship of the gamma distribution it also indirectly specifies the variance of the transaction rate at $\gamma/\hat{\beta}$ where $\hat{\beta}$ is an estimate of $\beta$.
 
 This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\gamma$ increases.
+In particular, 2) implies that the variance of transaction rates is always proportional to the mean, i.e.
+it is $\gamma/\beta$.
 We can mitigate 1) while making 2) if not worse, different.
@@
 This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different.
+In other words, it still assumes that the variance of transaction rates is still $\gamma/\beta$, but now $\gamma$ and $\beta$ are also stochastic with $\beta$ distributed according to its marginal posterior distribution and $\gamma|\beta$ distributed according to a shifted version of its conditional posterior given $\beta$.
+
  Essentially this requires that if we were in a higher transaction rate regime, our priors would have been different such that the posterior distribution would have obtained would be the shifted posterior we construct.
@@
 When block sizes are on average much smaller than 1 MB, increasing the variance of block sizes puts significantly more probability mass above 1 MB, but when the mean is near but still below 1 MB it has a relatively small effect.
-What is unexpected is that taking into account the uncertainty in $\beta$ and then in both $\gamma$ and $\beta$ seems to have essentially no effect on the probability that the next block fills up no matter the expected value of $\gamma$.
+What is unexpected is that taking into account the uncertainty in $\beta$ and then in both $\gamma$ and $\beta$ seems to have essentially no effect on the probability that the next block fills given a particular value for the expected value of $\gamma$.
 
-These estimates seem rather high considering no block have filled up yet.
-Transaction rates have been fluctuating around 0.05 for several months and even under constant transaction rates, the probability of a full block is over 0.1.
-Even at a constant transaction rate of 0.02, the probability of a full block is 0.067.
-In 100 blocks, which is approximately 17 hours, the probability of at least one block filling up is $1-(1-0.067)^{100}\approx 1$.
-Since no block has filled up yet, clearly something is happening that prevents blocks from filling up.
-One possibility is that transaction rates are usually low and in no danger of causing a full block, but occasionally there are large spikes in the rate.
-Another possibility is strategic action by Bitcoin traders and miners to prevent blocks from filling up.
-Traders may be waiting to trade until a new block replaces a nearly full block, which is not much of a cost since the wait is only about 10 minutes on average.
-Similarly miners have an incentive to keep blocks small since that reduces the change of an orphan block --- a mined block which is not accepted by the blockchain --- so the miner of the current block may be rejecting some transactions in order to prevent the limit from being reached.
+These estimates may seem rather high considering blocks only recently began filling up.
+Recent transaction rates have been fluctuating around 0.05 for several months and even under constant transaction rates, the probability of a full block is over 0.1.
+With a constant transaction rate of 0.02, the probability of a full block is 0.067.
+This seems small, but in 100 blocks, which is approximately 17 hours, the probability of at least one block filling up is $1-(1-0.067)^{100}\approx 1$.
+The apparent paradox is solved by acknowledging a soft limit that Bitcoin miners impose on block sizes.
+While the Bitcoin protocol limits block sizes to 1 MB, Bitcoin miners have the ability to set their own block size limit lower.
+Miners have an incentive to keep blocks small since that reduces the chance of an orphan block --- a mined block which is not accepted by the blockchain --- so they tend to exercise this option.
+Further, the default setting is to limit block sizes to 750 KB.
 

### 2015-10-22 paper/paper.tex

@@
 
-The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how the amount of transaction data is distributed requires more work.
+The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how transaction data is distributed requires more work.
 Crucial constraints on the models we can fit come from the data obtained from the block chain.
@@
 
-The gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$.
+The gamma process is particularly convenient since the distribution of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$.
 More generally the gamma distribution is infinitely divisible so that if $x\sim G(\alpha,\beta)$ then for any integer $n>0$ there exist iid random variables $y_1,y_2,\dots,y_n$ such that $y_1+y_2+\dots+y_n$ and $x$ have the same distribution.
@@
 The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g.
-seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the parameter increments.
+seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the transaction data increments.
 We will use the gamma process since it is widely available in statistical packages and in particular is available in Stan \citep{stan-software:2015}, unlike the main alternative, the inverse Gaussian.
@@
 For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision.
-This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$.
+This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the conditional dsitribution of $t_k$.
 A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative.
@@
 In other words, it still assumes that the variance of transaction rates is still $\gamma/\beta$, but now $\gamma$ and $\beta$ are also stochastic with $\beta$ distributed according to its marginal posterior distribution and $\gamma|\beta$ distributed according to a shifted version of its conditional posterior given $\beta$.
-
- Essentially this requires that if we were in a higher transaction rate regime, our priors would have been different such that the posterior distribution would have obtained would be the shifted posterior we construct.
-Explicitly constructing this prior would be involved and maybe intractable or even impossible.
+Essentially this assumption requires that if we were in a higher transaction rate regime, our priors would have been different such that the posterior distribution we would have obtained would be the shifted posterior we construct.
+Explicitly constructing this prior would be involved and maybe intractable or even theoretically impossible.
 Instead we estimate the probability of a block filling using all three approaches, but acknowledge that the estimates become less credible for values of $\gamma$ or $\mathrm{E}[\gamma]$ further away from their estimated values.
@@
 When block sizes are on average much smaller than 1 MB, increasing the variance of block sizes puts significantly more probability mass above 1 MB, but when the mean is near but still below 1 MB it has a relatively small effect.
-What is unexpected is that taking into account the uncertainty in $\beta$ and then in both $\gamma$ and $\beta$ seems to have essentially no effect on the probability that the next block fills given a particular value for the expected value of $\gamma$.
+What is unexpected is that no matter what the expected mean transaction rate is, the probability of a block filling up is largely unaffected by taking into account parameter uncertainty in $\beta$ or in both $\beta$ and $\gamma$, which can be seen from the final three lines of Table \ref{tab:probsims}.
 
-These estimates may seem rather high considering blocks only recently began filling up.
-Recent transaction rates have been fluctuating around 0.05 for several months and even under constant transaction rates, the probability of a full block is over 0.1.
-With a constant transaction rate of 0.02, the probability of a full block is 0.067.
-This seems small, but in 100 blocks, which is approximately 17 hours, the probability of at least one block filling up is $1-(1-0.067)^{100}\approx 1$.
-The apparent paradox is solved by acknowledging a soft limit that Bitcoin miners impose on block sizes.
-While the Bitcoin protocol limits block sizes to 1 MB, Bitcoin miners have the ability to set their own block size limit lower.
-Miners have an incentive to keep blocks small since that reduces the chance of an orphan block --- a mined block which is not accepted by the blockchain --- so they tend to exercise this option.
-Further, the default setting is to limit block sizes to 750 KB.
+%% In the next draft we can see what happens when the mean transaction rate varies deterministically over time -- e.g.
+through some prespecified seasonal pattern, then see how that impacts not the probability of the next block filling up, but one of the next 10 blocks, for example.
+
+%% These estimates seem rather high considering no block have filled up yet.
+Transaction rates have been fluctuating around 0.05 for several months and even under constant transaction rates, the probability of a full block is over 0.1.
+Even at a constant transaction rate of 0.02, the probability of a full block is 0.067.
+In 100 blocks, which is approximately 17 hours, the probability of at least one block filling up is $1-(1-0.067)^{100}\approx 1$.
+Since no block has filled up yet, clearly something is happening that prevents blocks from filling up.
+One possibility is that transaction rates are usually low and in no danger of causing a full block, but occasionally there are large spikes in the rate.
+Another possibility is strategic action by Bitcoin traders and miners to prevent blocks from filling up.
+Traders may be waiting to trade until a new block replaces a nearly full block, which is not much of a cost since the wait is only about 10 minutes on average.
+Similarly miners have an incentive to keep blocks small since that reduces the change of an orphan block --- a mined block which is not accepted by the blockchain --- so the miner of the current block may be rejecting some transactions in order to prevent the limit from being reached.
+
+%% [A PLACEHOLDER FOR NEXT DRAFT IN WHICH WE DISCUSS SOFT LIMITS AND HOW THEY AFFECT OUR MODEL?]   Matt: sounds good.
+
+vvvvvvvvvvvvvvvvvvvv
+\section{Discussion}
+Our analysis does not resolve the contentious block size issue, but it is an important input into weighing the tradeoff between supporting higher transaction rates and the security issues created by larger blocks.
+For example, at the existing 1 MB maximum, as average transaction rates increase from 500 KB/block to 600 KB/block, an additional five percent or so of blocks will hit the limit.
+These full blocks signal that not all transactions are being included in the first block after their transmission.
+
+In addition, a greater understanding of the dynamics of full blocks versus the average transaction rates could be used to help clients set fees appropriately, as a function of the observed rate of transactions over the last several blocks and the existing maximum block size.
+Creating a more dynamic fee market is another way, besides simply increasing the maximum block size, to scale the network effectively.
+^^^^^^^^^^^^^^^^^^^^
 

### 2015-10-22 paper/paper.tex
Ran a spellcheck and made minor changes

@@
 \]
-This probability depends crucially on the process that $x_\tau$ follows. If we assume that $x_{\tau}$ is determinstic with a constant transaction rate of $\gamma$ MB/min, then 
+This probability depends crucially on the process that $x_\tau$ follows. If we assume that $x_{\tau}$ is deterministic with a constant transaction rate of $\gamma$ MB/min, then 
 \[
@@
 Crucial constraints on the models we can fit come from the data obtained from the block chain.
-Since Bitcoin is a distributed network without a home node, there is no official timestamp for when each block arrives on the chain.
+Since Bitcoin is a distributed network without a home node, there is no official time-stamp for when each block arrives on the chain.
 Instead, the block miner's local time is recorded.
@@
 Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node with a maximum offset of 70 minutes.
-Because of the nature of the blocks' timestamps, occasionally a block has a timestamp which is earlier than one or more previous blocks in the chain.
-Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid timestamps define lower and upper bounds for the measurement error distribution.
+Because of the nature of the blocks' time-stamps, occasionally a block has a time-stamp which is earlier than one or more previous blocks in the chain.
+Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid time-stamps define lower and upper bounds for the measurement error distribution.
 Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the lower bound is at our disposal.
@@
 \citet{barndorff2012basics} for a fuller treatment.
-The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$.
+The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continuous time and has independent and stationary increments with $x_0=0$.
 Brownian motion is the most common example.
 A nondecreasing example is the gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$.
-The gamma process is called an infinite activitive process since for any time $\tau$ and increment $\delta>0$, $x_{\tau + \delta} - x_{\tau} > 0$.
+The gamma process is called an infinite activity process since for any time $\tau$ and increment $\delta>0$, $x_{\tau + \delta} - x_{\tau} > 0$.
 
@@
 For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision.
-This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the conditional dsitribution of $t_k$.
+This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the conditional distribution of $t_k$.
 A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative.
@@
 
-vvvvvvvvvvvvvvvvvvvv
 \section{Discussion}
@@
 These full blocks signal that not all transactions are being included in the first block after their transmission.
-
 In addition, a greater understanding of the dynamics of full blocks versus the average transaction rates could be used to help clients set fees appropriately, as a function of the observed rate of transactions over the last several blocks and the existing maximum block size.
 Creating a more dynamic fee market is another way, besides simply increasing the maximum block size, to scale the network effectively.
-^^^^^^^^^^^^^^^^^^^^
 

### 2015-10-22 paper/paper.tex
Made some final changes before the review and also added the pdf file to the repository

@@
 \[
-x_{\tau_k} - x_{\tau_{k-1}} = \gamma\delta_k \stackrel{ind}{\sim} Exp(1/\gamma 10)
+x_{\tau_k} - x_{\tau_{k-1}} = \gamma\delta_k \stackrel{ind}{\sim} Exp(1/10\gamma)
 \]
@@
 \[
-P\left(x_{\tau_k} - x_{\tau_{k-1}} > 1\right) = \int_1^\infty 1/10\gamma e^{-1/10\gamma x}dx = e^{-1/10\gamma}.
+P\left(x_{\tau_k} - x_{\tau_{k-1}} > 1\right) = \int_1^\infty \frac{1}{10\gamma}e^{-x/10\gamma}dx = e^{-1/10\gamma}.
 \]
@@
 For the full 5000 blocks, the posterior mean of $\alpha$ is 1.34 and the lower bound of the 95\% credible interval is 1.29, so we have strong evidence that $\alpha$ is nonzero and that a constant transaction rate is not appropriate.
-This is not surprising since Bitcoin usage has gone up over time, but on a short enough time scale usage may be approximately constant.
+This is not surprising since Bitcoin usage has gone up over time.
 A block is around 10 minutes on average, so 5000 blocks takes a little over a month on average.
-It is possible that on a short enough time-scale, constant transaction rates are more appropriate.
+It is possible that on a short enough time-scale, transaction rates are approximately constant.
 So we fit the gamma distribution to 1000 blocks at a time --- representing about a week --- by breaking the original 5000 blocks into five groups.
@@
 Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid time-stamps define lower and upper bounds for the measurement error distribution.
-Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the lower bound is at our disposal.
+Unfortunately network adjusted time is not recorded in the block chain, so the upper bound is not observed, but the lower bound is at our disposal.
 
@@
 These two constraints motivate a class of models for modeling Bitcoin transaction data.
-Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time between block $k$ and block $k-1$ so that $\tau_{k}=\tau_{k-1} + \delta_k$, and now let $t_k$ denote the observed arrival time in the blockchain.
+Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time between block $k$ and block $k-1$ so that $\tau_{k}=\tau_{k-1} + \delta_k$, and now let $t_k$ denote the observed arrival time in the block chain.
 Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k}} - x_{\tau_{k-1}}$ denotes the amount of data in block $k$.
@@
 This simplifies fitting the model because it allows us to write down a relatively simple exact likelihood function, facilitating both maximum likelihood estimation and Markov chain Monte Carlo methods.
-The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g.
+The parameters of the gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g.
 seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the transaction data increments.
@@
 where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$.
-To complete the model we assume $\sigma$, $\beta$, and $\gamma$ are independent in the prior with $\sigma \sim Cauchy(0, 2.5)$, $\beta \sim G(1, 1)$, and $\gamma \sim LN(0, 10)$.
+To complete the model we assume $\sigma$, $\beta$, and $\gamma$ are independent in the prior with $\sigma \sim half-Cauchy(0, 2.5)$, $\beta \sim G(1, 1)$, and $\gamma \sim LN(0, 10)$.
+Note that the half-Caucy distribution is the Cauchy truncated to the positive real line.
 Table \ref{tab:modelfit} contains the results of fitting the model.
@@
   \hline
-$E[\gamma] =$ & 0.02 & 0.03 & 0.04 & 0.05 & 0.06 & 0.07 & 0.08 & 0.09 \\ 
+$E[\gamma] = \hat{\gamma} =$ & 0.02 & 0.03 & 0.04 & 0.05 & 0.06 & 0.07 & 0.08 & 0.09 \\ 
   \hline
@@
 \end{tabular}
-\caption{Probability of a full block under a variety of parameter values.
-In the first row the transaction rate is constant, in the second through fourth rows the transaction rate is stochastic with fixed mean $\gamma$, and in the final row the transaction rate is stochastic and unknown with mean $\mathrm{E}[\gamma]$.
-Under all five regimes the expected mean transaction rate is the same --- it is the spread of the distribution of transaction rates and mean transaction rates that changes across regimes.}
+\caption{Probability of a full block under a variety of parameter values --- $\gamma$ is the mean transaction rate and $\beta$ is a dispersion parameter so that the variance of transaction rates is $\gamma/\beta$.
+In the first row the transaction rate is constant at $\hat{\gamma}$, in the second through fourth rows the transaction rate is stochastic with fixed mean $\hat{\gamma}$, and in the final row the transaction rate is stochastic and unknown with mean $\hat{\gamma}$.}
 \label{tab:probsims}
@@
 When block sizes are on average much smaller than 1 MB, increasing the variance of block sizes puts significantly more probability mass above 1 MB, but when the mean is near but still below 1 MB it has a relatively small effect.
-What is unexpected is that no matter what the expected mean transaction rate is, the probability of a block filling up is largely unaffected by taking into account parameter uncertainty in $\beta$ or in both $\beta$ and $\gamma$, which can be seen from the final three lines of Table \ref{tab:probsims}.
+What is unexpected is that no matter what the expected mean transaction rate is, the probability of a block filling up is largely unaffected by taking into account parameter uncertainty in $\beta$ or in both $\beta$ and $\gamma$, which can be seen from the final three rows of Table \ref{tab:probsims}.
 
@@
 Traders may be waiting to trade until a new block replaces a nearly full block, which is not much of a cost since the wait is only about 10 minutes on average.
-Similarly miners have an incentive to keep blocks small since that reduces the change of an orphan block --- a mined block which is not accepted by the blockchain --- so the miner of the current block may be rejecting some transactions in order to prevent the limit from being reached.
+Similarly miners have an incentive to keep blocks small since that reduces the change of an orphan block --- a mined block which is not accepted by the block chain --- so the miner of the current block may be rejecting some transactions in order to prevent the limit from being reached.
 

### 2015-12-03 postgres.R
Improved plot and backup of main chain

@@ -109,15 +109,29 @@ dbClearResult(res)
 
 
 #let's scatterplot all the blocks in our database by height and size
-res <- dbSendQuery(con, "select height, size from blocks")
+res <- dbSendQuery(con, "select height, size from blocks where branch=0")
 while(!dbHasCompleted(res)){
   chunk <- dbFetch(res, n = 5)
   print(nrow(chunk))
 }
 heightsize <- dbFetch(res)
 dbClearResult(res)
-plot(heightsize$height, heightsize$size, main="Block Sizes", 
-   xlab="Block Height ", ylab="Size in Bytes ", pch=19)
+plot(heightsize$height, heightsize$size, main="Bitcoin block sizes over time", 
+   xlab="Block Height ", ylab="Size in Bytes ", pch=".")
+
+
+
+#let's download all the data
+res <- dbSendQuery(con, "select * from blocks where branch=0")
+while(!dbHasCompleted(res)){
+  chunk <- dbFetch(res, n = 5)
+  print(nrow(chunk))
+}
+blockchain <- dbFetch(res)
+dbClearResult(res)
+blockchain$work <- vapply(blockchain$work, paste, collapse = ", ", character(1L))
+write.table(blockchain, file = "~/Desktop/blockchain.csv", sep = ",", col.names = NA,
+            qmethod = "double")
 
 
 

### 2015-12-04 paper/paper.tex
Fixed some typos

@@
 
-For the full 5000 blocks, the posterior mean of $\alpha$ is 1.34 and the lower bound of the 95\% credible interval is 1.29, so we have strong evidence that $\alpha$ is nonzero and that a constant transaction rate is not appropriate.
+For the full 5000 blocks, the posterior mean of $\alpha$ is 1.34 and the lower bound of the 95\% credible interval is 1.29, so we have strong evidence that $\alpha$ is not one and that a constant transaction rate is not appropriate.
 This is not surprising since Bitcoin usage has gone up over time.
@@
 
-The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how transaction data is distributed requires more work.
+The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining how transaction data is distributed requires more work.
 Crucial constraints on the models we can fit come from the data obtained from the block chain.
@@
 
-Since the observation times are observed with error, it is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form for any $tau$ and $\delta>0$.
+Since the observation times are observed with error, it is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form for any $\tau$ and $\delta>0$.
 This simplifies fitting the model because it allows us to write down a relatively simple exact likelihood function, facilitating both maximum likelihood estimation and Markov chain Monte Carlo methods.
@@
 where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$.
-To complete the model we assume $\sigma$, $\beta$, and $\gamma$ are independent in the prior with $\sigma \sim half-Cauchy(0, 2.5)$, $\beta \sim G(1, 1)$, and $\gamma \sim LN(0, 10)$.
-Note that the half-Caucy distribution is the Cauchy truncated to the positive real line.
+To complete the model we assume $\sigma$, $\beta$, and $\gamma$ are independent in the prior with $\sigma \sim \mathit{half-Cauchy}(0, 2.5)$, $\beta \sim G(1, 1)$, and $\gamma \sim LN(0, 10)$.
+Note that the half-Cauchy distribution is the Cauchy truncated to the positive real line.
 Table \ref{tab:modelfit} contains the results of fitting the model.

### 2015-12-21 postgres.R
Work on soft limits and stress tests

@@ -134,6 +134,54 @@ write.table(blockchain, file = "~/Desktop/blockchain.csv", sep = ",", col.names 
             qmethod = "double")
 
 
+# let's see if we can find the widely-used block size soft limits
+res <- dbSendQuery(con, "select height, size from blocks where branch=0")
+while(!dbHasCompleted(res)){
+  chunk <- dbFetch(res, n = 5)
+  print(nrow(chunk))
+}
+heightsize <- dbFetch(res)
+dbClearResult(res)
+d<-density(heightsize$size)
+plot(d)
+
+# 5 widely-used block size soft limits: 250kb, 750kb, something in the 300kb range, something in the 900kb range, and 1mb (also occasionally 500kb)
+# upon further research: 
+# 250kb limit introduced in July 2012
+# 350kb in version 0.8.6, released December 9, 2013 https://gist.github.com/gavinandresen/7670433#086-accept-into-block
+# 750kb in version 0.9.0, released March 19, 2014 https://bitcoin.org/en/release/v0.9.0
+# the value or values in the 900 range may be due almost exclusively to the Elegius mining pool. Still trying to pin down the exact value
+
+
+# let's also see if we can find the stress tests programmatically:
+# here's what we know
+# stress test data
+# 1. May 2015
+# blocks became full starting at block number 358596
+# remained full until block number 358609
+# citation: http://www.ofnumbers.com/2015/05/31/a-few-results-from-the-first-intentional-intentional-stress-test-on-a-communal-blockchain/
+# 2. June 22-23, 2015, 100 blocks
+# citation: http://www.coindesk.com/bitcoin-network-survives-stress-test/
+# 3. July 7-10, 2015
+# https://medium.com/blockcypher-blog/a-bitcoin-spam-attack-post-mortem-s-la-ying-alive-654e914edcf4#.y2rcrl10n
+# 4. September 2015
+# citation: https://bitcoinmagazine.com/articles/coinwallet-crowdsources-transactions-major-stress-test-bitcoin-giveaway-1442011334
+
+# install.packages("forecast")
+library(forecast)
+heightsize$mav <- ma(heightsize$size, 5000, centre=FALSE)
+plot(heightsize$height, heightsize$size, main="Bitcoin block sizes over time", 
+   xlab="Block Height ", ylab="Size in Bytes ", pch=".")
+lines(heightsize$height, heightsize$mav, type='l', col="blue")
[…]

### 2015-12-21 postgres.R
Fixed issues with moving average and rolling median plots

@@ -169,19 +169,23 @@ plot(d)
 
 # install.packages("forecast")
 library(forecast)
+heightsize <- heightsize[order(heightsize$height),]  ### make sure blocks are temporal order
 heightsize$mav <- ma(heightsize$size, 5000, centre=FALSE)
-plot(heightsize$height, heightsize$size, main="Bitcoin block sizes over time", 
-   xlab="Block Height ", ylab="Size in Bytes ", pch=".")
+## heightsize$mav2 <- filter(heightsize$size, rep(1/5000, 5000), "convolution", sides = 2)
+## moving average considering only past values
+## change to sides = 2 to consider future and past values -- equivalent to ma() above
+plot(heightsize$height, heightsize$size, main="Bitcoin block sizes over time",
+     xlab="Block Height ", ylab="Size in Bytes ", pch=".")
 lines(heightsize$height, heightsize$mav, type='l', col="blue")
-# produces a weird blue area toward the end of the series, am i doing something wrong?
 
 # install.packages("zoo")
 library(zoo)
 heightsize$rollmed <- rollmedian(heightsize$size, 999, fill=NA, align="right")
-plot(heightsize$height, heightsize$size, main="Bitcoin block sizes over time", 
-   xlab="Block Height ", ylab="Size in Bytes ", pch=".")
+plot(heightsize$height, heightsize$size, main="Bitcoin block sizes over time",
+     xlab="Block Height ", ylab="Size in Bytes ", pch=".")
 lines(heightsize$height, heightsize$rollmedian, type='l', col="blue")
 
 
 
 dbDisconnect(con)
+

### 2015-12-22 model/modelfits.R
Added new Gamma table, old one commented out; also deleted an extraneous line from the model fitting script & removed old data file

@@ -67,14 +67,13 @@ xtable(gsummary)
 
 ############## Full model fits
 
-nobs <- 1000
+nobs <- 500
 blockdatashort <- blockdata[(1000 - nobs + 1):1011,]
 lbs <- rep(0,nobs)
 for(i in 1:nobs){
   lbs[i] <- median(blockdatashort$min[i:(i+10)])
 }
 times <- blockdatashort$min[-c(1:11)]
-mbs <- blockdata$size[-c(1:11)]/1000000  ## convert to megabytes
 
 sum(lbs > times) ## check that each lower bound is below what it's bounding, should be 0
 

### 2015-12-23 model/modelfits.R
changed database query to only download blocks from branch 0

@@ -11,7 +11,7 @@ con <- dbConnect(RPostgres::Postgres(),dbname = 'toshi',
 
 dbListTables(con)
 
-res <- dbSendQuery(con, "select height, time, size, transactions_count from blocks order by height desc limit 5011")
+res <- dbSendQuery(con, "select branch, height, time, size, transactions_count from blocks where branch = 0 order by height desc limit 5011")
 while(!dbHasCompleted(res)){
   chunk <- dbFetch(res, n = 5)
   print(nrow(chunk))

### 2015-12-23 model/modelfits.R
Fixed accidental change of niter

@@ -67,7 +67,7 @@ xtable(gsummary)
 
 ############## Full model fits
 
-nobs <- 500
+nobs <- 1000
 blockdatashort <- blockdata[(1000 - nobs + 1):1011,]
 lbs <- rep(0,nobs)
 for(i in 1:nobs){

### 2015-12-30 model/modelfits-pre2k15.R
Results from fitting the model to pre-2015 data

@@ -1,0 +1,149 @@
+library(DBI)
+library(rstan)
+library(MCMCpack)
+library(xtable)
+
+con <- dbConnect(RPostgres::Postgres(),dbname = 'toshi', 
+                 host = 'toshi.cn6zzwcfsto5.us-east-1.rds.amazonaws.com',
+                 port = 5432,
+                 user = 'readonly',
+                 password = 'password')
+
+dbListTables(con)
+
+res <- dbSendQuery(con, "select branch, height, time, size, transactions_count from blocks where branch = 0 and height <= 385000 order by height desc limit 5011")
+while(!dbHasCompleted(res)){
+  chunk <- dbFetch(res, n = 5)
+  print(nrow(chunk))
+}
+
+blockdata <- dbFetch(res)
+dbClearResult(res)
+blockdata <- blockdata[order(blockdata$height),]
+blockdata$elapsed_time <- blockdata$time - c(NA,blockdata$time[-length(blockdata$time)])
+blockdata$min <- blockdata$time/60 ## convert to minutes
+
+save(blockdata, file = "blockdata-pre2015.RData")
+
+load("blockdata-pre2015.RData")
+
+
+############# Gamma fits
+
+## fit on all 5000
+mbs <- blockdata$size[-c(1:11)]/1000000  ## convert to megabytes
+gamma5000list <- list(mb = mbs, logalphapars = c(0,10), betapars = c(1, 1), nobs = length(mbs))
+g5000fit <- stan(file = 'gamma.stan', data = gamma5000list, chains = 4, iter = 4000)
+g5000out <- extract(g5000fit)
+g5000line <- c(summary(mcmc(matrix(g5000out$alpha, ncol=1)))[[1]][1], summary(mcmc(matrix(g5000out$alpha, ncol=1)))[[2]][c(1,5)])
+
+## fit on blocks of 1000
+g1000line <- NULL
+for(i in 1:5){
+  mbs <- blockdata$size[-c(1:11)]/1000000  ## convert to megabytes
+  mbs <- mbs[1:1000 + 1000*(i-1)]
+  gamma1000list <- list(mb = mbs, logalphapars = c(0,10), betapars = c(1, 1), nobs = length(mbs))
[…]

### 2015-12-30 charts/charts.R
charts R file

@@ -1,0 +1,101 @@
+# install.packages("devtools")
+# devtools::install_github("RcppCore/Rcpp")
+# devtools::install_github("rstats-db/DBI")
+# devtools::install_github("rstats-db/RPostgres")
+
+library(DBI)
+con <- dbConnect(RPostgres::Postgres(), dbname = "toshi", host = "toshi.cn6zzwcfsto5.us-east-1.rds.amazonaws.com", 
+	port = 5432, user = "readonly", password = "password")
+
+# LOAD ALL THE BLOCK HEADERS FROM THE MAIN CHAIN
+res <- dbSendQuery(con, "select * from blocks where branch=0")
+while (!dbHasCompleted(res)) {
+	chunk <- dbFetch(res, n = 5)
+	print(nrow(chunk))
+}
+blocks <- dbFetch(res)
+dbClearResult(res)
+dbDisconnect(con)
+
+
+# library(bit64)
+# unsigned <- function(x) {
+	# if (x >= 0) {
+		# return(as.integer64(x))
+	# } else {
+		# return(as.integer64(-x + 2147483648))
+	# }
+# }
+
+# CLEANING UP AND SAVING THE DATA
+blocks$created_at <- NULL
+blocks$id <- NULL
+blocks <- blocks[order(blocks$height), ]
+# foo <- lapply(blocks$fees, unsigned)
+save(blocks, file = "~/Documents/Research/Block Size/bitcointrans/data/blocks.Rda")
+blocks$work <- vapply(blocks$work, paste, collapse = ", ", character(1L))
+write.table(blocks, file = "~/Documents/Research/Block Size/bitcointrans/data/blocks.csv", sep = ",", col.names = NA, 
+	qmethod = "double")
+
+
+load("~/Documents/Research/Block Size/bitcointrans/data/blocks.Rda")
+
+
+# PLOT BLOCK SIZES VS BLOCK HEIGHT
+plot(blocks$height, blocks$size, main = "Bitcoin block sizes over time", xlab = "Block Height ", ylab = "Size in Bytes ", 
+	pch = ".")
+
+# LET'S VISUALIZE THE SOFT LIMITS IN USE
+d <- density(blocks$size, from = 0, to = 1e+06)
+plot(d)
+
+# LET'S SEE HOW DENSITY CHANGES OVER TIME
+d1 <- density(blocks$size[blocks$height <= 1e+05], from = 0, to = 1e+06)
[…]

### 2015-12-30 charts/charts.R
Added a couple lines for a mean-var scatter plot, and a higher order regression for mean-variance

@@ -99,3 +99,6 @@ bothreg <-lm(mean ~ sd + var,summary)
 summary(varreg)
 summary(sdreg)
 summary(bothreg)
+
+plot(summary$mean, summary$var)
+meanreg <- lm(var ~ mean + I(mean^2) + I(mean^3) - 1, data = summary)

### 2016-01-08 model/overflowmcmc.tex
Changed the overflow model to include a trend and began writing up the probable MCMC algorithm - not implemented yet

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{bm, bbm}
\usepackage[authoryear]{natbib}

\begin{document}
\section{Introduction}
This note describes the MCMC algorithm used to fit the overflow model. The bulk of this material may end up in a technical appendix or as a section of the paper. Section \ref{sec:model} briefly describes the model and section \ref{sec:mcmc} describes the MCMC strategy for computing the posterior distribution.

\section{The Overflow Model}\label{sec:model}
The overflow model assumes we observe several pieces of data, 1) a noisy measurement of the arrival time of block $k$, $t_k$, measured in minutes since the zero'th block, 2) the number of transactions in block $k$, $M_k - M_{k-1}$ so that $M_k$ is the number of transactions on the entire block chain up to and including block $k$, and 3) the amount of data the transactions require to be stored in the block, $D_k$, measured in MBs, all observed for $k=1,2,\dots,n_{b}$.

\subsection{Measurement error for block arrival times}
Let $m_k = \mathrm{median}(t_{k - 1:11})$ where $t_{k-1:11} = (t_{k-11}, t_{k-10},\dots,t_{k-1})'$. Then by design $t_k > m_k$ for all $k$. Let $\tau_k$ denote the actual arrival time of the $k$'th block. Then for $k=1,2,\dots,n_{b}$ $t_k$ has a truncated normal distribution:
\begin{align*}
\log t_k |t_{1:(k-1)}, \tau_{1:n_{b}},\sigma^2 \sim N(\log\tau_k, \sigma^2)1(t_k > m_k).
\end{align*}

\subsection{Block interarrival times}
By the design of the bitcoin protocol, blocks are a Poisson process with rate $1/10$. Interarrival times, defined as $\delta_k = \tau_k - \tau_{k-1}$ with $\tau_0 = 0$, are exponentially distributed with rate parameter $1/10$ so that $E[\delta_k] = 10$. In other words for $k=1,2\dots,n_{b}$
\begin{align*}
\delta_k \stackrel{iid}{\sim} Exp(1/10).
\end{align*}

\subsection{Transaction arrival process}
We will assume that transactions arrive according to a non-homogenous Poisson process. Let $N(\tau)$ denote the number of transaction in the entire chain up to time $\tau$, and let $\lambda(\tau)$ denote some intensity function. Then any collection of non-overlapping increments $\left\{N(\tau_i + \delta_i) - N(\tau_i)\right\}$ are independent with distribution
\begin{align*}
N(\tau_i + \delta_i) - N(\tau_i) \sim Poi(\Lambda(\tau_i + \delta_i) - \Lambda(\tau_i))
\end{align*}
where 
\begin{align*}
\Lambda(\tau) = \int_0^\tau \lambda(u)du.
\end{align*}
Suppose the intensity function is a gamma process with shape function $\phi\Psi(\tau)$ and rate parameter $\phi$. In other words
\begin{align*}
\Lambda(\tau_i + \delta_i) - \Lambda(\tau_i) \stackrel{ind}{\sim} G(\phi[\Psi(\tau_i + \delta_i) - \Psi(\tau_i)], \phi)
\end{align*}
with mean $\Psi(\tau_i + \delta_i) - \Psi(\tau_i)$. Then for block $k=1,2,\dots,n_{b}$ define $\lambda_k = \Lambda(\tau_k) - \Lambda(\tau_{k-1})$, $\eta_k = N(\tau_k) - N(\tau_{k-1})$, and $\psi_k = \Psi(\tau_k) - \Psi(\tau_{k-1})$. Then assume
\begin{align*}
\eta_k | \lambda_{1:n_{b}} &\stackrel{ind}{\sim} Poi(\lambda_k)\\
\lambda_k|\delta_{1:n_{b}} &\stackrel{ind}{\sim} G(\phi\psi_k, \phi).
\end{align*}

In order for the gamma process to be well defined we require $\phi>0$ and $\Psi(\tau)$ to be nondecreasing and right continuous with $\Psi(0)=0$. An easy way to achieve this is to define $\Psi(\tau) = \int_0^\tau\psi(t)dt$ where $\psi(t)$ is continuous and nonnegative. Then $\psi(t)$ is the intensity of the Gamma process, but is also intimately related to the intensity of the Poission process. The mean of $N(\tau + \delta) - N(\tau)$ is 
\begin{align*}
E[N(\tau + \delta) - N(\tau)] &= E\{E[N(\tau + \delta) - N(\tau)|\Lambda(\tau + \delta) - \Lambda(\tau)]\} = E[\Lambda(\tau + \delta) - \Lambda(\tau)]\\
 &= \Psi(\tau + \delta) - \Psi(\delta) = \int_{\tau}^{\tau + \delta}\psi(t)dt.
\end{align*}
In this way, $\psi(t)$ can be seen as the expected intensity function of the $N(\tau)$ process. The Gamma-Poisson structure also allows for a more flexible model on $N(\tau)$ by allowing for overdispersion. In a Poisson model the mean and variance are the same, but in this case they are proportional:
\begin{align*}
V[N(\tau + \delta) - N(\tau)] &= E\{V[N(\tau + \delta) - N(\tau)|\Lambda(\tau + \delta) - \Lambda(\tau)]\} + V\{E[N(\tau + \delta) - N(\tau)|\Lambda(\tau + \delta) - \Lambda(\tau)]\}\\
&= E[\Lambda(\tau + \delta) - \Lambda(\tau)] + Var[\Lambda(\tau + \delta) - \Lambda(\tau)]\\
&= [\Psi(\tau + \delta) - \Psi(\tau)](1 + 1/\phi) = \frac{\phi + 1}{\phi}\int_{\tau}^{\tau + \delta}\psi(t)dt.
\end{align*}
This allows for overdispersion but not underdispersion since $(\phi + 1)/\phi\in[1,\infty)$ for $\phi\ge 0$.

To complete this portion of the model we need to specify $\psi(t)$. A flexible structure that captures many modeling choices is $\psi(t) = \exp(\bm{z}(t)'\bm{\beta})$ where $\bm{z}(t)' = (z_1(t),\dots,z_p(t))$ is a vector of time-varying covariates continuously differentiable in time and $\bm{\beta}'=(\beta_1,\dots,\beta_p)$ is a vector of regression coefficients, for example $\psi(t) = \exp(\beta_1 + t\beta_2)$ in order to capture a simple trend. In that case we have
\begin{align*}
\Psi(\tau+\delta) - \Psi(\tau) &= \int_{\tau}^{\tau + \delta} e^{\bm{z}(t)'\bm{\beta}}dt = \int_\tau^{\tau + \delta} e^{\beta_1 + t\beta_2}dt = \frac{e^{\beta_1 + \tau\beta_2}}{\beta_2}\left(e^{\delta\beta_2} - 1\right)
\intertext{and thus}
\psi_k &= \frac{e^{\beta_1 + \tau_{k-1}\beta_2}}{\beta_2}\left(e^{\delta_k\beta_2} - 1\right).
\end{align*}
\subsection{Transaction data}
Let $i=1,2,\dots,n_{trans}$ index transactions and $d_i$ denote the amount of data for transaction $i$ measured in MBs. Then we assume that for $i=1,2,\dots,n_{trans}$
\begin{align*}
d_i \stackrel{iid}{\sim} G(\alpha_d,\beta_d)
\end{align*}
with mean $\alpha_d/\beta_d$. Along with the Poisson process for $N(\tau)$, this implies that $D(\tau)$, the amount of transaction data that has arrived by time $\tau$, is a compound non-homogenous Poisson process conditional on $\Lambda(\tau)$.

\subsection{Overflow and observations of $M_k$ and $D_k$}
Each block has an upper limit for the amount of transaction data it can hold defined by the block miner. Denote this upper limit by $\bar{D}_k$ --- because different miners use different limits and the same miner uses different limits at different times, each block has its own upper limit. Then the number of transactions in blocks $1,2,\dots,k$ combined, $M_k$, could potentially be less than the number of transactions that have arrive by the time the $k$'th block arrives, $N_k$. Specifically:
\begin{align*}
M_k &= \max\left\{m\in\left\{M_{k-1}, M_{k-1} + 1, \dots, N_k\right\}: \sum_{i = M_{k-1} + 1}^m d_i \le \bar{D}_k\right\}
\end{align*}
so that
\begin{align*}
D_k &= \sum_{i = M_{k-1} + 1}^{M_k} d_i
\end{align*}
with the convention that empty summations are defined as zero, i.e. $\sum_{i=5}^4d_i \equiv 0$. Finally, the upper limits are modeled as a categorical distribution over a set of prespecified possible limits, $\Delta_1,\dots,\Delta_{n_{limits}}$ with probability $p_{ik} = P(\bar{D}_k = \Delta_i)$, $\sum_ip_{ik} = 1$ and $p_{ik} \ge 0$. In other words
\begin{align*}
\bar{D}_k\stackrel{ind}{\sim} Cat(\Delta_{1:n_{limits}}, p_{1:n_{limits},k}).
\end{align*}

\section{Markov Chain Monte Carlo}\label{sec:mcmc}

In order to simulate from the posterior we construct a Gibbs sampler using data augmentation to expand the state space and Metropolis steps where appropriate. We use data augmentation in two ways. First, in order to more effectively deal with the truncated distribution in the measurement error model for block arrival times we introduce $\tilde{\bm{t}}_k=(\tilde{t}_{1k},\dots,\tilde{t}_{\tilde{n}_kk}, t_k)$ as the original $t_k$ along with $\tilde{n}_k$ draws from the untruncated lognormal distribution for $t_k$ that occurred before $t_k$ was drawn from the restricted support. Second, the models for $\eta_{1:n_{b}}$ and $d_{1:n_{trans}}$ induce a model for $(M_{1:n_{b}},D_{1:n_{trans}})$ since the latter are a detrministic function of the former, but this model is not easy work with. Instead, we explicitly use $\eta_{1:n_{b}}$ and $d_{1:n_{trans}}$ in the Gibbs sampler as additional data augmentation.

\subsection{Data augmentation for truncated distributions}
For block $k=1,2,n_{b}$ the measurement error model on block arrival times is
\begin{align*}
\log t_k \sim N(\log\tau_k,\sigma^2)1(t_k > m_k)
\end{align*}
where $m_k = \mathrm{median}(t_{k - 1:11})$ with density 
\begin{align*}
p(\log\bm{t_{1:n_{b}}}|\bm{\tau}_{1:n_{b}},\sigma^2) \propto \sigma^{-n_{b}}\frac{\exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}(\log t_K - \log \tau_k)^2\right]}{\prod_{k=1}^{n_{b}}\left[1 - \Phi\left(\frac{\log m_k - \log\tau_k}{\sigma}\right)\right]}
\end{align*}
where $\Phi(.)$ is the standard normal cdf. Drawing from the full conditional distributions of $\sigma^2$ and $\tau_k$ can be challenging since both parameters enter into the normalizing constant of $p(\log\bm{t_{1:n_{b}}}|\bm{\tau}_{1:n_{b}},\sigma^2)$. Consider an underlying process which gives rise to the truncated distribution --- instead of simply drawing $t_k$ from the truncated lognormal distribution, this process draws from the untruncated lognormal distribution until it obtains a draw from the restricted region of the parameter space, i.e. until it successfully obtains a $t_k$ such that $t_k > m_k$. Let $\tilde{n}_k$ be the number of failures before the observed success $t_k$. Then $\tilde{n}_k$ has a negative binomial distribution, specifically $\tilde{n}_k \sim NB(1, p_k)$ with density
\begin{align*}
p(\tilde{n}_k|\tau_k,\sigma^2) = p_k^{\tilde{n}_k}(1 - p_k)
\end{align*}
where $p_k$ is the probability of failure, i.e. $p_k = \Phi([\log m_k - \log \tau_k]/\sigma)$. Conditional on $\tilde{n}_k$ the density of the $\tilde{n}_k$ failures, $\tilde{t}_{1k},\dots,\tilde{t}_{\tilde{n}_kk}$, is again truncated lognormal, but this time truncated to the opposite end of the parameter space. That is for $i = 1,\dots,\tilde{n}_k$
\begin{align*}
\log\tilde{t}_{ik}\stackrel{iid}{\sim}N(\log\tau_k,\sigma^2)1(t_k < m_k).
\end{align*}

So the data augmentation step of the Gibbs sampler consists of the following substeps: for $k=1,2,\dots,n_{b}$,
\begin{enumerate}
\item Draw $\tilde{n}_k \sim NB(1,p_k)$ where $p_k = \Phi([\log m_k - \log \tau_k]/\sigma)$.
\item If $\tilde{n}_k > 0$, for $i=1,\dots,\tilde{n}_k$ draw $\tilde{t}_{ik} \sim N(\log\tau_k,\sigma^2)1(t_k < m_k)$ and form $\tilde{\bm{t}}_k = (\tilde{t}_{1k},\dots,\tilde{t}_{\tilde{n}_kk}, t_k)'$.
\end{enumerate}
For simplicity in the other Gibbs steps define $\tilde{t}_{\tilde{n}_k + 1,k}\equiv t_k$.

It is tempting to imbue $\tilde{n}_k$ and $\tilde{t_{ik}}$ with some interpretation relative to the model, e.g. as is [CITE GELMAN PAPER]. For example, $\tilde{n}_k$ may represent the number of blocks that were rejected by the Bitcoin protocol for having a timestamp earlier than the median of the previous elevent blocks. While tempting this interpretation is not strictly speak correct. When a block is rejected by the protocol a new block does not appear instantaneously since it takes time for another miner to discover the block. So the arrival times and thus measurement error distributions of the rejected block and the accepted block are not the same. While the interpretation is not correct, it is good to be inspired by computational tricks to create better model [CITE SAME GELMAN PAPER] and perhaps something similar to this negative binomial structure can be used to add rejected blocks as a component of the model.

\subsection{measurement error variance}
The full conditional distribution of the measurement error variance, $\sigma^2$, depends only on $\tilde{\bm{t}}_{1:n_{b}}$ and $\tau_{1:n_{b}}$. It is
\begin{align*}
p(\sigma^2|...) \propto \sigma^{-(n_{b} + \sum_{k=1}^{n_{b}}\tilde{n}_k)}\exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\log \tilde{t}_{ik} - \log\tau_k)^2\right]p(\sigma^2)
\end{align*}
where $p(\sigma^2)$ is the density of the prior distribution on $\sigma^2$. We will suppose that $\sigma$ has a half-$t$ prior, $\sigma \sim t_\nu(S)1(\sigma>0)$ with degrees of freedom and scale parameter $S$. This prior is equivalent to the marginal prior on $\sigma$ when $\sigma^2|\omega \sim IG(\nu/2, \omega)$ and $\omega \sim G(1/2, 1/(\nu S^2))$. Then letting $a_\sigma = (n_{b} + \sum_{k=1}^{n_b}\tilde{n}_k)/2$ and $b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\log \tilde{t}_{ik} - \log\tau_k)^2$ the full conditional distribution of $(\sigma^2,\omega)$ is
\begin{align*}
p(\sigma^2,\omega|...) \propto (\sigma^2)^{-(a_\sigma+\nu)/2-1}\left[-\frac{b_\sigma}{\sigma^2}\right]\exp\left[-\frac{\omega}{\sigma^2}\right]\omega^{(\nu + 1)/2 - 1}\exp\left[-\frac{\omega}{\nu S^2}\right].
\end{align*}
So we draw $\sigma^2$ and $\omega$ in two separate Gibbs steps:
\begin{enumerate}
\item Draw $\omega \sim G\left(\frac{\nu + 1}{2}, \frac{1}{\sigma^2} + \frac{1}{\nu S^2}\right)$ and form
\begin{align*}
a_\sigma = n_{b} + \sum_{k=1}^{n_b}\tilde{n}_k &&\mbox{  and  }&& b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\log \tilde{t}_{ik} - \log\tau_k)^2/2;
\end{align*}
\item Draw $\sigma^2 \sim  IG([a_\sigma + \nu]/2, b_\sigma + \omega)$.
\end{enumerate}

\subsection{block arrival times}
The full conditional distribution of the block arrival times is more complicated:
\begin{align*}
p(\bm{\tau}_{1:n_{b}}|...) \propto \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\log\tilde{t}_{ik} - \log\tau_k\right)^2 - \frac{1}{10}\tau_{n_{b}}\right]\frac{\phi^{\phi[\Psi(\tau_k) - \Psi(\tau_{k-1})]}}{\Gamma\left(\phi[\Psi(\tau_k) - \Psi(\tau_{k-1})]\right)}\lambda_k^{\phi[\Psi(\tau_k)-\Psi(\tau_{k-1})] - 1}
\end{align*}
where $\tau_0\equiv0$ and $\Gamma(x) = \int_0^\infty t^{x-1}e{-t}dt$ is the gamma function. Even $\Psi(\tau)$ is some nice, known form, this density is nonstandard because the $\tau_k$'s enter the Gamma function. As a result, we have several options for this step:
\begin{enumerate}
\item Draw each $\tau_k$ condtional on all the others using random walk Metropolis steps for $k=1,\dots,n_{b}$.
\item Approximate the density of $\tau_k$ sufficiently well and use independent Metropolis steps for $k=1,\dots,n_{b}$.
\item Approximate the density of $\tau_{1:n_{b}}$ sufficiently well and use an independent Metropolis step.
\end{enumerate}
Options 2 and 3 could both be replaced with a rejection sampling step under some circumstances, but a Metropolis step is likely cheaper computationally [CITE THE LIU PAPER].

Ideas:
\begin{enumerate}
\item ARMS -- ARS except with a Metropolis correction for non-log concave densities (look it up).
\item RW Metropolis and tune the proposal variances.
\end{enumerate}

\subsection{transaction process dispersion}
The transaction process dispersion parameter is $\phi$ --- $V[N(\tau)] = E[N(\tau)](1 + 1/\phi)$. It has the full conditional density
\begin{align*}
p(\phi|...) \propto \prod_{k=1}^{n_{b}}\frac{\phi^{\phi\psi_k}\lambda_k^{\phi\psi_k - 1}e^{\lambda_k\phi}}{\Gamma(\phi\psi_k)}p(\phi).
\end{align*}
This is another complicated density, but ARS should work decently to sample from it. If not, various Metropolis options are available. Another option: transform so that $\phi$ is absorbed into the shape function. Then $\tilde{\phi}$ has the full conditional
\begin{align*}
p(\phi|...) \propto \phi^{\sum_{k=1}^{n_{b}}\psi_k}e^{\phi\sum_{k=1}^{n_{b}}\lambda_k}p(\phi).
\end{align*}
With $\phi \sim G(a_\phi, b_\phi)$ in the prior, this becomes
\begin{enumerate}
\item Draw $\phi \sim G\left(a_\phi + \sum_{k=1}^{n_{b}}\psi_k, b_{\phi} + \sum_{k=1}^{n_{b}}\lambda_k \right)$.
\end{enumerate}

\subsection{regression parameters}

\subsection{transaction arrival times and data amounts}

\subsection{transaction data amount parameters}
\bibliographystyle{apalike} 
\bibliography{../paper/btc}

\end{document}

### 2016-01-12 model/overflowmcmc.tex
Mostly finished writing up model + sketch of algorithm

@@
 \section{The Overflow Model}\label{sec:model}
-The overflow model assumes we observe several pieces of data, 1) a noisy measurement of the arrival time of block $k$, $t_k$, measured in minutes since the zero'th block, 2) the number of transactions in block $k$, $M_k - M_{k-1}$ so that $M_k$ is the number of transactions on the entire block chain up to and including block $k$, and 3) the amount of data the transactions require to be stored in the block, $D_k$, measured in MBs, all observed for $k=1,2,\dots,n_{b}$.
+The overflow model assumes we observe several pieces of data: 1) a noisy measurement of the arrival time of block $k$ measured in minutes since the zero'th block and denoted by $t_k$, 2) the number of transactions in block $k$ denoted by $M_k - M_{k-1}$ so that $M_k$ is the number of transactions on the entire block chain up to and including block $k$, and 3) the amount of transaction data stored in block $k$ measured in MBs and denoted by $D_k$, all observed for $k=1,2,\dots,n_{b}$.
 
@@
 Let $\tau_k$ denote the actual arrival time of the $k$'th block.
-Then for $k=1,2,\dots,n_{b}$ $t_k$ has a truncated normal distribution:
+Then for $k=1,2,\dots,n_{b}$, $t_k$ has a truncated normal distribution:
 \begin{align*}
-\log t_k |t_{1:(k-1)}, \tau_{1:n_{b}},\sigma^2 \sim N(\log\tau_k, \sigma^2)1(t_k > m_k).
+t_k |t_{1:(k-1)}, \tau_{1:n_{b}},\sigma^2 \sim N(\tau_k, \sigma^2)1(t_k > m_k).
 \end{align*}
@@
 We will assume that transactions arrive according to a non-homogenous Poisson process.
-Let $N(\tau)$ denote the number of transaction in the entire chain up to time $\tau$, and let $\lambda(\tau)$ denote some intensity function.
-Then any collection of non-overlapping increments $\left\{N(\tau_i + \delta_i) - N(\tau_i)\right\}$ are independent with distribution
+Let $N(\tau)$ denote the number of transactions that have arrive by time $\tau$, and let $\lambda(\tau)$ denote some intensity function such that $\lambda(\tau) > 0$ for all times $\tau$.
+Then any collection of non-overlapping increments $\left\{N(\tau_i + \delta_i) - N(\tau_i)\right\}$ where $\delta_i>0$ for all $i$ are independent with distribution
 \begin{align*}
@@
 \end{align*}
-Suppose the intensity function is a gamma process with shape function $\phi\Psi(\tau)$ and rate parameter $\phi$. In other words
+Suppose the intensity function is a gamma process with shape function $\Psi(\tau)$ and rate parameter $\phi$. In other words
 \begin{align*}
-\Lambda(\tau_i + \delta_i) - \Lambda(\tau_i) \stackrel{ind}{\sim} G(\phi[\Psi(\tau_i + \delta_i) - \Psi(\tau_i)], \phi)
+\Lambda(\tau_i + \delta_i) - \Lambda(\tau_i) \stackrel{ind}{\sim} G(\Psi(\tau_i + \delta_i) - \Psi(\tau_i), \phi)
 \end{align*}
-with mean $\Psi(\tau_i + \delta_i) - \Psi(\tau_i)$.
+with mean $[\Psi(\tau_i + \delta_i) - \Psi(\tau_i)]/\phi$.
 Then for block $k=1,2,\dots,n_{b}$ define $\lambda_k = \Lambda(\tau_k) - \Lambda(\tau_{k-1})$, $\eta_k = N(\tau_k) - N(\tau_{k-1})$, and $\psi_k = \Psi(\tau_k) - \Psi(\tau_{k-1})$.
@@
 \eta_k | \lambda_{1:n_{b}} &\stackrel{ind}{\sim} Poi(\lambda_k)\\
-\lambda_k|\delta_{1:n_{b}} &\stackrel{ind}{\sim} G(\phi\psi_k, \phi).
+\lambda_k|\delta_{1:n_{b}} &\stackrel{ind}{\sim} G(\psi_k, \phi).
 \end{align*}
@@
 An easy way to achieve this is to define $\Psi(\tau) = \int_0^\tau\psi(t)dt$ where $\psi(t)$ is continuous and nonnegative.
-Then $\psi(t)$ is the intensity of the Gamma process, but is also intimately related to the intensity of the Poission process.
+While $\psi(t)$ can be thought of as the intensity of the Gamma process it is also intimately related to the intensity of the Poission process.
 The mean of $N(\tau + \delta) - N(\tau)$ is
@@
 E[N(\tau + \delta) - N(\tau)] &= E\{E[N(\tau + \delta) - N(\tau)|\Lambda(\tau + \delta) - \Lambda(\tau)]\} = E[\Lambda(\tau + \delta) - \Lambda(\tau)]\\
- &= \Psi(\tau + \delta) - \Psi(\delta) = \int_{\tau}^{\tau + \delta}\psi(t)dt.
+ &= [\Psi(\tau + \delta) - \Psi(\delta)]/\phi = \int_{\tau}^{\tau + \delta}\psi(t)dt/\phi.
 \end{align*}
-In this way, $\psi(t)$ can be seen as the expected intensity function of the $N(\tau)$ process.
+In this way, $\psi(t)/\phi$ can be seen as the expected intensity function of the $N(\tau)$ process.
 The Gamma-Poisson structure also allows for a more flexible model on $N(\tau)$ by allowing for overdispersion.
@@
 V[N(\tau + \delta) - N(\tau)] &= E\{V[N(\tau + \delta) - N(\tau)|\Lambda(\tau + \delta) - \Lambda(\tau)]\} + V\{E[N(\tau + \delta) - N(\tau)|\Lambda(\tau + \delta) - \Lambda(\tau)]\}\\
-&= E[\Lambda(\tau + \delta) - \Lambda(\tau)] + Var[\Lambda(\tau + \delta) - \Lambda(\tau)]\\
-&= [\Psi(\tau + \delta) - \Psi(\tau)](1 + 1/\phi) = \frac{\phi + 1}{\phi}\int_{\tau}^{\tau + \delta}\psi(t)dt.
+&= E[\Lambda(\tau + \delta) - \Lambda(\tau)] + V[\Lambda(\tau + \delta) - \Lambda(\tau)]\\
+&= [\Psi(\tau + \delta) - \Psi(\tau)](1/\phi + 1/\phi^2) = \frac{\phi + 1}{\phi}E[N(\tau + \delta) - N(\tau)].
 \end{align*}
@@
 \subsection{Transaction data}
-Let $i=1,2,\dots,n_{trans}$ index transactions and $d_i$ denote the amount of data for transaction $i$ measured in MBs. Then we assume that for $i=1,2,\dots,n_{trans}$
+Let $i=1,2,\dots,n_d \equiv N_{n_b}$ index transactions and $d_i$ denote the amount of data for transaction $i$ measured in MBs. Then we assume that for $i=1,2,\dots,n_d$
 \begin{align*}
@@
 with mean $\alpha_d/\beta_d$.
-Along with the Poisson process for $N(\tau)$, this implies that $D(\tau)$, the amount of transaction data that has arrived by time $\tau$, is a compound non-homogenous Poisson process conditional on $\Lambda(\tau)$.
+Along with the Poisson process for $N(\tau)$ this implies that the amount of transaction data that has arrived by time $\tau$, $D(\tau)$, is a compound non-homogenous Poisson process conditional on $\Lambda(\tau)$.
 
@@
 Denote this upper limit by $\bar{D}_k$ --- because different miners use different limits and the same miner uses different limits at different times, each block has its own upper limit.
-Then the number of transactions in blocks $1,2,\dots,k$ combined, $M_k$, could potentially be less than the number of transactions that have arrive by the time the $k$'th block arrives, $N_k$.
+Then the number of transactions in blocks $1,2,\dots,k$ combined, $M_k$, could potentially be less than the number of transactions that have arrived by the time the $k$'th block arrives, $N_k$.
 Specifically:
@@
 \end{align*}
-with the convention that empty summations are defined as zero, i.e.
-$\sum_{i=5}^4d_i \equiv 0$.
-Finally, the upper limits are modeled as a categorical distribution over a set of prespecified possible limits, $\Delta_1,\dots,\Delta_{n_{limits}}$ with probability $p_{ik} = P(\bar{D}_k = \Delta_i)$, $\sum_ip_{ik} = 1$ and $p_{ik} \ge 0$.
+with the convention that empty summations are defined as zero, i.e. $\sum_{i=5}^4d_i \equiv 0$.
+
+\section{Priors}
+The model has the following unknown parameters: $\sigma^2$, $\phi$, $\beta$, $\alpha_d$, $\beta_d$, and $\bar{D}_k$ for $k=1,2,\dots,n_b$.
+We will suppose that these parameters are mutually independent in the prior, i.e.
+\begin{align*}
+p(\sigma^2,\phi,\beta,\alpha_d,\beta_d,\bar{D}_1,\dots,\bar{D}_{n_{b}}) = p(\sigma^2)p(\phi)p(\beta)p(\alpha_d)p(\beta_d)p(\bar{D}_1)\hdots p(\bar{D}_{n_{b}}).
+\end{align*}
+
+\subsection{Prior on $\sigma^2$}
+The expected size of the measurement error on $t_k$ is controlled by $\sigma^2$ --- the larger $\sigma^2$, the farther away we expect $t_k$ to be from $\tau_k$.
+We will use a half-$t$ prior on $\sigma$ which can be written as the following prior on $\sigma^2$:
+\begin{align*}
+\sigma^2|\omega &\sim IG(v/2, \omega); && \omega \sim G(1/2, s^2/2).
+\end{align*}
+This yields a $t_{v}(0,s^2)1(\sigma>0)$ prior on $\sigma$ where $s^2$ is a scale parameter and $v$ is a degrees of freedom parameter.
+We use $v=1$ so that the prior on $\sigma$ has a half-Cauchy distribution with median $s$.
+To gain some intuition in order to set an appropriate value for $s$, suppose that the measurement error distribution was not truncated.
+Then we would expect $68\%$ of observed block arrival times to be within $\sigma$ of the actual arrival time.
+A reasonable a priori guess for $\sigma$ is then one minute, so we set $s=1$ so that the median value of $\sigma$ in its prior is one.
+This prior is only weakly informative --- the right tail is gently sloped so that the data will quickly overwhelm the prior as observations pile up [CITE GELMAN VARIANCE PAPER].
+
+\subsection{Prior on $\phi$}
+The conditionally conjugate prior for $\phi$ is a gamma distribution, so we will suppose that $\phi \sim G(a_{\phi}, b_{\phi})$.
+To choose $a_\phi$ and $b_\phi$ recall that $\phi$ is a dispersion parameter:
+\begin{align*}
+V[N(\tau)|\phi,\Psi(\tau)] = \frac{\phi + 1}{\phi}E[N(\tau)|\phi,\Psi(\tau)] = \frac{\phi + 1}{\phi}\frac{\Psi(\tau)}{\phi}.
+\end{align*}
+Setting $\phi = 1$ means that the variance of $N(\tau)$ is twice its mean.
+Absent strong prior information about $\phi$, we will center our prior on $2$ with a high level of uncertainty.
+The mean of the gamma distribution is $a_\phi/b_\phi$ with variance $a_\phi/b_\phi^2$, so we will set $b_\phi = 1/100$ to reflect a high degree of uncertainty, and $a_\phi = 1/50$ so that $a_\phi/b_\phi = 2$.
+
+\subsection{Prior on $\bm{\beta}$}
+As we will see in the next section, there is no convenient conditionally conjugate form for $\bm{\beta}$, so there is no computational trade-off associated with choosing a prior which more accurately reflects prior knowledge.
+We will consider $\bm{\beta}$ much like a regression parameter and set
+\begin{align*}
+\bm{\beta} \sim N(\bm{b}_\beta, \bm{S}_\beta)
+\end{align*}
+where $\bm{b}_\beta$ is a mean vector and $\bm{S}_\beta$ is a covariance matrix.
+We will use a weakly informative prior with $\bm{b}_\beta = \bm{0}$ and $\bm{S}_\beta = s^2_\beta\bm{I}$ where $s_\beta = 100$.
+[MAYBE SOMETHING WITH FATTER TAILS HERE?
+MAYBE NOTE THAT NONINFORMATIVE PRIOR IS NOT A GOOD IDEA SINCE WE WANT TO GUARANTEE POSTERIOR PROPRIETY --- SAME FOR $s_\beta$ TOO LARGE (VIVEK'S PAPER)]
+
+\subsection{Priors on $\alpha_d$ and $\beta_d$}
+
+\subsection{Priors on $\bar{D}_k$'s}
+
+Finally, the upper limits are modeled as a categorical distribution over a set of prespecified possible limits, $\Delta_1,\dots,\Delta_{n_{lim}}$ with probability $p_{ik} = P(\bar{D}_k = \Delta_i)$, $\sum_ip_{ik} = 1$ and $p_{ik} \ge 0$.
 In other words
 \begin{align*}
-\bar{D}_k\stackrel{ind}{\sim} Cat(\Delta_{1:n_{limits}}, p_{1:n_{limits},k}).
+\bar{D}_k\stackrel{ind}{\sim} Cat(\Delta_{1:n_{lim}}, p_{1:n_{lim},k}).
 \end{align*}
@@
 We use data augmentation in two ways.
-First, in order to more effectively deal with the truncated distribution in the measurement error model for block arrival times we introduce $\tilde{\bm{t}}_k=(\tilde{t}_{1k},\dots,\tilde{t}_{\tilde{n}_kk}, t_k)$ as the original $t_k$ along with $\tilde{n}_k$ draws from the untruncated lognormal distribution for $t_k$ that occurred before $t_k$ was drawn from the restricted support.
-Second, the models for $\eta_{1:n_{b}}$ and $d_{1:n_{trans}}$ induce a model for $(M_{1:n_{b}},D_{1:n_{trans}})$ since the latter are a detrministic function of the former, but this model is not easy work with.
-Instead, we explicitly use $\eta_{1:n_{b}}$ and $d_{1:n_{trans}}$ in the Gibbs sampler as additional data augmentation.
+First, in order to more effectively deal with the truncated distribution in the measurement error model for block arrival times we introduce $\tilde{\bm{t}}_k=(\tilde{t}_{1k},\dots,\tilde{t}_{\tilde{n}_kk}, t_k)$ as the original $t_k$ along with $\tilde{n}_k$ draws from the untruncated normal distribution for $t_k$ that occurred before $t_k$ was drawn from the restricted support.
+Second, the models for $\eta_{1:n_{b}}$ and $d_{1:n_d}$ induce a model for $(M_{1:n_{b}},D_{1:n_d})$ since the latter are a detrministic function of the former, but this model is not easy work with.
+Instead, we explicitly use $\eta_{1:n_{b}}$ and $d_{1:n_d}$ in the Gibbs sampler as additional data augmentation.
 
-\subsection{Data augmentation for truncated distributions}
+\subsection{Data augmentation for truncated distributions --- $\tilde{n}_k$ and $\tilde{t}_k$}
 For block $k=1,2,n_{b}$ the measurement error model on block arrival times is
 \begin{align*}
-\log t_k \sim N(\log\tau_k,\sigma^2)1(t_k > m_k)
+t_k|t_{1:(k-1)} \sim N(\tau_k,\sigma^2)1(t_k > m_k)
 \end{align*}
@@
 \begin{align*}
-p(\log\bm{t_{1:n_{b}}}|\bm{\tau}_{1:n_{b}},\sigma^2) \propto \sigma^{-n_{b}}\frac{\exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}(\log t_K - \log \tau_k)^2\right]}{\prod_{k=1}^{n_{b}}\left[1 - \Phi\left(\frac{\log m_k - \log\tau_k}{\sigma}\right)\right]}
+p(\bm{t_{1:n_{b}}}|\bm{\tau}_{1:n_{b}},\sigma^2) \propto \sigma^{-n_{b}}\frac{\exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}(t_k - \tau_k)^2\right]}{\prod_{k=1}^{n_{b}}\left[1 - \Phi\left(\frac{m_k - \tau_k}{\sigma}\right)\right]}
 \end{align*}
 where $\Phi(.)$ is the standard normal cdf.
-Drawing from the full conditional distributions of $\sigma^2$ and $\tau_k$ can be challenging since both parameters enter into the normalizing constant of $p(\log\bm{t_{1:n_{b}}}|\bm{\tau}_{1:n_{b}},\sigma^2)$.
-Consider an underlying process which gives rise to the truncated distribution --- instead of simply drawing $t_k$ from the truncated lognormal distribution, this process draws from the untruncated lognormal distribution until it obtains a draw from the restricted region of the parameter space, i.e.
-until it successfully obtains a $t_k$ such that $t_k > m_k$.
+Drawing from the full conditional distributions of $\sigma^2$ and $\tau_k$ can be challenging since both parameters enter into the normalizing constant of $p(\bm{t_{1:n_{b}}}|\bm{\tau}_{1:n_{b}},\sigma^2)$.
+Consider an underlying process which gives rise to the truncated distribution --- instead of simply drawing $t_k$ from the truncated normal distribution, this process draws from the untruncated normal distribution until it obtains a draw from the restricted region of the parameter space, i.e.
+until it successfully obtains a $t_k > m_k$.
 Let $\tilde{n}_k$ be the number of failures before the observed success $t_k$.
-Then $\tilde{n}_k$ has a negative binomial distribution, specifically $\tilde{n}_k \sim NB(1, p_k)$ with density
+Then $\tilde{n}_k$ has a negative binomial distribution, specifically $\tilde{n}_k \sim NB(1, \rho_k)$ with density
 \begin{align*}
-p(\tilde{n}_k|\tau_k,\sigma^2) = p_k^{\tilde{n}_k}(1 - p_k)
+p(\tilde{n}_k|\tau_k,\sigma^2) = \rho_k^{\tilde{n}_k}(1 - \rho_k)
 \end{align*}
-where $p_k$ is the probability of failure, i.e.
-$p_k = \Phi([\log m_k - \log \tau_k]/\sigma)$.
-Conditional on $\tilde{n}_k$ the density of the $\tilde{n}_k$ failures, $\tilde{t}_{1k},\dots,\tilde{t}_{\tilde{n}_kk}$, is again truncated lognormal, but this time truncated to the opposite end of the parameter space.
+where $\rho_k$ is the probability of failure, i.e.
+$\rho_k = \Phi([m_k - \tau_k]/\sigma)$.
+Conditional on $\tilde{n}_k$ the density of the $\tilde{n}_k$ failures, $\tilde{t}_{1k},\dots,\tilde{t}_{\tilde{n}_kk}$, is again truncated normal, but this time truncated to the opposite end of the parameter space.
 That is for $i = 1,\dots,\tilde{n}_k$
 \begin{align*}
-\log\tilde{t}_{ik}\stackrel{iid}{\sim}N(\log\tau_k,\sigma^2)1(t_k < m_k).
+\tilde{t}_{ik}\stackrel{iid}{\sim}N(\tau_k,\sigma^2)1(t_k < m_k).
 \end{align*}
-
 So the data augmentation step of the Gibbs sampler consists of the following substeps: for $k=1,2,\dots,n_{b}$,
 \begin{enumerate}
-\item Draw $\tilde{n}_k \sim NB(1,p_k)$ where $p_k = \Phi([\log m_k - \log \tau_k]/\sigma)$.
-\item If $\tilde{n}_k > 0$, for $i=1,\dots,\tilde{n}_k$ draw $\tilde{t}_{ik} \sim N(\log\tau_k,\sigma^2)1(t_k < m_k)$ and form $\tilde{\bm{t}}_k = (\tilde{t}_{1k},\dots,\tilde{t}_{\tilde{n}_kk}, t_k)'$.
+\item Draw $\tilde{n}_k \sim NB(1,\rho_k)$ where $\rho_k = \Phi([ m_k -  \tau_k]/\sigma)$.
+\item If $\tilde{n}_k > 0$, for $i=1,\dots,\tilde{n}_k$ draw $\tilde{t}_{ik} \sim N(\tau_k,\sigma^2)1(t_k < m_k)$ and form $\tilde{\bm{t}}_k = (\tilde{t}_{1k},\dots,\tilde{t}_{\tilde{n}_kk}, t_k)'$.
 \end{enumerate}
@@
 It is tempting to imbue $\tilde{n}_k$ and $\tilde{t_{ik}}$ with some interpretation relative to the model, e.g.
-as is [CITE GELMAN PAPER].
+as in [CITE GELMAN PAPER].
 For example, $\tilde{n}_k$ may represent the number of blocks that were rejected by the Bitcoin protocol for having a timestamp earlier than the median of the previous elevent blocks.
-While tempting this interpretation is not strictly speak correct.
+While tempting this interpretation is not strictly speaking correct.
 When a block is rejected by the protocol a new block does not appear instantaneously since it takes time for another miner to discover the block.
 So the arrival times and thus measurement error distributions of the rejected block and the accepted block are not the same.
-While the interpretation is not correct, it is good to be inspired by computational tricks to create better model [CITE SAME GELMAN PAPER] and perhaps something similar to this negative binomial structure can be used to add rejected blocks as a component of the model.
+While the interpretation is not correct, it is good to be inspired by computational tricks to create better model [CITE SAME GELMAN PAPER] and perhaps something similar to this negative binomial structure can be used to add rejected blocks as a component of the model, though we do not explore this possibility here.
 
-\subsection{measurement error variance}
+\subsection{Measurement error variance --- $\sigma^2$}
 The full conditional distribution of the measurement error variance, $\sigma^2$, depends only on $\tilde{\bm{t}}_{1:n_{b}}$ and $\tau_{1:n_{b}}$. It is
 \begin{align*}
-p(\sigma^2|...) \propto \sigma^{-(n_{b} + \sum_{k=1}^{n_{b}}\tilde{n}_k)}\exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\log \tilde{t}_{ik} - \log\tau_k)^2\right]p(\sigma^2)
+p(\sigma^2|...) \propto \sigma^{-(n_{b} + \sum_{k=1}^{n_{b}}\tilde{n}_k)}\exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\tilde{t}_{ik} - \tau_k)^2\right]p(\sigma^2)
 \end{align*}
 where $p(\sigma^2)$ is the density of the prior distribution on $\sigma^2$.
-We will suppose that $\sigma$ has a half-$t$ prior, $\sigma \sim t_\nu(S)1(\sigma>0)$ with degrees of freedom and scale parameter $S$.
-This prior is equivalent to the marginal prior on $\sigma$ when $\sigma^2|\omega \sim IG(\nu/2, \omega)$ and $\omega \sim G(1/2, 1/(\nu S^2))$.
-Then letting $a_\sigma = (n_{b} + \sum_{k=1}^{n_b}\tilde{n}_k)/2$ and $b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\log \tilde{t}_{ik} - \log\tau_k)^2$ the full conditional distribution of $(\sigma^2,\omega)$ is
+With the half-$t$ prior on $\sigma$ discussed above and letting $a_\sigma = (n_{b} + \sum_{k=1}^{n_b}\tilde{n}_k)/2$ and $b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\tilde{t}_{ik} - \tau_k)^2$ the full conditional distribution of $(\sigma^2,\omega)$ is
 \begin{align*}
-p(\sigma^2,\omega|...) \propto (\sigma^2)^{-(a_\sigma+\nu)/2-1}\left[-\frac{b_\sigma}{\sigma^2}\right]\exp\left[-\frac{\omega}{\sigma^2}\right]\omega^{(\nu + 1)/2 - 1}\exp\left[-\frac{\omega}{\nu S^2}\right].
+p(\sigma^2,\omega|...) \propto (\sigma^2)^{-(a_\sigma+v)/2-1}\exp\left[-\frac{b_\sigma}{\sigma^2}\right]\exp\left[-\frac{\omega}{\sigma^2}\right]\omega^{(v + 1)/2 - 1}\exp\left[-\frac{\omega}{v s^2}\right].
 \end{align*}
@@
 \begin{enumerate}
-\item Draw $\omega \sim G\left(\frac{\nu + 1}{2}, \frac{1}{\sigma^2} + \frac{1}{\nu S^2}\right)$ and form
+\item Draw $\omega \sim G\left(\frac{v + 1}{2}, \frac{1}{\sigma^2} + \frac{1}{v s^2}\right)$ and form
 \begin{align*}
-a_\sigma = n_{b} + \sum_{k=1}^{n_b}\tilde{n}_k &&\mbox{  and  }&& b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\log \tilde{t}_{ik} - \log\tau_k)^2/2;
+a_\sigma = n_{b} + \sum_{k=1}^{n_b}\tilde{n}_k &&\mbox{  and  }&& b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\tilde{t}_{ik} - \tau_k)^2/2;
 \end{align*}
-\item Draw $\sigma^2 \sim  IG([a_\sigma + \nu]/2, b_\sigma + \omega)$.
+\item Draw $\sigma^2 \sim  IG\left(\frac{a_\sigma + v}{2}, b_\sigma + \omega\right)$.
 \end{enumerate}
 
-\subsection{block arrival times}
+\subsection{Block arrival times --- $\tau_k$}
 The full conditional distribution of the block arrival times is more complicated:
 \begin{align*}
-p(\bm{\tau}_{1:n_{b}}|...) \propto \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\log\tilde{t}_{ik} - \log\tau_k\right)^2 - \frac{1}{10}\tau_{n_{b}}\right]\frac{\phi^{\phi[\Psi(\tau_k) - \Psi(\tau_{k-1})]}}{\Gamma\left(\phi[\Psi(\tau_k) - \Psi(\tau_{k-1})]\right)}\lambda_k^{\phi[\Psi(\tau_k)-\Psi(\tau_{k-1})] - 1}
+p(\bm{\tau}_{1:n_{b}}|...) \propto \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \tau_k\right)^2 - \frac{1}{10}\tau_{n_{b}}\right]\frac{\phi^{\Psi(\tau_k) - \Psi(\tau_{k-1})}}{\Gamma\left[\Psi(\tau_k) - \Psi(\tau_{k-1})\right]}\lambda_k^{\Psi(\tau_k)-\Psi(\tau_{k-1}) - 1}
 \end{align*}
-where $\tau_0\equiv0$ and $\Gamma(x) = \int_0^\infty t^{x-1}e{-t}dt$ is the gamma function.
-Even $\Psi(\tau)$ is some nice, known form, this density is nonstandard because the $\tau_k$'s enter the Gamma function.
-As a result, we have several options for this step:
+where $\tau_0\equiv0$ and $\Gamma(x) = \int_0^\infty t^{x-1}e^{-t}dt$ is the gamma function.
+Even if $\Psi(\tau)$ is some nice known form, this density is nonstandard because the $\tau_k$'s enter the Gamma function.
+As a result, we are limited to several options for this step:
 \begin{enumerate}
@@
 
-\subsection{transaction process dispersion}
-The transaction process dispersion parameter is $\phi$ --- $V[N(\tau)] = E[N(\tau)](1 + 1/\phi)$. It has the full conditional density
+\subsection{Transaction process dispersion --- $\phi$}
+The full conditional density of $\phi$ is
 \begin{align*}
-p(\phi|...) \propto \prod_{k=1}^{n_{b}}\frac{\phi^{\phi\psi_k}\lambda_k^{\phi\psi_k - 1}e^{\lambda_k\phi}}{\Gamma(\phi\psi_k)}p(\phi).
+p(\phi|...) \propto \prod_{k=1}^{n_{b}}\phi^{\psi_k}e^{\lambda_k\phi}p(\phi) \propto \phi^{\sum_{k=1}^{n_{b}}\psi_k}e^{\phi\sum_{k=1}^{n_{b}}\lambda_k}p(\phi).
 \end{align*}
-This is another complicated density, but ARS should work decently to sample from it.
-If not, various Metropolis options are available.
-Another option: transform so that $\phi$ is absorbed into the shape function.
-Then $\tilde{\phi}$ has the full conditional
-\begin{align*}
-p(\phi|...) \propto \phi^{\sum_{k=1}^{n_{b}}\psi_k}e^{\phi\sum_{k=1}^{n_{b}}\lambda_k}p(\phi).
-\end{align*}
-With $\phi \sim G(a_\phi, b_\phi)$ in the prior, this becomes
+With $\phi \sim G(a_\phi, b_\phi)$ in the prior, this is a Gamma density so the step is simple:
 \begin{enumerate}
@@
 
-\subsection{regression parameters}
+\subsection{Regression parameters --- $\bm{\beta}$}
+The full conditional posterior of $\bm{\beta}$ is closely connected to the functional form of the covariates in time.
+No matter the functional form, however, the density is complex and a Metropolis step is likely warranted.
+Recall that since we defined $\psi(t) = e^{\bm{z}(t)'\bm{\beta}}$ we have $\psi_k(\bm{\beta}) = \int_{\tau_{k-1}}^{\tau^k}e^{\bm{z}(t)'\bm{\beta}}dt$ --- in this subsection we write $\psi_k(\bm{\beta})$ to explicitly acknowledge that $\psi_k$ is a function of $\bm{\beta}$.
+Then the full conditional of $\bm{\beta}$ is
+\begin{align*}
+p(\bm{\beta}|\dots) \propto p(\bm{\beta})\prod_{k=1}^{n_b}\frac{\phi^{\psi_k(\bm{\beta})}\lambda_k^{\psi_k(\bm{\beta}) - 1}}{\Gamma[\psi_k(\bm{\beta})]}
+\end{align*}
+where in the prior $\bm{\beta} \sim p(\bm{\beta})$ independent of the other parameters. This density is complex and nonstandard since $\psi_k(\bm{\beta})$ enters the gamma function.
 
-\subsection{transaction arrival times and data amounts}
+\subsection{Transaction arrival times and sizes in MB --- $N(\tau_k)$ and $D_k$}
+First consider transaction sizes, $\bm{d}=(d_1,d_2,\dots,d_{n_d})'$.
+Let $\bm{d}_k$ denote the sizes of the $M_k - M_{k-1}$ transactions that were stored in block $k$ so that $\bm{d} = (\bm{d}_1',\dots,\bm{d}_{n_{b}}')'$.
+The full conditional distribution of the $\bm{d}_k$'s is
+\begin{align*}
+p(\bm{d}_{1:n_{b}}|...)\propto \prod_{k=1}^{n_b}\prod_{i=1}^{M_k - M_{k-1}} d_i^{\alpha_d - 1}e^{-d_i\beta_d}1\left(\sum_{i=1}^{M_k - M_{k-1}} d_{ik}= D_k\right)1\left(d_{1k} > \bar{D}_{k-1} - D_{k-1} \mathrm{\ or\ } M_{k-1} = N_{k-1}\right).
+\end{align*}
+The second indicator function is required because if $M_{k-1} < N_{k-1}$ then $N_{k-1} - M_{k-1}$ transactionss must have overflowed into block $k$.
+So the amount of data from the first transaction of block $k$, $d_{1k}$, must have been greater than the difference between the amount of data in block $k-1$ and the data cap for block $k-1$.
+Similarly if $M_{k-1} = N_{k-1}$ then we know there was no overflow into block $k$, so we have no reason to think that $d_{1k}$ has a restricted support.
+Let $\bm{o}_k = \bm{d}_k/D_k$.
+Then the $\bm{o}_k$'s have potentially truncated independent Dirichlet distributions.
+That is
+\begin{align*}
+p(\bm{o}_{1:n_b}|...)\propto \prod_{k=1}^{n_b} \prod_{i=1}^{M_k - M_{k-1}}o_{ik}^{\alpha_d - 1} 1\left(d_{1k} > [\bar{D}_{k-1} - D_{k-1}]/D_{k} \mathrm{\ or\ } M_{k-1} = N_{k-1}\right).
+\end{align*}
+So the $\bm{d}$ step to draw from $p(\bm{d}|...)$ is as follows: Let $\bm{1}_n$ denote an $n$-vector of ones. For $k=1,2,\dots,n_b$
+\begin{enumerate}
+\item Draw $\bm{o}_{k} \sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1}})1\big(d_{1k} > 1[N_{k-1} - M_{k-1} > 0][\bar{D}_{k-1} - D_{k-1}]/D_{k}\big)$.
+\item Set $\bm{d}_k = \bm{o}_kD_k$.
+\end{enumerate}
+When the dirichlet distribution is truncated the draw can be accomplished by rejection sampling, though other methods are available.
+[FILL IN WHICHEVER METHOD I END UP USING --- PROBABLY AUGMENTATION LIKE MY BLOG POST]
 
-\subsection{transaction data amount parameters}
+To draw the $N_k$'s first note that their support is restricted by the value of the $M_k$'s. Specifically for $k=1,2,\dots,n_b$ and $N_0 = 0$ we have $M_k \le N_k \le N_{k+1}$. Second, when 
+\begin{align*}
+\sum_{i=1}^{M_k - M_{k-1} + 1}d_{M_{k-1} + i} \le \bar{D}_k
+\end{align*}
+we know that $N_k = M_k$ because otherwise transaction $M_k + 1$ would be in block $k$.
+If the summation is greater than $D_k$ then it is still possible that $N_k = M_k$ because transaction $d_{M_k + 1}$ could still have arrived later than block $k$.
+In that case, the density of $N_k$ given $N_{0:(k-1)}$ and all other parameters and processes is given by
+\begin{align*}
+p(N_k|N_{0:(k-1)}, \dots) \propto \frac{\lambda_k^{N_k - N_{k-1}}e^{-\lambda_k}}{(N_k - N_{k-1})!}1(N_k \ge M_k)1(N_k \ge N_{k-1}).
+\end{align*}
+In other words, $\eta_k = N_k - N_{k-1}$ has a truncated Poisson distribution where the support is restricted to $\{M_k - N_{k-1}, M_k - N_{k-1} + 1, \dots \}$.
+[PUT INSTRUCTIONS FOR HOW TO DRAW FROM THE TRUNCATED POISSON HERE] So we can draw from $p(\bm{N}_{1:n_b}|\dots)$ as follows: For $k=1,2,\dots,n_b$
+\begin{enumerate}
+\item If $\sum_{i=1}^{M_k - M_{k-1} + 1} d_{M_k + i} \le \bar{D}_k$ set $N_k = M_k$.
+
+Otherwise, draw $\eta_k \sim Poi(\lambda_k)1(\eta_k \ge M_k - N_{k-1})$ and set $N_k = N_{k-1} + \eta_k$.
+\end{enumerate}
+\subsection{Block size caps --- $\bar{D}_k$}
+Suppose for $k=1,\dots,n_b$ we have independent categorical priors on each of the $\bar{D}_k$'s.
+That is $\bar{D}_k \sim Cat(\bm{a}_{1:n_{cat}}, \bm{p}_{k})$ where $P(\bar{D}_k = a_i) = p_{ik}$ so that $p_{ik}\ge 0$ for $i=1,2,\dots,n_{cat}$ and $\sum_{i=1}^{n_{cat}}p_{ik} = 1$, both for $k=1,2,\dots,n_b$.
+In the posterior we learn about $\bar{D}_k$ through the observed $D_k$'s.
+The caps are then independent in their full conditional and for each block $k$, there are two possibilities:
+\begin{enumerate}
+\item When $N_k > M_k$, $D_k \le \bar{D}_k < D_k + d_{M_k+1}$.
+\item When $N_k = M_k$, $D_k \le \bar{D}_k$.
+\end{enumerate}
+In the first case, this should completely determine the value of $\bar{D}_k$ since transaction sizes are much smaller than the differences between possible caps.
+For both cases we simple draw $\bar{D}_k$ from a categorical distribution where the possible categories are only those remaining which satisfy the relevant constraint fiven above, and the probability of obtaining each category is proportional to its probability in the prior.
+For example, suppose only $a_1$ and $a_2$ satisfy the relevant constraint for $\bar{D}_k$.
+Then we draw $\bar{D}_k$ with probabilities $P(\bar{D}_k = a_1) = p_{1k}/(p_{1k} + p_{2k})$ and $P(\bar{D}_k = a_2) = p_{2k}/(p_{1k} + p_{2k})$.
+\subsection{Transaction data parameters --- $\alpha_d$ and $\beta_d$}
+The full conditional distribution of $(\alpha_d,\beta_d)$ is 
+\begin{align*}
+p(\alpha_d,\beta_d|...) \propto p(\alpha_d)p(\beta_d) \prod_{i=1}^{n_d}\frac{\beta_d^{\alpha_d}}{\Gamma(\alpha_d)}d_i^{\alpha_d - 1}e^{-\beta_dd_i}.
+\end{align*}
+When $\beta \sim G(a_\beta, b_\beta)$ in the prior, its full conditional becomes
+\begin{align*}
+p(\beta_d|...) \propto \beta_d^{\alpha_dn_d + a_{\beta} - 1}e^{-\beta_d\left(\sum_{i=1}^{n_d}d_i + b_{\beta}\right)},
+\end{align*}
+which is another gamma distribution. The full conditional of $\alpha_d$ is complex no matter what its prior is since $\alpha_d$ enters the gamma function, and has the form
+\begin{align*}
+p(\alpha_d|...) \propto p(\alpha_d)\frac{\left(\prod_{i=1}^{n_d}d_i + \beta_d^{n_d}\right)^{\alpha_d}}{\Gamma(\alpha_d)^{n_d}}.
+\end{align*}
+This density can be drawn from using rejection sampling or an independence Metropolis step, e.g. ARS or ARMS.
 \bibliographystyle{apalike} 

### 2016-01-13 model/overflowmcmc.tex
Priors and algorithm sketch finished

@@
 \subsection{Priors on $\alpha_d$ and $\beta_d$}
+The conditionally conjugate prior for $beta_d$ is a gamma while $\alpha_d$ has no convenient conditionally conjugate form.
+We will assume that $\beta_d \sim G(a_\beta, b_\beta)$ and $\alpha_d \sim G(a_\alpha, b_\alpha)$.
+In order to choose $a_\alpha$, $b_\alpha$, $a_\beta$, and $b_\beta$, consider that $E[d_i|\alpha_d,\beta_d] = \alpha_d/\beta_d$.
+But
+\begin{align*}
+E\left[\frac{\alpha_d}{\beta_d}\right] = E[\alpha_d]E\left[\frac{1}{\beta_d}\right] = \frac{a_\alpha}{b_\alpha}\frac{b_\beta}{a_\beta - 1}.
+\end{align*}
+Similarly
+\begin{align*}
+V\left[\frac{\alpha_d}{\beta_d}\right] = V[\alpha_d] + V\left[\frac{1}{\beta_d}\right] = \frac{a_\alpha}{b_\alpha^2} + \frac{b_\beta^2}{(a_\beta - 1)^2(a_\beta - 2)}.
+\end{align*}
+We will choose these parameters so that $E[\alpha_d/\beta_d] \approx 0.01$ MB and $V[\alpha_d/\beta_d]$ is large so that the prior is weakly informative.
+We will set $a_\beta = 2.1$ to ensure the variance calculated above exists.
+Then $a_\alpha = 0.1$, $b_\alpha = 1$, and $b_\beta = 1$ yields
+\begin{align*}
+E\left[\frac{\alpha_d}{\beta_d}\right] = \frac{1}{21} \approx 0.09; && V\left[\frac{\alpha_d}{\beta_d}\right] \approx 10.
+\end{align*}
 
-\subsection{Priors on $\bar{D}_k$'s}
+\subsection{Priors on the $\bar{D}_k$'s}
+We will put a discrete prior on each of the $\bar{D}_k$'s.
+Theoretically a miner can put any upper limit she wants on a mined bitcoin, but in practice most miners use one of several possible values.
+We will use the set of values $\mathcal{D} = \{0.25, 0.35, 0.5, 0.75, 0.9, 0.95, 1\}$ [DOUBLE CHECK THESE --- IN PARTICULAR, ARE THERE OTHER COMMON VALUES?
+WHAT ABOUT THE ELGIUS MINING POOL'S CAPS?].
+For each $\bar{D}_k$ we will use a discrete uniform prior on $\mathcal{D}$ so that $P(\bar{D}_k = j) = 1/7$ for all $j \in \mathcal{D}$.
 
-Finally, the upper limits are modeled as a categorical distribution over a set of prespecified possible limits, $\Delta_1,\dots,\Delta_{n_{lim}}$ with probability $p_{ik} = P(\bar{D}_k = \Delta_i)$, $\sum_ip_{ik} = 1$ and $p_{ik} \ge 0$.
-In other words
-\begin{align*}
-\bar{D}_k\stackrel{ind}{\sim} Cat(\Delta_{1:n_{lim}}, p_{1:n_{lim},k}).
-\end{align*}
+[OPTIONS FOR MORE INTERESTING THINGS HERE: DIFFERENT PRIOR FOR EACH $k$ SO THAT LATER BLOCKS HAVE A HIGHER PROBABILITY OF HAVING A HIGHER CAP.
+THIS USES THE DATA TWICE, SO A BETTER OPTION MAY JUST BE MODELING THE PROBABILITIES ON $\mathcal{D}$ AS EVOLVING OVER TIME BASED ON SOME HIGHER LEVEL PARAMETERS]
 
@@
 Even if $\Psi(\tau)$ is some nice known form, this density is nonstandard because the $\tau_k$'s enter the Gamma function.
-As a result, we are limited to several options for this step:
+Some ideas for handlilng this step:
 \begin{enumerate}
@@
 \item Approximate the density of $\tau_{1:n_{b}}$ sufficiently well and use an independent Metropolis step.
-\end{enumerate}
-Options 2 and 3 could both be replaced with a rejection sampling step under some circumstances, but a Metropolis step is likely cheaper computationally [CITE THE LIU PAPER].
-
-Ideas:
-\begin{enumerate}
-\item ARMS -- ARS except with a Metropolis correction for non-log concave densities (look it up).
-\item RW Metropolis and tune the proposal variances.
+\item Use adaptive rejection sampling (ARS) to draw from the density of $\tau_k$.
+\item Use ARMS --- ARS except with a Metropolis correction for densities which are not log concave (look it up).
 \end{enumerate}
@@
 \end{align*}
-where in the prior $\bm{\beta} \sim p(\bm{\beta})$ independent of the other parameters. This density is complex and nonstandard since $\psi_k(\bm{\beta})$ enters the gamma function.
+where in the prior $\bm{\beta} \sim p(\bm{\beta})$ independent of the other parameters.
+This density is complex and nonstandard since $\psi_k(\bm{\beta})$ enters the gamma function.
+A Metropolis step is likely warranted here as well.
 
@@
 \subsection{Block size caps --- $\bar{D}_k$}
-Suppose for $k=1,\dots,n_b$ we have independent categorical priors on each of the $\bar{D}_k$'s.
-That is $\bar{D}_k \sim Cat(\bm{a}_{1:n_{cat}}, \bm{p}_{k})$ where $P(\bar{D}_k = a_i) = p_{ik}$ so that $p_{ik}\ge 0$ for $i=1,2,\dots,n_{cat}$ and $\sum_{i=1}^{n_{cat}}p_{ik} = 1$, both for $k=1,2,\dots,n_b$.
-In the posterior we learn about $\bar{D}_k$ through the observed $D_k$'s.
-The caps are then independent in their full conditional and for each block $k$, there are two possibilities:
+Since the caps are independent discrete uniform in the prior, they are independent in their full conditional and for each block $k$, there are two possibilities:
 \begin{enumerate}
@@
 \end{enumerate}
-In the first case, this should completely determine the value of $\bar{D}_k$ since transaction sizes are much smaller than the differences between possible caps.
-For both cases we simple draw $\bar{D}_k$ from a categorical distribution where the possible categories are only those remaining which satisfy the relevant constraint fiven above, and the probability of obtaining each category is proportional to its probability in the prior.
-For example, suppose only $a_1$ and $a_2$ satisfy the relevant constraint for $\bar{D}_k$.
-Then we draw $\bar{D}_k$ with probabilities $P(\bar{D}_k = a_1) = p_{1k}/(p_{1k} + p_{2k})$ and $P(\bar{D}_k = a_2) = p_{2k}/(p_{1k} + p_{2k})$.
+In the first case this should completely determine the value of $\bar{D}_k$ since transaction sizes are much smaller than the differences between possible caps.
+In both cases we draw $\bar{D}_k$ from a discrete uniform distribution over the set of possible caps given the relevant constraint above --- sometimes that distribution puts all of the mass on a single value.
+
 \subsection{Transaction data parameters --- $\alpha_d$ and $\beta_d$}
@@
 \end{align*}
-When $\beta \sim G(a_\beta, b_\beta)$ in the prior, its full conditional becomes
+When $\beta_d \sim G(a_\beta, b_\beta)$ in the prior, its full conditional becomes
 \begin{align*}
@@
 \begin{align*}
-p(\alpha_d|...) \propto p(\alpha_d)\frac{\left(\prod_{i=1}^{n_d}d_i + \beta_d^{n_d}\right)^{\alpha_d}}{\Gamma(\alpha_d)^{n_d}}.
+p(\alpha_d|...) &\propto p(\alpha_d)\frac{\left(\beta_d^{n_d}\prod_{i=1}^{n_d}d_i\right)^{\alpha_d}}{\Gamma(\alpha_d)^{n_d}}\\
+\intertext{which in the case of a gamma prior on $\alpha_d$ is}
+p(\alpha_d|...) &\propto \alpha_d^{a_\alpha - 1}\frac{\left(e^{-b_\alpha}\beta_d^{n_d}\prod_{i=1}^{n_d}d_i\right)^{\alpha_d}}{\Gamma(\alpha_d)^{n_d}}.
 \end{align*}

### 2016-01-13 model/overflowmcmc.tex
Began writing the algorithm -- just the rejection/metropolis steps left. MCMC document updated on the fly to reflect specific algorithms used for e.g. truncated distributions

@@
 \begin{enumerate}
-\item Draw $\omega \sim G\left(\frac{v + 1}{2}, \frac{1}{\sigma^2} + \frac{1}{v s^2}\right)$ and form
+\item Draw $\omega \sim G\left(\frac{v + 1}{2}, \frac{1}{\sigma^2} + \frac{1}{v s^2}\right)$ and form\footnote{The inconsistency in notation between the definition of $a_\sigma$ and $b_\sigma$ is a consequence of matching notation to the R code supplied for implementing this MCMC algorithm.}
 \begin{align*}
-a_\sigma = n_{b} + \sum_{k=1}^{n_b}\tilde{n}_k &&\mbox{  and  }&& b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\tilde{t}_{ik} - \tau_k)^2/2;
+a_\sigma = \frac{n_{b} + \sum_{k=1}^{n_b}\tilde{n}_k + v}{2} &&\mbox{  and  }&& b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\tilde{t}_{ik} - \tau_k)^2;
 \end{align*}
-\item Draw $\sigma^2 \sim  IG\left(\frac{a_\sigma + v}{2}, b_\sigma + \omega\right)$.
+\item Draw $\sigma^2 \sim  IG\left(a_\sigma, b_\sigma/2 + \omega\right)$.
 \end{enumerate}
@@
 \begin{enumerate}
-\item Draw $\bm{o}_{k} \sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1}})1\big(d_{1k} > 1[N_{k-1} - M_{k-1} > 0][\bar{D}_{k-1} - D_{k-1}]/D_{k}\big)$.
+\item If $N_k = M_k$ draw $\bm{o}_{k} \sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1}})$.
+
+  Otherwise draw $\bm{o}_{k} \sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1}})1(d_{1k} > [\bar{D}_{k-1} - D_{k-1}]/D_k)$.
 \item Set $\bm{d}_k = \bm{o}_kD_k$.
 \end{enumerate}
-When the dirichlet distribution is truncated the draw can be accomplished by rejection sampling, though other methods are available.
-[FILL IN WHICHEVER METHOD I END UP USING --- PROBABLY AUGMENTATION LIKE MY BLOG POST]
+To draw from the truncated Dirichlet distribution above, first note that by the properties of the Dirhichlet the marginal distribution of $o_{1k}$ is $Beta(\alpha_d, \alpha_d(M_k - M_{k-1} - 1))$ and conditional on $o_{1k}$, $\bm{o}_{-1,k}/(1 - o_{1k}) \sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1} - 1})$.
+Then $o_{1k}$ can be drawn from the truncated Beta distribution using the inverse cdf method, resulting in the following two step algorithm to draw from the desired truncated Dirichlet distribution:
+\begin{enumerate}
+\item Draw $u \sim U(0,1)$ and set $o_{1k} = F^{-1}(u*[1 - F((\bar{D}_{k-1} - D_{k-1})/D_k)])$ where $F$ is the cdf of the $Beta(\alpha_d, \alpha_d(M_k - M_{k-1} - 1))$ distribution.
+\item Draw $\tilde{\bm{o}}_{2:(M_k - M_{k-1})}\sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1} - 1})$ and set $o_{ik} = \tilde{o}_i$ for $i = 2,3,\dots,M_k - M_{k-1}$.
+\end{enumerate}
 
@@
 In other words, $\eta_k = N_k - N_{k-1}$ has a truncated Poisson distribution where the support is restricted to $\{M_k - N_{k-1}, M_k - N_{k-1} + 1, \dots \}$.
-[PUT INSTRUCTIONS FOR HOW TO DRAW FROM THE TRUNCATED POISSON HERE] So we can draw from $p(\bm{N}_{1:n_b}|\dots)$ as follows: For $k=1,2,\dots,n_b$
+We can easily simulate from this distribution using the inverse cdf method.
+So we can draw from $p(\bm{N}_{1:n_b}|\dots)$ as follows: For $k=1,2,\dots,n_b$
 \begin{enumerate}

### 2016-01-13 model/overflowmcmc.R

@@ -1,0 +1,83 @@
+library(MCMCpack)
+library(truncnorm) ## rtruncnorm is faster than rtnorm in the msm package
+
+## steps left: tau, beta, alpha_d
+## all will be ARS / ARMS / Metropolis steps
+## then put it all together and test
+
+aomega <- (vsigma + 1)/2
+bomega <- 1/(vsigma + s^2)
+
+tfull <- list()
+
+blocks <- matrix(1:nblocks, ncol = 1)
+
+listdiffsq <- function(k, tlist, tauvec){
+  out <- sum((tlist[[k]] - tauvec[k])^2)
+  return(out)
+}
+
+## DA step: \tilde{n}_k and \tilde{t}_k
+## Easily parallelizable
+bsigma <- 0 ## running tally, used in sigma step
+for(k in 1:nblock){
+  rho <- pnorm((m[k] - tau[k])/sigma) ## rho = P(t_k < m_k) = P(fail)
+  nrej[k] <- rnbinom(1, 1, prob = 1 - rho)
+  if(nrej[k] > 0){
+    tmiss <- rtruncnorm(nrej[k], mean = tau[k], sd = sigma, a = -Inf, b = m[k])
+    tfull[[k]] <- c(tmiss, tobs[k])
+  } else {
+    tfull[[k]] <- tobs[k]
+  }
+  bsigma <- bsigma + sum((tfull[[k]] - tau[k])^2)
+}
+
+## sigma / sigma^2 step
+omega <- rgamma(aomega, bomega + 1/sigma)
+asigma <- (nblocks + sum(nrej) + vsigma)/2
+sigma2 <- 1/rgamma(1, asigma, bsigma/2 + omega )
+sigma <- sqrt(sigma^2)
+
+## tau step -> complicated, probably not parallelizable
+
+## phi step
+phi <- rgamma(1, aphi + sum(psi), bphi + sum(lambda))
+
+## beta step -> complicated
+
+## d step
+nds <- diff(M) ## M[1] = 0 and N[1] = 0
+d <- list()
+for(k in 1:nblocks){
+  if(N[k] == M[k]){
+    d[[k]] <- rdirichlet(1, rep(alphad, nds[k]))*D[k]
+  } else {
+    u <- runif(1)
+    pupper <- pbeta(d[k], alphad, betad)
+    d1 <- qbeta(pupper + u * (1 - pupper), alphad, betad)
[…]

### 2016-01-14 model/overflowmcmc.tex
Fixed a couple typos

@@
 An easy way to achieve this is to define $\Psi(\tau) = \int_0^\tau\psi(t)dt$ where $\psi(t)$ is continuous and nonnegative.
-While $\psi(t)$ can be thought of as the intensity of the Gamma process it is also intimately related to the intensity of the Poission process.
+While $\psi(t)$ can be thought of as the intensity of the gamma process it is also intimately related to the intensity of the Poission process.
 The mean of $N(\tau + \delta) - N(\tau)$ is
@@
 In this way, $\psi(t)/\phi$ can be seen as the expected intensity function of the $N(\tau)$ process.
-The Gamma-Poisson structure also allows for a more flexible model on $N(\tau)$ by allowing for overdispersion.
+The gamma-Poisson structure also allows for a more flexible model on $N(\tau)$ by allowing for overdispersion.
 In a Poisson model the mean and variance are the same, but in this case they are proportional:
@@
 \begin{align*}
-p(\bm{\tau}_{1:n_{b}}|...) \propto \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \tau_k\right)^2 - \frac{1}{10}\tau_{n_{b}}\right]\frac{\phi^{\Psi(\tau_k) - \Psi(\tau_{k-1})}}{\Gamma\left[\Psi(\tau_k) - \Psi(\tau_{k-1})\right]}\lambda_k^{\Psi(\tau_k)-\Psi(\tau_{k-1}) - 1}
+p(\bm{\tau}_{1:n_{b}}|...) \propto \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \tau_k\right)^2 - \frac{1}{10}\tau_{n_{b}}\right]\prod_{k=1}^{n_b}\frac{\phi^{\Psi(\tau_k) - \Psi(\tau_{k-1})}}{\Gamma\left[\Psi(\tau_k) - \Psi(\tau_{k-1})\right]}\lambda_k^{\Psi(\tau_k)-\Psi(\tau_{k-1}) - 1}
 \end{align*}
 where $\tau_0\equiv0$ and $\Gamma(x) = \int_0^\infty t^{x-1}e^{-t}dt$ is the gamma function.
-Even if $\Psi(\tau)$ is some nice known form, this density is nonstandard because the $\tau_k$'s enter the Gamma function.
-Some ideas for handlilng this step:
+Even if $\Psi(\tau)$ is some nice known form, this density is nonstandard because the $\tau_k$'s enter the gamma function.
+Some ideas for handling this step:
 \begin{enumerate}
@@
 \end{align*}
-With $\phi \sim G(a_\phi, b_\phi)$ in the prior, this is a Gamma density so the step is simple:
+With $\phi \sim G(a_\phi, b_\phi)$ in the prior, this is a gamma density so the step is simple:
 \begin{enumerate}

### 2016-01-21 model/overflowmcmc.R
Four no-trend algorithms now working; trend algorithm has numerical problems

@@ -1,83 +1,1126 @@
 library(MCMCpack)
-library(truncnorm) ## rtruncnorm is faster than rtnorm in the msm package
+library(truncnorm) ## rtruncnorm() is faster than rtnorm() in the msm package
+library(ars)
+library(compiler)
 
-## steps left: tau, beta, alpha_d
-## all will be ARS / ARMS / Metropolis steps
-## then put it all together and test
-
-aomega <- (vsigma + 1)/2
-bomega <- 1/(vsigma + s^2)
-
-tfull <- list()
-
-blocks <- matrix(1:nblocks, ncol = 1)
-
-listdiffsq <- function(k, tlist, tauvec){
-  out <- sum((tlist[[k]] - tauvec[k])^2)
+ftaumcmc <- function(niter, data, initial, prior, rwsds){
+  ## Initial values
+  tau <- initial$tau
+  tau0 <- c(0, tau)
+  delta <- diff(tau0)
+  nblocks <- length(tau)
+  sigma <- initial$sigma
+  lambda <- initial$lambda
+  gam <- initial$gam
+  phi <- initial$phi
+  N <- initial$N
+  Ntotal <- N[nblocks + 1]
+  Dbar <- initial$Dbar
+  alphad <- initial$alphad
+  betad <- initial$betad
+  ## RW sds and bookkeeping
+  taulogrwsd <- rwsds$taulogrwsd
+  gamlogrwsd <- rwsds$gamlogrwsd
+  alphadlogrwsd <- rwsds$alphadlogrwsd
+  taurwsd <- exp(taulogrwsd)
+  gamrwsd <- exp(gamlogrwsd)
+  alphadrwsd <- exp(alphadlogrwsd)
+  taurejs <- matrix(0, ncol = 1, nrow = niter)
+  gamrejs <- matrix(0, ncol = 1, nrow = niter)
+  alphadrejs <- matrix(0, ncol = 1, nrow = niter)
+  rwc <- rwsds$rwc
+  H <- rwsds$H
+  tune <- rwsds$tune
+  lowtarget <- rwsds$lowtarget
+  hightarget <- rwsds$hightarget
+  tauchol <- rwsds$tauchol
+  ## Priors
+  vsigma <- prior$vsigma
+  ssigma <- prior$ssigma
+  aphi <- prior$aphi
+  bphi <- prior$bphi
+  agam <- prior$agamma
+  bgam <- prior$bgamma
+  Dbars <- prior$Dbars
+  aalpha <- prior$aalpha
[…]

### 2016-01-21 model/overflowmcmcout.R
Created for testing algorithms - a mess right now

@@ -1,0 +1,334 @@
+source("overflowmcmc.R")
+library(MCMCpack)
+
+load("../charts/fullblocks.RData")
+cleanblocks <- blocks[,c("height", "time", "transactions_count", "size")]
+cleanblocks$elapsed.min<- c(NA, diff(cleanblocks$time))/60
+
+nblocks <- 100
+
+block1k <- cleanblocks[(nrow(cleanblocks) - nblocks - 10):nrow(cleanblocks),]
+block1k$tobs <- cumsum(block1k$elapsed.min)
+block1k$ms <- 0
+for(i in 1:nblocks){
+  block1k$ms[i + 11] <- median(block1k$tobs[i:(i + 10)])
+}
+
+datain <- list()
+datain$tobs <- block1k$tobs[11 + 1:nblocks] - block1k$tobs[10]
+datain$m <- block1k$ms[11 + 1:nblocks] - block1k$tobs[10]
+datain$M <- c(0, cumsum(block1k$transactions_count[11 + 1:nblocks]))
+datain$D <- block1k$size[11 + 1:nblocks]/1000 ## convert to kilobytes
+
+lfun <- function(a, m, d){
+  out <- rep(0, length(a))
+  M <- sum(m)
+  dbar <- sum(d)/M
+  b <- a/dbar
+  for(i in 1:length(a)){
+    out[i] <- sum(dgamma(d, a[i]*m, b[i], log=TRUE))
+  }
+  return(out)
+}
+
+ahat <- optimize(lfun, c(0.00001, 1000), m = diff(datain$M), d = datain$D, maximum = TRUE)$maximum
+bhat <- ahat/(sum(datain$D)/datain$M[nblocks + 1])
+
+
+ssigma <- 1
+vsigma <- 1
+aphi <- 1/50
+bphi <- 1/100
+agamma <- 1
+bgamma <- 1
+balpha <- 1/10
+aalpha <- ahat*balpha ## 1/10
+bbeta <- 1/10
+abeta <- bhat*bbeta
+Dbars <- c(0.25, 0.35, 0.5, 0.75, 0.9, 0.95, 1)*1000
+priorin <- list(ssigma = ssigma, vsigma = vsigma, aphi = aphi, bphi = bphi, agamma = agamma,
+                bgamma = bgamma,
[…]

### 2016-01-21 model/overflowmcmc.tex
Changed to reflect changes in the actual algorithm used; currently out of date compared to code

@@
 \begin{align*}
-p(\bm{\tau}_{1:n_{b}}|...) \propto \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \tau_k\right)^2 - \frac{1}{10}\tau_{n_{b}}\right]\prod_{k=1}^{n_b}\frac{\phi^{\Psi(\tau_k) - \Psi(\tau_{k-1})}}{\Gamma\left[\Psi(\tau_k) - \Psi(\tau_{k-1})\right]}\lambda_k^{\Psi(\tau_k)-\Psi(\tau_{k-1}) - 1}
+p(\bm{\tau}_{1:n_{b}}|...) \propto & \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \tau_k\right)^2 - \frac{1}{10}\tau_{n_{b}}\right]\\
+&\times \prod_{k=1}^{n_b}\frac{\phi^{\Psi(\tau_k) - \Psi(\tau_{k-1})}}{\Gamma\left[\Psi(\tau_k) - \Psi(\tau_{k-1})\right]}\lambda_k^{\Psi(\tau_k)-\Psi(\tau_{k-1}) - 1}1(0\le \tau_1 \le \tau_2 \le \hdots \le \tau_{n_b})\\
+\propto & \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \tau_k\right)^2 - \frac{1}{10}\tau_{n_{b}}\right] \phi^{\Psi(\tau_{n_b})}\\
+&\times \prod_{k=1}^{n_b}\frac{\lambda_k^{\Psi(\tau_k)-\Psi(\tau_{k-1})}}{\Gamma\left[\Psi(\tau_k) - \Psi(\tau_{k-1})\right]}1(0\le \tau_1 \le \tau_2 \le \hdots \le \tau_{n_b})\\
 \end{align*}
@@
 \end{enumerate}
+The full conditional distribution of $\tau_k$ for $k=1,2,\dots,n_{b}-1$ is
+\begin{align*}
+p(\tau_k|...) \propto \exp\left[-\frac{1}{2\sigma^2}\sum_{i=1}^{\tilde{n}_k + 1}(\tilde{t}_{ik} - \tau_k)^2\right] \frac{\left(\frac{\lambda_k}{\lambda_{k+1}}\right)^{\Psi(\tau_k)}}{\Gamma\left[\Psi(\tau_{k+1}) - \Psi(\tau_k)\right]\Gamma\left[\Psi(\tau_{k}) - \Psi(\tau_{k-1})\right]}1(\tau_{k-1}\le \tau_k \le \tau_{k+1})
+\end{align*}
+and the full conditional of $\tau_{n_b}$ is
+\begin{align*}
+p(\tau_{n_b}|...) \propto \exp\left[-\frac{1}{2\sigma^2}\sum_{i=1}^{\tilde{n}_{n_b} + 1}(\tilde{t}_{in_b} - \tau_{n_b})^2\right] \frac{\lambda_{n_b}^{\Psi(\tau_{n_b})}}{\Gamma\left[\Psi(\tau_{n_b}) - \Psi(\tau_{n_b-1})\right]}1(\tau_{n_b-1}\le \tau_{n_b}).
+\end{align*}
+Define the sets $A_k = [\tau_{k-1}, \tau_{k+1}$ for $k = 1, 2, \dots, n_b-1$ and $A_{n_b} = [\tau_{n_b -1}, \infty)$.
+Let $l_k$ and $u_k$ denote the lower and upper bounds of $A_k$ respectivfely.
+Then we will use a random walk Metropolis algorithm on each coordinate with a truncated normal proposal distribution $N_{A_k}(\tau_k, h_k^2)$.
+Then the acceptance rate of the Metropolis step for $\tau_k$ is
+\begin{align*}
+R_k = \frac{p(\tau_k^{(prop)}|\dots)\left[\Phi\left(\frac{u_k - \tau_k^{(old)}}{h_k}\right) - \Phi\left(\frac{l_k - \tau_k^{(old)}}{h_k}\right)\right]}{p(\tau_k^{(old)}|\dots)\left[\Phi\left(\frac{u_k - \tau_k^{(prop)}}{h_k}\right) - \Phi\left(\frac{l_k - \tau_k^{(prop)}}{h_k}\right)\right]}.
+\end{align*}
+where $\tau_k^{(old)}$ is the previous value in the chain and $\tau_k^{(prop)}$ is the proposed value. So the $\tau_k$ step is as follows:
+\begin{enumerate}
+\item Draw $\tau_k^{(prop)} \sim N_{A_k}(\tau_k^{(t)},h_k^2)$ and compute $R_k$.
+\item Set $\tau_k^{(t+1)} = \tau_k^{(prop)}$ with probability $\min(R_k, 1)$ and $\tau_k^{(t+1)} = \tau_k^{(t)}$ otherwise.
+\end{enumerate}
+For each $k$ the scale parameter $h_k$ must be chosen by the user, which is practically impossible with a large number of blocks in the analysis.
+So we automatically tune $h_k$ during the the burn in period as follows [CITATION]:
+\begin{enumerate}
+\item Run the algorithm above for $H$ iterations.
+\item For each $k$, compute the observed acceptance rate of the last $H$ iterations: $r_k\#(\tau_k^{(t)} = \tau_k^{(t-1)})/H$.
+\item If $r_k$ is larger than some target range $[r_l^*, r_h^*]$ update $\log h_k = \log h_k + c$; if lower update $\log h_k = \log h_k - c$.
+\end{enumerate}
+In certain simple cases the optimal acceptance rate for a one-dimensional target has been computed to be approximately $0.44$, so we use the range $[0.4, 0.45]$ as a rough guideline [CITATION].
+In order to ensure that the Markov chain converges to the target posterior of distribution this adaptive process must asymptotically stop in some sense --- restricting the process to the burn in is a sufficient condition.
+In practice as the chain converges in can help to start $H$ small and $c$ large so that the $h_k$'s quickly move to reasonable values, then over time increase $H$ and decrease $c$ for more precise fine tuning.
+We start with $H=50$ and $c=0.1$
+
+The previous algorithm for the $\tau_k$ steps works but will suffer from slow convergence because the $\tau_k$'s are being updated separately instead of in a block.
+It is helpful for getting a crude estimate of $\Sigma$, the posterior covariance matrix for $\tau_{1:n_b}$.
+Given this estimate $\hat{\Sigma}$, we can implement a blockwise Metropolis update for $\tau_{1:n_b}$, which will substantially decrease the number of draws required to adequately characterize the posterior distribution after burn in.
+First we transform to $\delta_{1:n_b}$ where $\delta_k = \tau_k - \tau_{k-1}$, then to $\zeta_k = \log\delta_k$ to obtain
+\begin{align*}
+p(\zeta_{1:n_b}|...) &\propto  \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \sum_{l=1}^ke^{\zeta_l}\right)^2 - \frac{1}{10}\sum_{k=1}^{n_{b}}e^{\zeta_k} + \sum_{k}^{n_b}\zeta_k\right]\\
+&\times \phi^{\Psi(\sum_{k=1}^{n_b}e^{\zeta_k})}\prod_{k=1}^{n_b}\frac{\lambda_k^{\Psi(\sum_{l=1}^{k}e^{\zeta_l})-\Psi(\sum_{l=1}^{k-1}e^{\zeta_l})}}{\Gamma\left[\Psi(\sum_{l=1}^ke^{\zeta_l}) - \Psi(\sum_{l=1}^{k-1}e^{\zeta_l})\right]}.
+\end{align*}
+Then the Metropolis proposal is drawn from a $N(\zeta_{1:n_b}^{(t)}, h^2\hat{\Sigma})$ distribution where $\zeta_{1:n_b}^{(t)}$ is the previous value of $\zeta_{1:n_b}$ in the chain, $h$ is a user chosen scale parameter, and $\hat{\Sigma}$ is a crude estimate of the posterior covariance matrix for $\zeta_{1:n_b}$ obtained using the single dimensional random walk Metropolis algorithm used above.
+Then $h$ can be tuned just like the $h_k$'s with a target rate of about $0.24$ that is more appropriate for the high dimensional case, though again this rate was derived for simpler models than our own [CITATION].
+If the correlation structure of the $\zeta$'s depends heavily on the values of the other parameters in the model then this algorithm may reject too frequently no matter the value of $h$ because $\hat{\Sigma}$ is a crude estimate of the {\it marginal} posterior for $\zeta_{1:n_b}$.
+In that case we can simply revert to the original one dimensional random walk Metropolis algorithm.
 
@@
 This density is complex and nonstandard since $\psi_k(\bm{\beta})$ enters the gamma function.
-A Metropolis step is likely warranted here as well.
+In the case of $\bm{\beta} = (beta_1, \beta_2)'$ and $\bm{z}(t) = (1, t)'$ we use separate random walk Metropolis steps with a normal proposal and separately tuned standard deviations --- analogous to the algorithm used for the $\tau_k$'s above.
 
@@
 \end{align*}
-This density can be drawn from using rejection sampling or an independence Metropolis step, e.g. ARS or ARMS.
+This density is log concave so we will use adaptive rejection sampling to draw from it [CITATION].
 \bibliographystyle{apalike} 

### 2016-01-22 charts/charts.R
Checked that the variance of transactions sizes is linear in the mean, which confirms gamma distribution for modeling this piece.

@@ -139,3 +139,38 @@ summary(weightedreg1)
 summary(weightedreg2)
 summary(weightedreg3)
 summary(weightedreg4)
+
+
+### added for checking basic distributional assumptions about transaction sizes
+### upshot: gamma distribution seems reasonable (variance of trans size is linear in mean)
+weights <- blocks$transactions_count
+sizes <- blocks$size
+means <- weights/sizes
+heights <- blocks$height
+maxheight <- max(heights)
+
+sequence <- seq(1,maxheight,by=500)
+weightedsummary <- NULL
+for(i in 1:length(sequence)){
+  tempsizes <- means[heights >= sequence[i] & heights < sequence[i] + 500]
+  tempweights <- weights[heights >= sequence[i] & heights < sequence[i] + 500]
+  tempmean <- wtd.mean(tempsizes, tempweights, TRUE)
+  tempvar <- wtd.var(tempsizes, tempweights, TRUE)
+  tempsd <- sqrt(tempvar)
+  totalweight <- sum(tempweights)
+  weightedsummary <- rbind(weightedsummary,c(i,tempmean,tempvar,tempsd,totalweight))
+}
+colnames(weightedsummary) <- c("id", "mean", "var", "sd", "wt")
+weightedsummary <- data.frame(weightedsummary)
+
+plot(weightedsummary$mean, weightedsummary$var)
+
+weightedreg1 <- lm(var ~ mean, data = weightedsummary, weights = wt)
+weightedreg2 <- lm(var ~ mean + I(mean^2), data = weightedsummary, weights = wt)
+weightedreg3 <- lm(var ~ mean + I(mean^2) + I(mean^3), data = weightedsummary, weights = wt)
+weightedreg4 <- lm(var ~ mean + I(mean^2) + I(mean^3) + I(mean^4), data = weightedsummary, weights = wt)
+
+summary(weightedreg1)
+summary(weightedreg2)
+summary(weightedreg3)
+summary(weightedreg4)

### 2016-02-18 model/overflowmcmc.R
Major debugging and added cpp version of the overflow trend model - still calls a couple functions from R

@@ -2,32 +2,35 @@ library(MCMCpack)
 library(truncnorm) ## rtruncnorm() is faster than rtnorm() in the msm package
 library(ars)
 library(compiler)
+library(aster)
 
+## no trend, no overflow, full tau step
 ftaumcmc <- function(niter, data, initial, prior, rwsds){
   ## Initial values
   tau <- initial$tau
-  tau0 <- c(0, tau)
-  delta <- diff(tau0)
   nblocks <- length(tau)
+  tau0 <- c(0, tau[-nblocks])
+  delta <- diff(c(0,tau))
   sigma <- initial$sigma
-  lambda <- initial$lambda
   gam <- initial$gam
   phi <- initial$phi
   N <- initial$N
+  eta <- diff(N)
   Ntotal <- N[nblocks + 1]
   Dbar <- initial$Dbar
-  alphad <- initial$alphad
-  betad <- initial$betad
+  alpha <- initial$alpha
+  beta <- initial$beta
+  psi <- psifunC(tau, c(gam, 0), tau0)
   ## RW sds and bookkeeping
   taulogrwsd <- rwsds$taulogrwsd
   gamlogrwsd <- rwsds$gamlogrwsd
-  alphadlogrwsd <- rwsds$alphadlogrwsd
+  alphalogrwsd <- rwsds$alphalogrwsd
   taurwsd <- exp(taulogrwsd)
   gamrwsd <- exp(gamlogrwsd)
-  alphadrwsd <- exp(alphadlogrwsd)
+  alpharwsd <- exp(alphalogrwsd)
   taurejs <- matrix(0, ncol = 1, nrow = niter)
   gamrejs <- matrix(0, ncol = 1, nrow = niter)
-  alphadrejs <- matrix(0, ncol = 1, nrow = niter)
+  alpharejs <- matrix(0, ncol = 1, nrow = niter)
   rwc <- rwsds$rwc
   H <- rwsds$H
   tune <- rwsds$tune
@@ -39,6 +42,8 @@ ftaumcmc <- function(niter, data, initial, prior, rwsds){
   ssigma <- prior$ssigma
   aphi <- prior$aphi
   bphi <- prior$bphi
+  mugam <- prior$mugamma
+  sig2gam <- prior$sig2gamma
   agam <- prior$agamma
   bgam <- prior$bgamma
   Dbars <- prior$Dbars
@@ -55,20 +60,20 @@ ftaumcmc <- function(niter, data, initial, prior, rwsds){
   Mtotal <- M[nblocks + 1]
[…]

### 2016-02-18 model/overflowmcmcC.cpp

@@ -1,0 +1,425 @@
+/* To do:
+   4) several functions called from R - need to written in C++
+   5) N_K = M_K must be enforced. Needs to be fixed.
+*/
+
+//#include <Rcpp.h>
+#include <RcppArmadillo.h>
+#include <RcppArmadilloExtensions/sample.h>
+// [[Rcpp::depends(RcppArmadillo)]]
+
+using namespace Rcpp;
+using namespace RcppArmadillo;
+
+// used for generating log gamma distributed random variates when shape < 1
+arma::vec rloggamma(int n, double shape, double scale){
+    arma::vec temp = log(arma::randg(n, arma::distr_param(shape + 1.0, 1.0))) + 
+      log(arma::randu(n))/shape + log(scale);
+  return temp;
+}
+
+// simple rejection sampler for gamma distributed random variates
+arma::vec rgammarej(double shape, double scale, double lower){
+  arma::vec x = arma::randg(1, arma::distr_param(shape, scale));
+  while(x(0) < lower){
+    x = arma::randg(1, arma::distr_param(shape, scale));
+  }
+  return x;
+}
+
+// compute psi vector
+// [[Rcpp::export]]
+arma::vec psifun(arma::vec tau, arma::vec gam, arma::vec tau0){
+  arma::vec deltau = tau - tau0;
+  arma::vec out(tau.size());
+  if(gam(1) == 0){
+    out = deltau*exp(gam(0));
+  } else {
+    // abs(arma::vec) works as expected
+    // abs(double) coerces to integer, causing lots of problems
+    // std::abs(double)
+    out = log(abs(exp(deltau*gam(1)) - 1)) + gam(0) + gam(1)*tau0 - log(std::abs(gam(1)));
+    out = exp(out);
+  }
+  return out;
+}
+
+// [[Rcpp::export]]
+List overtrendmcmcCPP(int niter, arma::vec tobs, arma::vec D, arma::uvec M, arma::vec m,
+			 arma::vec tau, double sigma, arma::vec gam, double phi, arma::uvec N,
+			 arma::vec Dbar, double alpha, double beta, 
[…]

### 2016-02-18 model/overflowmcmcCwrap.R

@@ -1,0 +1,69 @@
+Sys.setenv("PKG_CXXFLAGS"="-std=c++11")
+## required so that armadillo uses c++11 for gamma random number generation
+library(MCMCpack)
+library(truncnorm) ## rtruncnorm() is faster than rtnorm() in the msm package
+library(ars)
+library(compiler)
+library(aster)
+library(Rcpp)
+library(RcppArmadillo)
+source("overflowmcmc.R")
+sourceCpp("overflowmcmcC.cpp")
+
+overtrendmcmcCwrap <- function(niter, data, initial, prior, rwsds){
+  ## Initial values
+  tau <- as.numeric(c(initial$tau))
+  nblocks <- length(tau)
+  sigma <- as.numeric(initial$sigma)
+  gam <- as.numeric(c(initial$gam))
+  phi <- as.numeric(initial$phi)
+  N <- as.numeric(c(initial$N))
+  Dbar <- as.numeric(c(initial$Dbar))
+  alpha <- as.numeric(initial$alpha)
+  beta <- as.numeric(initial$beta)
+  ## RW sds and bookkeeping
+  taulogrwsd <- as.numeric(rwsds$taulogrwsd)
+  gamlogrwsds <- as.numeric(rwsds$gamlogrwsds)
+  alphalogrwsd <- as.numeric(rwsds$alphalogrwsd)
+  rwc <- rwsds$rwc
+  H <- rwsds$H
+  tune <- (rwsds$tune)*1
+  lowtarget <- rwsds$lowtarget
+  hightarget <- rwsds$hightarget
+  tauchol <- rwsds$tauchol
+  ## Priors
+  vsigma <- prior$vsigma
+  ssigma <- prior$ssigma
+  aphi <- prior$aphi
+  bphi <- prior$bphi
+  mugam <- prior$mugamma
+  sig2gam <- prior$sig2gamma
+  Dbars <- prior$Dbars
+  aalpha <- prior$aalpha
+  balpha <- prior$balpha
+  abeta <- prior$abeta
+  bbeta <- prior$bbeta
+  ## Data
+  tobs <- data$tobs
+  m <- data$m
+  M <- data$M
+  D <- data$D
+  out <- overtrendmcmcCPP(niter, tobs, D, M, m, tau, sigma, gam, phi, N, Dbar, alpha, beta,
+                        vsigma, ssigma, aphi, bphi, mugam, sig2gam, Dbars, aalpha, balpha,
[…]

### 2016-02-18 model/overflowmcmcout.R

@@ -1,5 +1,5 @@
 source("overflowmcmc.R")
-library(MCMCpack)
+source("overflowmcmcCwrap.R")
 
 load("../charts/fullblocks.RData")
 cleanblocks <- blocks[,c("height", "time", "transactions_count", "size")]
@@ -13,6 +13,7 @@ block1k$ms <- 0
 for(i in 1:nblocks){
   block1k$ms[i + 11] <- median(block1k$tobs[i:(i + 10)])
 }
+
 
 datain <- list()
 datain$tobs <- block1k$tobs[11 + 1:nblocks] - block1k$tobs[10]
@@ -34,38 +35,39 @@ lfun <- function(a, m, d){
 ahat <- optimize(lfun, c(0.00001, 1000), m = diff(datain$M), d = datain$D, maximum = TRUE)$maximum
 bhat <- ahat/(sum(datain$D)/datain$M[nblocks + 1])
 
-
 ssigma <- 1
 vsigma <- 1
-aphi <- 1/50
-bphi <- 1/100
-agamma <- 1
-bgamma <- 1
-balpha <- 1/10
-aalpha <- ahat*balpha ## 1/10
-bbeta <- 1/10
-abeta <- bhat*bbeta
+aphi <- 2
+bphi <- 2
+mugamma <- 0
+sig2gamma <- 1
+balpha <- 1000000
+aalpha <- 1*balpha
+bbeta <- balpha
+abeta <- (bhat/ahat*1)*bbeta
 Dbars <- c(0.25, 0.35, 0.5, 0.75, 0.9, 0.95, 1)*1000
-priorin <- list(ssigma = ssigma, vsigma = vsigma, aphi = aphi, bphi = bphi, agamma = agamma,
-                bgamma = bgamma,
+
+Dbars <- c(1)*1000
+
+priorin <- list(ssigma = ssigma, vsigma = vsigma, aphi = aphi, bphi = bphi, mugamma = mugamma,
+                sig2gamma = sig2gamma,
                 aalpha = aalpha, balpha = balpha, abeta = abeta, bbeta = bbeta, Dbars = Dbars)
 
 initial <- list()
 initial$tau <- datain$tobs[order(datain$tobs)] + cumsum(rexp(nblocks, 100000))
 initial$sigma <- ssigma
-initial$gam <- 1
-initial$phi <- aphi/bphi
+initial$gam <- 0
[…]

### 2016-02-18 model/overflowmcmc.tex
Mostly updated the MCMC document to reflect current code - missing some details for the d's and for the algorithms for interesting sub-models

--- 2016-01-14 model/overflowmcmc.tex
@@
 \end{align*}
-Suppose the intensity function is a gamma process with shape function $\Psi(\tau)$ and rate parameter $\phi$. In other words
+Suppose the intensity function is a gamma process with shape function $\Psi(\tau)$ and rate parameter $\theta$. In other words
 \begin{align*}
-\Lambda(\tau_i + \delta_i) - \Lambda(\tau_i) \stackrel{ind}{\sim} G(\Psi(\tau_i + \delta_i) - \Psi(\tau_i), \phi)
+\Lambda(\tau_i + \delta_i) - \Lambda(\tau_i) \stackrel{ind}{\sim} G(\Psi(\tau_i + \delta_i) - \Psi(\tau_i), \theta)
 \end{align*}
-with mean $[\Psi(\tau_i + \delta_i) - \Psi(\tau_i)]/\phi$.
+with mean $[\Psi(\tau_i + \delta_i) - \Psi(\tau_i)]/\theta$.
 Then for block $k=1,2,\dots,n_{b}$ define $\lambda_k = \Lambda(\tau_k) - \Lambda(\tau_{k-1})$, $\eta_k = N(\tau_k) - N(\tau_{k-1})$, and $\psi_k = \Psi(\tau_k) - \Psi(\tau_{k-1})$.
-Then assume
+Then we assume
 \begin{align*}
 \eta_k | \lambda_{1:n_{b}} &\stackrel{ind}{\sim} Poi(\lambda_k)\\
-\lambda_k|\delta_{1:n_{b}} &\stackrel{ind}{\sim} G(\psi_k, \phi).
+\lambda_k|\delta_{1:n_{b}} &\stackrel{ind}{\sim} G(\psi_k, \theta).
 \end{align*}
 
-In order for the gamma process to be well defined we require $\phi>0$ and $\Psi(\tau)$ to be nondecreasing and right continuous with $\Psi(0)=0$.
+In order for the gamma process to be well defined we require $\theta>0$ and $\Psi(\tau)$ to be nondecreasing and right continuous with $\Psi(0)=0$.
 An easy way to achieve this is to define $\Psi(\tau) = \int_0^\tau\psi(t)dt$ where $\psi(t)$ is continuous and nonnegative.
@@
 E[N(\tau + \delta) - N(\tau)] &= E\{E[N(\tau + \delta) - N(\tau)|\Lambda(\tau + \delta) - \Lambda(\tau)]\} = E[\Lambda(\tau + \delta) - \Lambda(\tau)]\\
- &= [\Psi(\tau + \delta) - \Psi(\delta)]/\phi = \int_{\tau}^{\tau + \delta}\psi(t)dt/\phi.
+ &= [\Psi(\tau + \delta) - \Psi(\delta)]/\theta = \int_{\tau}^{\tau + \delta}\psi(t)dt/\theta.
 \end{align*}
-In this way, $\psi(t)/\phi$ can be seen as the expected intensity function of the $N(\tau)$ process.
+In this way, $\psi(t)/\theta$ can be seen as the expected intensity function of the $N(\tau)$ process.
 The gamma-Poisson structure also allows for a more flexible model on $N(\tau)$ by allowing for overdispersion.
@@
 &= E[\Lambda(\tau + \delta) - \Lambda(\tau)] + V[\Lambda(\tau + \delta) - \Lambda(\tau)]\\
-&= [\Psi(\tau + \delta) - \Psi(\tau)](1/\phi + 1/\phi^2) = \frac{\phi + 1}{\phi}E[N(\tau + \delta) - N(\tau)].
+&= [\Psi(\tau + \delta) - \Psi(\tau)](1/\theta + 1/\theta^2) = \frac{\theta + 1}{\theta}E[N(\tau + \delta) - N(\tau)].
 \end{align*}
-This allows for overdispersion but not underdispersion since $(\phi + 1)/\phi\in[1,\infty)$ for $\phi\ge 0$.
+This allows for overdispersion but not underdispersion since $(\theta + 1)/\theta\in[1,\infty)$ for $\theta\ge 0$.
+
+We can represent this part of the model more compactly by integrating out the $\lambda_k$'s. Define $\phi = \theta/(1 + \theta)$. Then $\eta_k$ has a negative binomial distribution:
+\begin{align*}
+\eta_k|\delta_{1:n_{b}} &\stackrel{ind}{\sim} NB(\psi_k, 1-\phi)
+\end{align*}
+with pmf
+\begin{align*}
+P(\eta_k = \eta|\delta_{1:n_{b}}) \frac{\Gamma(\psi_k + \eta)}{\Gamma(\eta + 1)}\phi^{\psi_k}(1-\phi)^{\eta}
+\end{align*}
+for $\eta=0,1,\dots$ We will use this representation of the model both for simplified exposition and later for cheaper MCMC.
 
 To complete this portion of the model we need to specify $\psi(t)$.
-A flexible structure that captures many modeling choices is $\psi(t) = \exp(\bm{z}(t)'\bm{\beta})$ where $\bm{z}(t)' = (z_1(t),\dots,z_p(t))$ is a vector of time-varying covariates continuously differentiable in time and $\bm{\beta}'=(\beta_1,\dots,\beta_p)$ is a vector of regression coefficients, for example $\psi(t) = \exp(\beta_1 + t\beta_2)$ in order to capture a simple trend.
+A flexible structure that captures many modeling choices is $\psi(t) = \exp(\bm{z}(t)'\bm{\gamma})$ where $\bm{z}(t)' = (z_1(t),\dots,z_p(t))$ is a vector of time-varying covariates continuously differentiable in time and $\bm{\gamma}'=(\gamma_1,\dots,\gamma_p)$ is a vector of regression coefficients, for example $\psi(t) = \exp(\gamma_1 + t\gamma_2)$ in order to capture a simple trend.
 In that case we have
 \begin{align*}
-\Psi(\tau+\delta) - \Psi(\tau) &= \int_{\tau}^{\tau + \delta} e^{\bm{z}(t)'\bm{\beta}}dt = \int_\tau^{\tau + \delta} e^{\beta_1 + t\beta_2}dt = \frac{e^{\beta_1 + \tau\beta_2}}{\beta_2}\left(e^{\delta\beta_2} - 1\right)
+\Psi(\tau+\delta) - \Psi(\tau) &= \int_{\tau}^{\tau + \delta} e^{\bm{z}(t)'\bm{\gamma}}dt = \int_\tau^{\tau + \delta} e^{\gamma_1 + t\gamma_2}dt = \frac{e^{\gamma_1 + \tau\gamma_2}}{\gamma_2}\left(e^{\delta\gamma_2} - 1\right)
 \intertext{and thus}
-\psi_k &= \frac{e^{\beta_1 + \tau_{k-1}\beta_2}}{\beta_2}\left(e^{\delta_k\beta_2} - 1\right).
+\psi_k &= \frac{e^{\gamma_1 + \tau_{k-1}\gamma_2}}{\gamma_2}\left(e^{\delta_k\gamma_2} - 1\right).
 \end{align*}
+
 \subsection{Transaction data}
@@
 \begin{align*}
-d_i \stackrel{iid}{\sim} G(\alpha_d,\beta_d)
+d_i \stackrel{iid}{\sim} G(\alpha,\beta)
 \end{align*}
-with mean $\alpha_d/\beta_d$.
+with mean $\alpha/\beta$.
 Along with the Poisson process for $N(\tau)$ this implies that the amount of transaction data that has arrived by time $\tau$, $D(\tau)$, is a compound non-homogenous Poisson process conditional on $\Lambda(\tau)$.
@@
 Denote this upper limit by $\bar{D}_k$ --- because different miners use different limits and the same miner uses different limits at different times, each block has its own upper limit.
-Then the number of transactions in blocks $1,2,\dots,k$ combined, $M_k$, could potentially be less than the number of transactions that have arrived by the time the $k$'th block arrives, $N_k$.
+Then the number of transactions in blocks $1,2,\dots,k$ combined could potentially be less than the number of transactions that have arrived by the time the $k$'th block arrives.
+Let $M_k$ denote the number of transactions total in blocks $1,2,\dots,k$.
+Then possibly $M_k \le N_k$.
 Specifically:
@@
 \section{Priors}
-The model has the following unknown parameters: $\sigma^2$, $\phi$, $\beta$, $\alpha_d$, $\beta_d$, and $\bar{D}_k$ for $k=1,2,\dots,n_b$.
+The model has the following unknown parameters: $\sigma^2$, $\phi$, $\beta$, $\alpha$, $\beta$, and $\bar{D}_k$ for $k=1,2,\dots,n_b$.
 We will suppose that these parameters are mutually independent in the prior, i.e.
@@
 \subsection{Prior on $\phi$}
-The conditionally conjugate prior for $\phi$ is a gamma distribution, so we will suppose that $\phi \sim G(a_{\phi}, b_{\phi})$.
-To choose $a_\phi$ and $b_\phi$ recall that $\phi$ is a dispersion parameter:
+The conditionally conjugate prior for $\phi$ is a beta distribution, so we will suppose that $\phi \sim B(a_{\phi}, b_{\phi})$.
+To choose $a_\phi$ and $b_\phi$ recall that $\phi$ is related to $\theta$, a dispersion parameter:
 \begin{align*}
-V[N(\tau)|\phi,\Psi(\tau)] = \frac{\phi + 1}{\phi}E[N(\tau)|\phi,\Psi(\tau)] = \frac{\phi + 1}{\phi}\frac{\Psi(\tau)}{\phi}.
+V[N(\tau)|\theta,\Psi(\tau)] = \frac{\theta + 1}{\theta}E[N(\tau)|\theta,\Psi(\tau)] = \frac{1}{\phi}E[N(\tau)|\theta,\Psi(\tau)] = \frac{1}{\phi}\Psi(\tau)\frac{1-\phi}{\phi}.
 \end{align*}
-Setting $\phi = 1$ means that the variance of $N(\tau)$ is twice its mean.
-Absent strong prior information about $\phi$, we will center our prior on $2$ with a high level of uncertainty.
-The mean of the gamma distribution is $a_\phi/b_\phi$ with variance $a_\phi/b_\phi^2$, so we will set $b_\phi = 1/100$ to reflect a high degree of uncertainty, and $a_\phi = 1/50$ so that $a_\phi/b_\phi = 2$.
+Setting $\phi = 0.5$ means that the variance of $N(\tau)$ is twice its mean.
+Absent strong prior information about $\phi$, we will center our prior on $2/3$ with a high level of uncertainty.
+The mean of the beta distribution is $a_\phi/(a_\phi + b_\phi)$ with variance $a_\phi b_\phi/[(a_\phi + b_\phi + 1)(a_\phi + b_\phi)^2]$, so we will set $a_\phi = 0.2$ and $b_\phi = 0.1$ to reflect a high degree of uncertainty.
 
-\subsection{Prior on $\bm{\beta}$}
-As we will see in the next section, there is no convenient conditionally conjugate form for $\bm{\beta}$, so there is no computational trade-off associated with choosing a prior which more accurately reflects prior knowledge.
-We will consider $\bm{\beta}$ much like a regression parameter and set
+\subsection{Prior on $\bm{\gamma}$}
+As we will see in the next section, there is no convenient conditionally conjugate form for $\bm{\gamma}$, so there is no computational trade-off associated with choosing a prior which more accurately reflects prior knowledge.
+We will consider $\bm{\gamma}$ much like a regression parameter and set
 \begin{align*}
-\bm{\beta} \sim N(\bm{b}_\beta, \bm{S}_\beta)
+\bm{\beta} \sim N(\bm{g}_\gamma, \bm{S}_\gamma)
 \end{align*}
-where $\bm{b}_\beta$ is a mean vector and $\bm{S}_\beta$ is a covariance matrix.
-We will use a weakly informative prior with $\bm{b}_\beta = \bm{0}$ and $\bm{S}_\beta = s^2_\beta\bm{I}$ where $s_\beta = 100$.
+where $\bm{g}_\gamma$ is a mean vector and $\bm{S}_\gamma$ is a covariance matrix.
+We will use a weakly informative prior with $\bm{g}_\gamma = \bm{0}$ and $\bm{S}_\gamma = s^2_\gamma\bm{I}$ where $s_\gamma = 100$.
 [MAYBE SOMETHING WITH FATTER TAILS HERE?
-MAYBE NOTE THAT NONINFORMATIVE PRIOR IS NOT A GOOD IDEA SINCE WE WANT TO GUARANTEE POSTERIOR PROPRIETY --- SAME FOR $s_\beta$ TOO LARGE (VIVEK'S PAPER)]
+MAYBE NOTE THAT NONINFORMATIVE PRIOR IS NOT A GOOD IDEA SINCE WE WANT TO GUARANTEE POSTERIOR PROPRIETY --- SAME FOR $s_\gamma$ TOO LARGE (VIVEK'S PAPER)]
 
-\subsection{Priors on $\alpha_d$ and $\beta_d$}
-The conditionally conjugate prior for $beta_d$ is a gamma while $\alpha_d$ has no convenient conditionally conjugate form.
-We will assume that $\beta_d \sim G(a_\beta, b_\beta)$ and $\alpha_d \sim G(a_\alpha, b_\alpha)$.
-In order to choose $a_\alpha$, $b_\alpha$, $a_\beta$, and $b_\beta$, consider that $E[d_i|\alpha_d,\beta_d] = \alpha_d/\beta_d$.
+\subsection{Priors on $\alpha$ and $\beta$}
+The conditionally conjugate prior for $beta$ is a gamma while $\alpha$ has no convenient conditionally conjugate form.
+We will assume that $\beta \sim G(a_\beta, b_\beta)$ and $\alpha \sim G(a_\alpha, b_\alpha)$.
+In order to choose $a_\alpha$, $b_\alpha$, $a_\beta$, and $b_\beta$, consider that $E[d_i|\alpha,\beta] = \alpha/\beta$.
 But
 \begin{align*}
-E\left[\frac{\alpha_d}{\beta_d}\right] = E[\alpha_d]E\left[\frac{1}{\beta_d}\right] = \frac{a_\alpha}{b_\alpha}\frac{b_\beta}{a_\beta - 1}.
+E\left[\frac{\alpha}{\beta}\right] = E[\alpha]E\left[\frac{1}{\beta}\right] = \frac{a_\alpha}{b_\alpha}\frac{b_\beta}{a_\beta - 1}.
 \end{align*}
@@
 \begin{align*}
-V\left[\frac{\alpha_d}{\beta_d}\right] = V[\alpha_d] + V\left[\frac{1}{\beta_d}\right] = \frac{a_\alpha}{b_\alpha^2} + \frac{b_\beta^2}{(a_\beta - 1)^2(a_\beta - 2)}.
+V\left[\frac{\alpha}{\beta}\right] = V[\alpha] + V\left[\frac{1}{\beta}\right] = \frac{a_\alpha}{b_\alpha^2} + \frac{b_\beta^2}{(a_\beta - 1)^2(a_\beta - 2)}.
 \end{align*}
-We will choose these parameters so that $E[\alpha_d/\beta_d] \approx 0.01$ MB and $V[\alpha_d/\beta_d]$ is large so that the prior is weakly informative.
+We will choose these parameters so that $E[\alpha/\beta] \approx 0.01$ MB and $V[\alpha/\beta]$ is large so that the prior is weakly informative.
 We will set $a_\beta = 2.1$ to ensure the variance calculated above exists.
@@
 \begin{align*}
-E\left[\frac{\alpha_d}{\beta_d}\right] = \frac{1}{21} \approx 0.09; && V\left[\frac{\alpha_d}{\beta_d}\right] \approx 10.
+E\left[\frac{\alpha}{\beta}\right] = \frac{1}{21} \approx 0.09; && V\left[\frac{\alpha}{\beta}\right] \approx 10.
 \end{align*}
@@
 Theoretically a miner can put any upper limit she wants on a mined bitcoin, but in practice most miners use one of several possible values.
-We will use the set of values $\mathcal{D} = \{0.25, 0.35, 0.5, 0.75, 0.9, 0.95, 1\}$ [DOUBLE CHECK THESE --- IN PARTICULAR, ARE THERE OTHER COMMON VALUES?
-WHAT ABOUT THE ELGIUS MINING POOL'S CAPS?].
+We will use the set of values $\mathcal{D} = \{0.25, 0.35, 0.5, 0.75, 0.935, 0.95, 1\}$.
 For each $\bar{D}_k$ we will use a discrete uniform prior on $\mathcal{D}$ so that $P(\bar{D}_k = j) = 1/7$ for all $j \in \mathcal{D}$.
@@
 \begin{align*}
-p(\bm{\tau}_{1:n_{b}}|...) \propto \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \tau_k\right)^2 - \frac{1}{10}\tau_{n_{b}}\right]\prod_{k=1}^{n_b}\frac{\phi^{\Psi(\tau_k) - \Psi(\tau_{k-1})}}{\Gamma\left[\Psi(\tau_k) - \Psi(\tau_{k-1})\right]}\lambda_k^{\Psi(\tau_k)-\Psi(\tau_{k-1}) - 1}
+p(\bm{\tau}_{1:n_{b}}|...) \propto & \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \tau_k\right)^2 - \frac{1}{10}\tau_{n_{b}}\right]\\
+&\times \prod_{k=1}^{n_b}\Gamma[\psi_k(\tau_{k}, \tau_{k-1}) + \eta_k]\phi^{\psi_k(\tau_k,\tau_{k-1})}1(0\le \tau_1 \le \tau_2 \le \hdots \le \tau_{n_b})
 \end{align*}
 where $\tau_0\equiv0$ and $\Gamma(x) = \int_0^\infty t^{x-1}e^{-t}dt$ is the gamma function.
+Here we make explicit that $\psi_k$ is a function of $\tau_k$ and $\tau_{k-1}$.
 Even if $\Psi(\tau)$ is some nice known form, this density is nonstandard because the $\tau_k$'s enter the gamma function.
@@
 
+Let $\delta_k = \tau_k - \tau_{k-1}$ for $k=1,2,\dots,n_b$. Then the determinant of the Jacobian is $1$ full conditional of $\delta_{1:n_b}$ is
+\begin{align*}
+p(\bm{\delta}_{1:n_{b}}|...) \propto & \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \sum_{l=1}^k\delta_l\right)^2 - \frac{1}{10}\sum_{k=1}^{n_b}\delta_{k}\right]\\
+&\times \prod_{k=1}^{n_b}\Gamma(\psi_k + \eta_k)\phi^{\psi_k}1(0\le \delta_k).
+\end{align*}
+Finally let $\zeta_k = \log(\delta_k)$ for $k = 1, 2, \dots, n_b$. Then the determinate of the Jacobian is $\prod_{k=1}^{n_b}e^{\zeta_k}$ and the full conditional of $\zeta_{1:n_b}$ is
+\begin{align*}
+p(\bm{\zeta}_{1:n_{b}}|...) \propto & \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \sum_{l=1}^ke^{\zeta_l}\right)^2 - \frac{1}{10}\sum_{k=1}^{n_b}e^{\zeta_k} + \sum_{k=1}^{n_b}{\zeta_k}\right] \prod_{k=1}^{n_b}\Gamma(\psi_k + \eta_k)\phi^{\psi_k}.
+\end{align*}
+
+For an intial sampler, we will use a random walk Metropolis algorithm on each coordinate with a normal proposal distribution $N(\zeta_k, h_k^2)$.
+Then the acceptance rate of the Metropolis step for $\zeta_k$ is
+\begin{align*}
+R_k = \frac{p(\zeta_k^{(prop)}|\dots)}{p(\zeta_k^{(old)}|\dots)} = \frac{p(\bm{\zeta}_{1:n_b}^{(prop)}|\dots)}{p(\zeta_{1:n_b}^{(old)}|\dots)}
+\end{align*}
+where $\zeta_k^{(old)}$ is the previous value in the chain, $\zeta_k^{(prop)}$ is the proposed value, $\bm{\zeta}_{1:n_b}^{(old)}$ is the previous value of the entire vector $\bm{\zeta}_{1:n_b}$, and $\bm{\zeta}_{1:n_b}^{(prop)}$ is the same except for $\zeta_k^{(prop)}$ in the $k$'th coordinate.
+So the $\zeta_k$ step is as follows: for $k=1,2,\dots,n_b$
+\begin{enumerate}
+\item Draw $\zeta_k^{(prop)} \sim N(\zeta_k^{(t)},h_k^2)$.
+\item Compute $\delta_{1:n_b}^{(prop)}$, $\tau_{1:n_b}^{(prop)}$, and $\psi_{1:n_b}^{(prop)}$.
+\item Compute the acceptance rate $R_k$.
+\item Set $\zeta_k^{(t+1)} = \zeta_k^{(prop)}$, $\delta_{1:n_b} = \delta_{1:n_b}^{(prop)}$, $\tau_{1:n_b} = \tau_{1:n_b}^{(prop)}$, and $\psi_{1:n_b} = \psi_{1:n_b}^{(prop)}$ with probability $\min(R_k, 1)$ and $\zeta_k^{(t+1)} = \zeta_k^{(t)}$ otherwise.
+\end{enumerate}
+For each $k$ the scale parameter $h_k$ must be chosen by the user, which is practically impossible with a large number of blocks in the analysis.
+So we automatically tune $h_k$ during the the burn in period as follows [CITATION]:
+\begin{enumerate}
+\item Run the algorithm above for $H$ iterations.
+\item For each $k$, compute the observed acceptance rate of the last $H$ iterations: $r_k\#(\zeta_k^{(t)} = \zeta_k^{(t-1)})/H$.
+\item If $r_k$ is larger than some target range $[r_l^*, r_h^*]$ update $\log h_k = \log h_k + c$; if lower update $\log h_k = \log h_k - c$.
+\end{enumerate}
+In certain simple cases the optimal acceptance rate for a one-dimensional target has been computed to be approximately $0.44$, so we use the range $[0.4, 0.45]$ as a rough guideline [CITATION].
+In order to ensure that the Markov chain converges to the target posterior of distribution this adaptive process must asymptotically stop in some sense --- restricting the process to the burn in is a sufficient condition.
+In practice as the chain converges in can help to start $H$ small and $c$ large so that the $h_k$'s quickly move to reasonable values, then over time increase $H$ and decrease $c$ for more precise fine tuning.
+We start with $H=50$ and $c=0.1$
+
+The previous algorithm for the $\zeta_k$ steps works but will suffer from slow convergence because the $\zeta_k$'s are being updated separately instead of in a block.
+It is helpful for getting a crude estimate of the posterior covariance matrix for $\zeta_{1:n_b}$, which we will denote by $\hat{\Sigma}$.
+Using this we can implement a blockwise Metropolis update for $\zeta_{1:n_b}$, which will substantially decrease the number of draws required to adequately characterize the posterior distribution after burn in.
+Then the Metropolis proposal is drawn from a $N(\zeta_{1:n_b}^{(t)}, h^2\hat{\Sigma})$ distribution where $\zeta_{1:n_b}^{(t)}$ is the previous value of $\zeta_{1:n_b}$ in the chain, $h$ is a user chosen scale parameter, and $\hat{\Sigma}$ is a crude estimate of the posterior covariance matrix for $\zeta_{1:n_b}$ obtained using the single dimensional random walk Metropolis algorithm used above.
+The Metropolis acceptance ratio is then $p(\bm{\zeta}_{1:n_b}^{(prop)}|...)/p(\bm{\zeta}_{1:n_b}^{(old)}|...)$.
+Then $h$ can be tuned just like the $h_k$'s with a target rate of about $0.24$ that is more appropriate for the high dimensional case.
+
 \subsection{Transaction process dispersion --- $\phi$}
@@
 \begin{align*}
-p(\phi|...) \propto \prod_{k=1}^{n_{b}}\phi^{\psi_k}e^{\lambda_k\phi}p(\phi) \propto \phi^{\sum_{k=1}^{n_{b}}\psi_k}e^{\phi\sum_{k=1}^{n_{b}}\lambda_k}p(\phi).
+p(\phi|...) \propto \phi^{a_\phi + \sum_{k=1}^{n_b}\psi_k}(1-\phi)^{b_\phi + \sum_{k=1}^{n_b}\eta_k}
 \end{align*}
-With $\phi \sim G(a_\phi, b_\phi)$ in the prior, this is a gamma density so the step is simple:
+which is a beta distribution. So this step is simple:
 \begin{enumerate}
-\item Draw $\phi \sim G\left(a_\phi + \sum_{k=1}^{n_{b}}\psi_k, b_{\phi} + \sum_{k=1}^{n_{b}}\lambda_k \right)$.
+\item Draw $\phi \sim B\left(a_\phi + \sum_{k=1}^{n_{b}}\psi_k, b_{\phi} + \sum_{k=1}^{n_{b}}\eta_k \right)$.
 \end{enumerate}
 
-\subsection{Regression parameters --- $\bm{\beta}$}
-The full conditional posterior of $\bm{\beta}$ is closely connected to the functional form of the covariates in time.
+\subsection{Regression parameters --- $\bm{\gamma}$}
+The full conditional posterior of $\bm{\gamma}$ is closely connected to the functional form of the covariates in time.
 No matter the functional form, however, the density is complex and a Metropolis step is likely warranted.
-Recall that since we defined $\psi(t) = e^{\bm{z}(t)'\bm{\beta}}$ we have $\psi_k(\bm{\beta}) = \int_{\tau_{k-1}}^{\tau^k}e^{\bm{z}(t)'\bm{\beta}}dt$ --- in this subsection we write $\psi_k(\bm{\beta})$ to explicitly acknowledge that $\psi_k$ is a function of $\bm{\beta}$.
-Then the full conditional of $\bm{\beta}$ is
+Recall that since we defined $\psi(t) = e^{\bm{z}(t)'\bm{\gamma}}$ we have $\psi_k(\bm{\gamma}) = \int_{\tau_{k-1}}^{\tau^k}e^{\bm{z}(t)'\bm{\gamma}}dt$ --- in this subsection we write $\psi_k(\bm{\gamma})$ to explicitly acknowledge that $\psi_k$ is a function of $\bm{\gamma}$.
+Then the full conditional of $\bm{\gamma}$ is
 \begin{align*}
-p(\bm{\beta}|\dots) \propto p(\bm{\beta})\prod_{k=1}^{n_b}\frac{\phi^{\psi_k(\bm{\beta})}\lambda_k^{\psi_k(\bm{\beta}) - 1}}{\Gamma[\psi_k(\bm{\beta})]}
+p(\bm{\gamma}|\dots) \propto p(\bm{\gamma})\prod_{k=1}^{n_b}\phi^{\psi_k(\bm{\gamma})}\Gamma[\psi_k(\bm{\gamma}) + \eta_k]
 \end{align*}
-where in the prior $\bm{\beta} \sim p(\bm{\beta})$ independent of the other parameters.
-This density is complex and nonstandard since $\psi_k(\bm{\beta})$ enters the gamma function.
-A Metropolis step is likely warranted here as well.
+where in the prior $\bm{\gamma} \sim p(\bm{\gamma})$ independent of the other parameters.
+This density is complex and nonstandard since $\psi_k(\bm{\gamma})$ enters the gamma function.
+In the case of $\bm{\gamma} = (\gamma_1, \gamma_2)'$ and $\bm{z}(t) = (1, t)'$ we use separate random walk Metropolis steps with a normal proposal and separately tuned standard deviations --- analogous to the algorithm used for the $\zeta_k$'s above.
 
@@
 \begin{align*}
-p(\bm{d}_{1:n_{b}}|...)\propto \prod_{k=1}^{n_b}\prod_{i=1}^{M_k - M_{k-1}} d_i^{\alpha_d - 1}e^{-d_i\beta_d}1\left(\sum_{i=1}^{M_k - M_{k-1}} d_{ik}= D_k\right)1\left(d_{1k} > \bar{D}_{k-1} - D_{k-1} \mathrm{\ or\ } M_{k-1} = N_{k-1}\right).
+p(\bm{d}_{1:n_{b}}|...)\propto \prod_{k=1}^{n_b}\prod_{i=1}^{M_k - M_{k-1}} d_i^{\alpha - 1}e^{-d_i\beta}1\left(\sum_{i=1}^{M_k - M_{k-1}} d_{ik}= D_k\right)1\left(d_{1k} > \bar{D}_{k-1} - D_{k-1} \mathrm{\ or\ } M_{k-1} = N_{k-1}\right).
 \end{align*}
-The second indicator function is required because if $M_{k-1} < N_{k-1}$ then $N_{k-1} - M_{k-1}$ transactionss must have overflowed into block $k$.
+The second indicator function is required because if $M_{k-1} < N_{k-1}$ then $N_{k-1} - M_{k-1}$ transactions must have overflowed into block $k$.
 So the amount of data from the first transaction of block $k$, $d_{1k}$, must have been greater than the difference between the amount of data in block $k-1$ and the data cap for block $k-1$.
-Similarly if $M_{k-1} = N_{k-1}$ then we know there was no overflow into block $k$, so we have no reason to think that $d_{1k}$ has a restricted support.
+Similarly if $M_{k-1} = N_{k-1}$ then we know there was no overflow into block $k$, so $d_{1k}$ does not have restricted support.
 Let $\bm{o}_k = \bm{d}_k/D_k$.
@@
 \begin{align*}
-p(\bm{o}_{1:n_b}|...)\propto \prod_{k=1}^{n_b} \prod_{i=1}^{M_k - M_{k-1}}o_{ik}^{\alpha_d - 1} 1\left(d_{1k} > [\bar{D}_{k-1} - D_{k-1}]/D_{k} \mathrm{\ or\ } M_{k-1} = N_{k-1}\right).
+p(\bm{o}_{1:n_b}|...)\propto \prod_{k=1}^{n_b} \prod_{i=1}^{M_k - M_{k-1}}o_{ik}^{\alpha - 1} 1\left(d_{1k} > [\bar{D}_{k-1} - D_{k-1}]/D_{k} \mathrm{\ or\ } M_{k-1} = N_{k-1}\right).
 \end{align*}
@@
 \begin{enumerate}
-\item If $N_k = M_k$ draw $\bm{o}_{k} \sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1}})$.
+\item If $N_k = M_k$ draw $\bm{o}_{k} \sim Dir(\alpha\bm{1}_{M_k - M_{k-1}})$.
 
-  Otherwise draw $\bm{o}_{k} \sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1}})1(d_{1k} > [\bar{D}_{k-1} - D_{k-1}]/D_k)$.
+  Otherwise draw $\bm{o}_{k} \sim Dir(\alpha\bm{1}_{M_k - M_{k-1}})1(d_{1k} > [\bar{D}_{k-1} - D_{k-1}]/D_k)$.
 \item Set $\bm{d}_k = \bm{o}_kD_k$.
 \end{enumerate}
-To draw from the truncated Dirichlet distribution above, first note that by the properties of the Dirhichlet the marginal distribution of $o_{1k}$ is $Beta(\alpha_d, \alpha_d(M_k - M_{k-1} - 1))$ and conditional on $o_{1k}$, $\bm{o}_{-1,k}/(1 - o_{1k}) \sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1} - 1})$.
+To draw from the truncated Dirichlet distribution above, first note that by the properties of the Dirhichlet the marginal distribution of $o_{1k}$ is $Beta(\alpha, \alpha(M_k - M_{k-1} - 1))$ and conditional on $o_{1k}$, $\bm{o}_{-1,k}/(1 - o_{1k}) \sim Dir(\alpha\bm{1}_{M_k - M_{k-1} - 1})$.
 Then $o_{1k}$ can be drawn from the truncated Beta distribution using the inverse cdf method, resulting in the following two step algorithm to draw from the desired truncated Dirichlet distribution:
 \begin{enumerate}
-\item Draw $u \sim U(0,1)$ and set $o_{1k} = F^{-1}(u*[1 - F((\bar{D}_{k-1} - D_{k-1})/D_k)])$ where $F$ is the cdf of the $Beta(\alpha_d, \alpha_d(M_k - M_{k-1} - 1))$ distribution.
-\item Draw $\tilde{\bm{o}}_{2:(M_k - M_{k-1})}\sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1} - 1})$ and set $o_{ik} = \tilde{o}_i$ for $i = 2,3,\dots,M_k - M_{k-1}$.
+\item Draw $u \sim U(0,1)$ and set $o_{1k} = F^{-1}(u*[1 - F((\bar{D}_{k-1} - D_{k-1})/D_k)])$ where $F$ is the cdf of the $Beta(\alpha, \alpha(M_k - M_{k-1} - 1))$ distribution.
+\item Draw $\tilde{\bm{o}}_{2:(M_k - M_{k-1})}\sim Dir(\alpha\bm{1}_{M_k - M_{k-1} - 1})$ and set $o_{ik} = \tilde{o}_i$ for $i = 2,3,\dots,M_k - M_{k-1}$.
 \end{enumerate}
+
+Note that to draw from a Dirichlet distribution, $Dir(a_1, a_2, \dots, a_n)$, you simply draw $g_i\stackrel{ind}{\sim} G(a_i,b)$ for any $b>0$, then set $o_i = g_i/\sum_{i=1}^ng_i$.
+Then to avoid drawing $g_i$'s that are numerically identical to zero --- which occurs ocassionally if $a_i < 1$ --- note that if $x\sim G(a+1,b)$ and $u\sim U(0,1)$ independently, then $y = u^{1/a}x \sim G(a,b)$.
+[CITATION].
+
+[ADD DETAILS ABOUT TRUNCATED DRAWS, DEALING WITH NUMERICAL ISSUES, AND WHAT TO DO WHEN FOR $N_{n_b}$]
 
@@
 If the summation is greater than $D_k$ then it is still possible that $N_k = M_k$ because transaction $d_{M_k + 1}$ could still have arrived later than block $k$.
-In that case, the density of $N_k$ given $N_{0:(k-1)}$ and all other parameters and processes is given by
+In that case, the pmf of $N_k$ given $N_{-k}$ and all other parameters and processes is given by
 \begin{align*}
-p(N_k|N_{0:(k-1)}, \dots) \propto \frac{\lambda_k^{N_k - N_{k-1}}e^{-\lambda_k}}{(N_k - N_{k-1})!}1(N_k \ge M_k)1(N_k \ge N_{k-1}).
+p(N_k|\bm{N}_{-k}, \dots) \propto \frac{\Gamma(\psi_k + N_k - N_{k-1})}{\Gamma(N_k - N_{k-1} + 1)} \frac{\Gamma(\psi_{k+1} + N_{k+1} - N_{k})}{\Gamma(N_{k+1} - N_{k} + 1)}1(N_{k-1}\le N_k \le N_{k+1})1(M_k\le N_k)
 \end{align*}
-In other words, $\eta_k = N_k - N_{k-1}$ has a truncated Poisson distribution where the support is restricted to $\{M_k - N_{k-1}, M_k - N_{k-1} + 1, \dots \}$.
-We can easily simulate from this distribution using the inverse cdf method.
-So we can draw from $p(\bm{N}_{1:n_b}|\dots)$ as follows: For $k=1,2,\dots,n_b$
+In other words the full condtional of $\eta_k = N_k - N_{k-1}$ given $N_{-k}$ is
+\begin{align*}
+p(\eta_k|\bm{N}_{-k}, \dots) \propto \frac{\Gamma(\psi_k + \eta_k)}{\Gamma(\eta_k + 1)} \frac{\Gamma(\psi_{k+1} + N_{k+1} - N_{k-1} -\eta_k)}{\Gamma(N_{k+1} - N_{k-1} - \eta_k + 1)}1(0\le \eta_k \le N_{k+1} - N_{k-1})1(M_k - N_{k-1}\le \eta_k).
+\end{align*}
+This is a discrete distribution on a finite set and can easily be drawn by computing the probability of each element of the set, then sampling one from the set with the computed probabilities.
+
+For $N_{n_b}$ the pmf is slightly different:
+\begin{align*}
+p(\eta_{n_b}|\bm{N}_{-{n_b}}, \dots) \propto \frac{\Gamma(\psi_{n_b} + \eta_{n_b})}{\Gamma(\eta_{n_b} + 1)} \phi^{\psi_{n_b}}(1-\phi)^{\eta_{n_b}}1(0\le \eta_{n_b})1(M_{n_b} - N_{n_b-1}\le \eta_{n_b}).
+\end{align*}
+This is a (potentially) truncated negative binomial distribution, and can be drawn from using the algorithm in [CITATION].
+So we can easily simulate each of $N_k$ given $N_{-k}$ as follows: For $k=1,2,\dots,n_b-1$
 \begin{enumerate}
@@
 
-Otherwise, draw $\eta_k \sim Poi(\lambda_k)1(\eta_k \ge M_k - N_{k-1})$ and set $N_k = N_{k-1} + \eta_k$.
+Otherwise, compute $p_\eta$ for each $\eta \in A = \{ \max(0, M_k - N_{k-1}), \max(0, M_k - N_{k-1}) + 1, \dots, N_{k+1} - N_{k-1}\}$ and draw $\eta_k\sim (A,\bm{p}_\eta)$ and set $N_k = N_{k-1} + \eta_k$.
 \end{enumerate}
+Then for $N_{n_b}$:
+\begin{enumerate}
+\item If $\sum_{i=1}^{M_{n_b} - M_{n_b-1} + 1} d_{M_{n_b} + i} \le \bar{D}_{n_b}$ set $N_{n_b} = M_{n_b}$.
+
+Otherwise draw $\eta_{n_b}\sim NB(\psi_{n_b}, 1-\phi)1(\eta_{n_b}\ge \max(0, M_{n_b} - N_{n_b-1}))$ and set $N_{n_b} = \eta_{n_b} + N_{n_b - 1}$.
+\end{enumerate}
+Note that the intermediate values of the $\eta_k$ generated in these steps are NOT the correct values of the $\eta_k$'s after each step is completed.
+
 \subsection{Block size caps --- $\bar{D}_k$}
@@
 
-\subsection{Transaction data parameters --- $\alpha_d$ and $\beta_d$}
-The full conditional distribution of $(\alpha_d,\beta_d)$ is 
+\subsection{Transaction data parameters --- $\alpha$ and $\beta$}
+The full conditional distribution of $(\alpha,\beta)$ is 
 \begin{align*}
-p(\alpha_d,\beta_d|...) \propto p(\alpha_d)p(\beta_d) \prod_{i=1}^{n_d}\frac{\beta_d^{\alpha_d}}{\Gamma(\alpha_d)}d_i^{\alpha_d - 1}e^{-\beta_dd_i}.
+p(\alpha,\beta|...) \propto p(\alpha)p(\beta) \prod_{i=1}^{n}\frac{\beta^{\alpha}}{\Gamma(\alpha)}d_i^{\alpha - 1}e^{-\beta d_i}.
 \end{align*}
-When $\beta_d \sim G(a_\beta, b_\beta)$ in the prior, its full conditional becomes
+When $\beta \sim G(a_\beta, b_\beta)$ in the prior, its full conditional becomes
 \begin{align*}
-p(\beta_d|...) \propto \beta_d^{\alpha_dn_d + a_{\beta} - 1}e^{-\beta_d\left(\sum_{i=1}^{n_d}d_i + b_{\beta}\right)},
+p(\beta|...) \propto \beta^{\alpha n_d + a_{\beta} - 1}e^{-\beta\left(\sum_{i=1}^{n_d}d_i + b_{\beta}\right)},
 \end{align*}
-which is another gamma distribution. The full conditional of $\alpha_d$ is complex no matter what its prior is since $\alpha_d$ enters the gamma function, and has the form
+which is another gamma distribution. The full conditional of $\alpha$ is complex no matter what its prior is since $\alpha$ enters the gamma function, and has the form
 \begin{align*}
-p(\alpha_d|...) &\propto p(\alpha_d)\frac{\left(\beta_d^{n_d}\prod_{i=1}^{n_d}d_i\right)^{\alpha_d}}{\Gamma(\alpha_d)^{n_d}}\\
-\intertext{which in the case of a gamma prior on $\alpha_d$ is}
-p(\alpha_d|...) &\propto \alpha_d^{a_\alpha - 1}\frac{\left(e^{-b_\alpha}\beta_d^{n_d}\prod_{i=1}^{n_d}d_i\right)^{\alpha_d}}{\Gamma(\alpha_d)^{n_d}}.
+p(\alpha|...) &\propto p(\alpha)\frac{\left(\beta^{n_d}\prod_{i=1}^{n_d}d_i\right)^{\alpha}}{\Gamma(\alpha)^{n_d}}\\
+\intertext{which in the case of a gamma prior on $\alpha$ is}
+p(\alpha|...) &\propto \alpha^{a_\alpha - 1}\frac{\left(e^{-b_\alpha}\beta^{n_d}\prod_{i=1}^{n_d}d_i\right)^{\alpha}}{\Gamma(\alpha)^{n_d}}.
 \end{align*}
-This density can be drawn from using rejection sampling or an independence Metropolis step, e.g. ARS or ARMS.
+This density is log concave when $a_\alpha \ge 0$ but not in general, so we will use a random walk Metropolis step. First full conditional of $\xi = \log(\alpha)$ is
+\begin{align*}
+p(\xi|...) &\propto e^{a_\alpha\xi}\frac{\left(e^{-b_\alpha}\beta^{n_d}\prod_{i=1}^{n_d}d_i\right)^{e^{\xi}}}{\Gamma(e^{\xi})^{n_d}}.
+\end{align*}
+Then the Metropolis proposal is $\xi^{(prop)}\sim N(\xi^{(t)},h_{\xi}^2)$, the acceptance ratio is $R_{\xi} = p(\xi^{(prop)}|...)/p(\xi^{(t)}|...)$, and $h_{\xi}$ is a user defined parameter that can be tuned analogously to the previous random walk Metropolis steps.
 \bibliographystyle{apalike} 

### 2017-01-17 model/overflowmcmc.R
changed the rw update to more modern technology

@@ -4,6 +4,499 @@ library(ars)
 library(compiler)
 library(aster)
 
+## overflow compound gamma testing
+overcgammamcmc <- function(niter, data, initial, prior, rwsds){
+  ## Initial values
+  print("initial")
+  tau <- initial$tau
+  nblocks <- length(tau)
+  tau0 <- c(0, tau[-nblocks])
+  delta <- diff(c(0, tau))
+  sigma <- initial$sigma
+  gam <- initial$gam
+  phi <- initial$phi
+  N <- initial$N
+  eta <- diff(N)
+  Ntotal <- N[nblocks + 1]
+  Dbar <- initial$Dbar
+  alpha <- initial$alpha
+  beta <- initial$beta
+  ps <- initial$ps
+  q <- initial$q
+  psi <- psifunC(tau, c(gam,0), tau0)
+  print("rwsds")
+  ## RW sds and bookkeeping
+  taulogrwsd <- rwsds$taulogrwsd
+  gamlogrwsd <- rwsds$gamlogrwsd
+  alphalogrwsd <- rwsds$alphalogrwsd
+  plogrwsds <- rwsds$plogrwsds  
+  taurwsd <- exp(taulogrwsd)
+  gamrwsd <- exp(gamlogrwsd)
+  alpharwsd <- exp(alphalogrwsd)
+  prwsds <- exp(plogrwsds)  
+  tauaccs <- matrix(0, ncol = 1, nrow = niter)
+  gamaccs <- matrix(0, ncol = 1, nrow = niter)
+  alphaaccs <- matrix(0, ncol = 1, nrow = niter)
+  print("probably expensive")
+  paccs <- matrix(0, ncol = length(initial$ps), nrow = niter)
+  print("expensive done")
+  rwc <- rwsds$rwc
+  tune <- rwsds$tune
+  accprob <- rwsds$accprob
+  tauchol <- rwsds$tauchol
+  print("priors")
+  ## Priors
+  vsigma <- prior$vsigma
+  ssigma <- prior$ssigma
+  aphi <- prior$aphi
+  bphi <- prior$bphi
+  mugam <- prior$mugamma
+  sig2gam <- prior$sig2gamma
+  agam <- prior$agamma
+  bgam <- prior$bgamma
+  Dbars <- prior$Dbars
+  aalpha <- prior$aalpha
+  balpha <- prior$balpha
+  abeta <- prior$abeta
+  bbeta <- prior$bbeta
+  aq <- prior$aq
+  bq <- prior$bq
[…]

### 2017-01-17 model/overflowmcmcout.R

@@ -32,26 +32,47 @@ lfun <- function(a, m, d){
   return(out)
 }
 
+lcgamma <- function(par, dat){
+  k <- exp(par[1])
+  alpha <- exp(par[2])
+  beta <- exp(par[3])
+  m <- dat[,1]
+  d <- dat[,2]
+  a <- m*k
+  out <- - lbeta(alpha, a) - a*log(beta) + (a-1)*log(d) - (alpha + a)*log(1 + d/beta)
+  return(-sum(out))
+}
+
+
 ahat <- optimize(lfun, c(0.00001, 1000), m = diff(datain$M), d = datain$D, maximum = TRUE)$maximum
 bhat <- ahat/(sum(datain$D)/datain$M[nblocks + 1])
+
+test <- optim(c(0,0,0), lcgamma, dat=cbind(diff(datain$M), datain$D))
+testpars <- exp(test$par)
+mmm <- tail(diff(datain$M), 1)
+nnn <- 10000
+
+outsim <- matrix(rgamma(nnn*mmm, rgamma(nnn*mmm, 1, 1.9), .07), nrow = mmm)
+outsim <- matrix(rgamma(nnn*mmm, .04, .07), nrow = mmm)
+outsim2 <- apply(outsim, 2, sum)
 
 ssigma <- 1
 vsigma <- 1
 aphi <- 2
 bphi <- 2
-mugamma <- 0
+mugamma <- 3.5
 sig2gamma <- 1
-balpha <- 1000000
+balpha <- 100 ##0000
 aalpha <- 1*balpha
 bbeta <- balpha
 abeta <- (bhat/ahat*1)*bbeta
 Dbars <- c(0.25, 0.35, 0.5, 0.75, 0.9, 0.95, 1)*1000
 
-Dbars <- c(1)*1000
+##Dbars <- c(1)*1000
 
 priorin <- list(ssigma = ssigma, vsigma = vsigma, aphi = aphi, bphi = bphi, mugamma = mugamma,
-                sig2gamma = sig2gamma,
-                aalpha = aalpha, balpha = balpha, abeta = abeta, bbeta = bbeta, Dbars = Dbars)
+                sig2gamma = sig2gamma, aalpha = aalpha, balpha = balpha,
+                abeta = abeta, bbeta = bbeta, aq = aalpha, bq = balpha, Dbars = Dbars)
 
 initial <- list()
 initial$tau <- datain$tobs[order(datain$tobs)] + cumsum(rexp(nblocks, 100000))
[…]

### 2017-01-17 paper/paper.tex
Added new Gamma table, old one commented out; also deleted an extraneous line from the model fitting script & removed old data file
Updated gamma fits after fixing the sql query issue
Updated the tables based on fitting the gamma process model to the end of the block chain
mode changed on a bunch of files

@@
 
-% latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Mon Oct 12 10:44:18 2015
+% latex table generated in R 3.2.3 by xtable 1.7-4 package
+% Wed Dec 23 10:33:41 2015
 \begin{table}[ht]
 \centering
-\begin{tabular}{lrrr}
+\begin{tabular}{rrrr}
   \hline
@@
   \hline
-       $n=5000$ & 1.34 & 1.29 & 1.39 \\ 
-$n=1000$, $g=1$ & 1.26 & 1.17 & 1.37 \\ 
-$n=1000$, $g=2$ & 1.36 & 1.26 & 1.47 \\ 
-$n=1000$, $g=3$ & 1.37 & 1.26 & 1.47 \\ 
-$n=1000$, $g=4$ & 1.27 & 1.17 & 1.37 \\ 
-$n=1000$, $g=5$ & 1.41 & 1.30 & 1.52 \\ 
- $n=500$, $g=1$ & 1.29 & 1.15 & 1.45 \\ 
- $n=500$, $g=2$ & 1.23 & 1.10 & 1.37 \\ 
- $n=500$, $g=3$ & 1.46 & 1.31 & 1.63 \\ 
- $n=500$, $g=4$ & 1.28 & 1.14 & 1.42 \\ 
- $n=500$, $g=5$ & 1.55 & 1.38 & 1.74 \\ 
- $n=500$, $g=6$ & 1.21 & 1.08 & 1.34 \\ 
- $n=500$, $g=7$ & 1.24 & 1.11 & 1.38 \\ 
- $n=500$, $g=8$ & 1.29 & 1.15 & 1.44 \\ 
- $n=500$, $g=9$ & 1.36 & 1.21 & 1.52 \\ 
- $n=500$, $g=10$ & 1.46 & 1.30 & 1.63 \\ 
+  $n=5000$ & 1.04 & 1.01 & 1.08 \\ 
+  $n=1000$, $g=1$ & 0.93 & 0.86 & 1.00 \\ 
+  $n=1000$, $g=2$ & 0.93 & 0.86 & 1.00 \\ 
+  $n=1000$, $g=3$ & 1.00 & 0.92 & 1.08 \\ 
+  $n=1000$, $g=4$ & 1.04 & 0.96 & 1.12 \\ 
+  $n=1000$, $g=5$ & 1.52 & 1.41 & 1.64 \\ 
+  $n=500$, $g=1$ & 0.89 & 0.80 & 0.99 \\ 
+  $n=500$, $g=2$ & 0.97 & 0.86 & 1.08 \\ 
+  $n=500$, $g=3$ & 0.92 & 0.83 & 1.02 \\ 
+  $n=500$, $g=4$ & 0.94 & 0.84 & 1.04 \\ 
+  $n=500$, $g=5$ & 1.26 & 1.12 & 1.40 \\ 
+  $n=500$, $g=6$ & 0.83 & 0.74 & 0.92 \\ 
+  $n=500$, $g=7$ & 1.04 & 0.93 & 1.16 \\ 
+  $n=500$, $g=8$ & 1.04 & 0.92 & 1.15 \\ 
+  $n=500$, $g=9$ & 1.40 & 1.25 & 1.57 \\ 
+  $n=500$, $g=10$ & 1.67 & 1.49 & 1.86 \\ 
    \hline
@@
 \end{table}
+
+% latex table generated in R 3.2.3 by xtable 1.7-4 package
+% Wed Dec 30 12:33:51 2015
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrr}
+  \hline
+ & Mean & 2.5\% & 97.5\% \\ 
+  \hline
+  $n=5000$ & 1.04 & 1.00 & 1.07 \\ 
+  $n=1000$, $g=1$ & 1.16 & 1.08 & 1.25 \\ 
+  $n=1000$, $g=2$ & 1.01 & 0.93 & 1.09 \\ 
+  $n=1000$, $g=3$ & 1.10 & 1.01 & 1.18 \\ 
+  $n=1000$, $g=4$ & 1.00 & 0.92 & 1.08 \\ 
+  $n=1000$, $g=5$ & 0.96 & 0.89 & 1.04 \\ 
+  $n=500$,  $g=1$ & 1.24 & 1.11 & 1.38 \\ 
+  $n=500$,  $g=2$ & 1.09 & 0.98 & 1.21 \\ 
+  $n=500$,  $g=3$ & 0.98 & 0.88 & 1.09 \\ 
+  $n=500$,  $g=4$ & 1.04 & 0.93 & 1.16 \\ 
+  $n=500$,  $g=5$ & 1.27 & 1.13 & 1.41 \\ 
+  $n=500$,  $g=6$ & 0.98 & 0.88 & 1.09 \\ 
+  $n=500$,  $g=7$ & 0.92 & 0.83 & 1.02 \\ 
+  $n=500$,  $g=8$ & 1.09 & 0.97 & 1.21 \\ 
+  $n=500$,  $g=9$ & 1.12 & 1.00 & 1.24 \\ 
+  $n=500$,  $g=10$ & 0.86 & 0.77 & 0.95 \\ 
+   \hline
+\end{tabular}
+\caption{Posterior mean and 95\% credible intervals for $\alpha$ from fitting a gamma distribution to block sizes of the last 5000 blocks in our database.
+For sample sizes less than 5000, blocks were broken into groups of equal size and the model was fit separately to each group.
+Sample size is indicated by $n$, group id by $g$.
+Each model was fit in RStan \citep{rstan-software:2015,stan-software:2015} using 4 chains, each with 4000 iterations, 2000 of which used for warmup/burn in.
+PRE 2015}
+\end{table}
+
+% OLD TABLE
+% latex table generated in R 3.2.2 by xtable 1.7-4 package
+% Mon Oct 12 10:44:18 2015
+%% \begin{table}[ht]
+%% \centering
+%% \begin{tabular}{lrrr}
+%%   \hline
+%%  & Mean & 2.5\% & 97.5\% \\ 
+%%   \hline
+%%        $n=5000$ & 1.34 & 1.29 & 1.39 \\ 
+%% $n=1000$, $g=1$ & 1.26 & 1.17 & 1.37 \\ 
+%% $n=1000$, $g=2$ & 1.36 & 1.26 & 1.47 \\ 
+%% $n=1000$, $g=3$ & 1.37 & 1.26 & 1.47 \\ 
+%% $n=1000$, $g=4$ & 1.27 & 1.17 & 1.37 \\ 
+%% $n=1000$, $g=5$ & 1.41 & 1.30 & 1.52 \\ 
+%%  $n=500$, $g=1$ & 1.29 & 1.15 & 1.45 \\ 
+%%  $n=500$, $g=2$ & 1.23 & 1.10 & 1.37 \\ 
+%%  $n=500$, $g=3$ & 1.46 & 1.31 & 1.63 \\ 
+%%  $n=500$, $g=4$ & 1.28 & 1.14 & 1.42 \\ 
+%%  $n=500$, $g=5$ & 1.55 & 1.38 & 1.74 \\ 
+%%  $n=500$, $g=6$ & 1.21 & 1.08 & 1.34 \\ 
+%%  $n=500$, $g=7$ & 1.24 & 1.11 & 1.38 \\ 
+%%  $n=500$, $g=8$ & 1.29 & 1.15 & 1.44 \\ 
+%%  $n=500$, $g=9$ & 1.36 & 1.21 & 1.52 \\ 
+%%  $n=500$, $g=10$ & 1.46 & 1.30 & 1.63 \\ 
+%%  \hline
+%% \end{tabular}
+%% \caption{Posterior mean and 95\% credible intervals for $\alpha$ from fitting a gamma distribution to block sizes of the last 5000 blocks in our database.
+For sample sizes less than 5000, blocks were broken into groups of equal size and the model was fit separately to each group.
+Sample size is indicated by $n$, group id by $g$.
+Each model was fit in RStan \citep{rstan-software:2015,stan-software:2015} using 4 chains, each with 4000 iterations, 2000 of which used for warmup/burn in.}
+%% \label{tab:gammafit}
+%% \end{table}
 
@@
 
-% latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Mon Oct 12 22:27:11 2015
+% latex table generated in R 3.2.3 by xtable 1.7-4 package
+% Mon Dec 28 10:59:52 2015
 \begin{table}[ht]
@@
   \hline
-$\gamma$ & 0.0240 & 0.0004 & 0.0000 & 0.0232 & 0.0237 & 0.0240 & 0.0243 & 0.0248 \\ 
-  $\beta$ & 15.0951 & 0.9228 & 0.0103 & 13.3725 & 14.4461 & 15.0760 & 15.6962 & 16.9591 \\ 
-  $\sigma$ & 0.0024 & 0.0001 & 0.0000 & 0.0023 & 0.0023 & 0.0024 & 0.0024 & 0.0025 \\ 
+$\gamma$ & 0.0597 & 0.0012 & 0.0000 & 0.0574 & 0.0589 & 0.0597 & 0.0605 & 0.0621 \\ 
+  $\beta$ & 4.4923 & 0.2382 & 0.0027 & 4.0328 & 4.3303 & 4.4866 & 4.6498 & 4.9660 \\ 
+  $\sigma$ & 0.0008 & 0.0000 & 0.0000 & 0.0007 & 0.0008 & 0.0008 & 0.0008 & 0.0009 \\ 
    \hline
@@
 We use RStan \citep{rstan-software:2015,stan-software:2015} to fit the model, obtaining four chains each of 4000 iterations, 2000 of which used for warmup/burn in.}
- \label{tab:modelfit}
 \end{table}
+
+% latex table generated in R 3.2.3 by xtable 1.7-4 package
+% Wed Dec 30 12:31:30 2015
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrrrrrrr}
+  \hline
+ & Mean & SD & SE(Mean) & 2.5\% & 25\% & 50\% & 75\% & 97.5\% \\ 
+  \hline
+$\gamma$ & 0.0561 & 0.0009 & 0.0000 & 0.0543 & 0.0554 & 0.0561 & 0.0566 & 0.0578 \\ 
+  $\beta$ & 7.0450 & 0.4327 & 0.0046 & 6.2235 & 6.7486 & 7.0402 & 7.3299 & 7.9080 \\ 
+  $\sigma$ & 0.0025 & 0.0001 & 0.0000 & 0.0024 & 0.0025 & 0.0025 & 0.0026 & 0.0027 \\ 
+   \hline
+\end{tabular}
+ \caption{Results of fitting the measurement error model with stochastic measurement times.
+We use RStan \citep{rstan-software:2015,stan-software:2015} to fit the model, obtaining four chains each of 4000 iterations, 2000 of which used for warmup/burn in.
+PRE 2015}
+\end{table}
+
+%%% OLD MODEL FITS TABLE
+%% % latex table generated in R 3.2.2 by xtable 1.7-4 package
+%% % Mon Oct 12 22:27:11 2015
+%% \begin{table}[ht]
+%% \centering
+%% \begin{tabular}{rrrrrrrrr}
+%%   \hline
+%%  & Mean & SD & SE(Mean) & 2.5\% & 25\% & 50\% & 75\% & 97.5\% \\ 
+%%   \hline
+%% $\gamma$ & 0.0240 & 0.0004 & 0.0000 & 0.0232 & 0.0237 & 0.0240 & 0.0243 & 0.0248 \\ 
+%%   $\beta$ & 15.0951 & 0.9228 & 0.0103 & 13.3725 & 14.4461 & 15.0760 & 15.6962 & 16.9591 \\ 
+%%   $\sigma$ & 0.0024 & 0.0001 & 0.0000 & 0.0023 & 0.0023 & 0.0024 & 0.0024 & 0.0025 \\ 
+%%    \hline
+%% \end{tabular}
+%%  \caption{Results of fitting the measurement error model with stochastic measurement times.
+We use RStan \citep{rstan-software:2015,stan-software:2015} to fit the model, obtaining four chains each of 4000 iterations, 2000 of which used for warmup/burn in.}
+%%  \label{tab:modelfit}
+%% \end{table}
 
@@
 Probabilities in the fifth and final row were computed by simulating 100 blocks for each of the 8000 draws from the posterior of $\beta$ and the shifted posterior of $\gamma$, where a draw from the shifted posterior of $\gamma$ is a draw from the posterior plus the difference between the specified $\mathrm{E}[\gamma]$ and $\gamma$'s posterior mean.
-% latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Tue Oct 13 00:53:54 2015
+
+% latex table generated in R 3.2.3 by xtable 1.7-4 package
+% Mon Dec 28 11:02:20 2015
 \begin{table}[ht]
 \centering
-\begin{tabular}{lrrrrrrrr}
+\begin{tabular}{rrrrrrrrr}
   \hline
@@
 $\beta = \infty$ & 0.0067 & 0.0357 & 0.0821 & 0.1353 & 0.1889 & 0.2397 & 0.2865 & 0.3292 \\ 
-  $\beta = 3\times\beta^{(0.5)}$ & 0.0082 & 0.0387 & 0.0853 & 0.1383 & 0.1915 & 0.2420 & 0.2874 & 0.3293 \\ 
-  $\beta = \beta^{(0.5)}$ & 0.0117 & 0.0446 & 0.0917 & 0.1435 & 0.1958 & 0.2446 & 0.2881 & 0.3300 \\ 
-  $\beta \sim p(\beta|x,t)$ & 0.0119 & 0.0452 & 0.0919 & 0.1437 & 0.1954 & 0.2438 & 0.2892 & 0.3304 \\ 
-  $(\beta,\gamma) \sim p(\beta,\gamma|x,t)$ & 0.0118 & 0.0448 & 0.0916 & 0.1433 & 0.1955 & 0.2442 & 0.2896 & 0.3301 \\ 
+  $\beta = 3\times\beta^{(0.5)}$ & 0.0126 & 0.0462 & 0.0935 & 0.1445 & 0.1954 & 0.2445 & 0.2895 & 0.
+3306 \\ 
+  $\beta = \beta^{(0.5)}$ & 0.0274 & 0.0644 & 0.1102 & 0.1579 & 0.2057 & 0.2488 & 0.2911 & 0.3287 \\ 
+  $\beta \sim p(\beta|x,t)$ & 0.0271 & 0.0641 & 0.1104 & 0.1582 & 0.2051 & 0.2509 & 0.2908 & 0.3290 \\ 
+  $(\beta,\gamma) \sim p(\beta,\gamma|x,t)$ & 0.0271 & 0.0645 & 0.1100 & 0.1579 & 0.2058 & 0.2499 & 0.2920 & 0.3290 \\ 
    \hline
@@
 In the first row the transaction rate is constant at $\hat{\gamma}$, in the second through fourth rows the transaction rate is stochastic with fixed mean $\hat{\gamma}$, and in the final row the transaction rate is stochastic and unknown with mean $\hat{\gamma}$.}
-\label{tab:probsims}
 \end{table}
+
+% latex table generated in R 3.2.3 by xtable 1.7-4 package
+% Wed Dec 30 12:32:39 2015
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrrrrrrr}
+  \hline
+$E[\gamma] = \hat{\gamma} = $ & 0.02 & 0.03 & 0.04 & 0.05 & 0.06 & 0.07 & 0.08 & 0.09 \\ 
+  \hline
+$\beta = \infty$ & 0.0067 & 0.0357 & 0.0821 & 0.1353 & 0.1889 & 0.2397 & 0.2865 & 0.3292 \\ 
+  $\beta = 3\times\beta^{(0.5)}$ & 0.0103 & 0.0423 & 0.0891 & 0.1416 & 0.1938 & 0.2440 & 0.2892 & 0.3312 \\ 
+  $\beta = \beta^{(0.5)}$ & 0.0191 & 0.0549 & 0.1016 & 0.1519 & 0.2010 & 0.2478 & 0.2906 & 0.3303 \\ 
+  $\beta \sim p(\beta|x,t)$ & 0.0191 & 0.0553 & 0.1014 & 0.1516 & 0.2017 & 0.2483 & 0.2907 & 0.3305 \\ 
+  $(\beta,\gamma) \sim p(\beta,\gamma|x,t)$ & 0.0194 & 0.0550 & 0.1016 & 0.1517 & 0.2010 & 0.2472 & 0.2901 & 0.3310 \\ 
+   \hline
+\end{tabular}
+\caption{Probability of a full block under a variety of parameter values --- $\gamma$ is the mean transaction rate and $\beta$ is a dispersion parameter so that the variance of transaction rates is $\gamma/\beta$.
+In the first row the transaction rate is constant at $\hat{\gamma}$, in the second through fourth rows the transaction rate is stochastic with fixed mean $\hat{\gamma}$, and in the final row the transaction rate is stochastic and unknown with mean $\hat{\gamma}$.
+PRE 2015}
+\end{table}
+
+%%%% OLD BLOCK FILL PROBABILITIES TABLE
+%% % latex table generated in R 3.2.2 by xtable 1.7-4 package
+%% % Tue Oct 13 00:53:54 2015
+%% \begin{table}[ht]
+%% \centering
+%% \begin{tabular}{lrrrrrrrr}
+%%   \hline
+%% $E[\gamma] = \hat{\gamma} =$ & 0.02 & 0.03 & 0.04 & 0.05 & 0.06 & 0.07 & 0.08 & 0.09 \\ 
+%%   \hline
+%% $\beta = \infty$ & 0.0067 & 0.0357 & 0.0821 & 0.1353 & 0.1889 & 0.2397 & 0.2865 & 0.3292 \\ 
+%%   $\beta = 3\times\beta^{(0.5)}$ & 0.0082 & 0.0387 & 0.0853 & 0.1383 & 0.1915 & 0.2420 & 0.2874 & 0.3293 \\ 
+%%   $\beta = \beta^{(0.5)}$ & 0.0117 & 0.0446 & 0.0917 & 0.1435 & 0.1958 & 0.2446 & 0.2881 & 0.3300 \\ 
+%%   $\beta \sim p(\beta|x,t)$ & 0.0119 & 0.0452 & 0.0919 & 0.1437 & 0.1954 & 0.2438 & 0.2892 & 0.3304 \\ 
+%%   $(\beta,\gamma) \sim p(\beta,\gamma|x,t)$ & 0.0118 & 0.0448 & 0.0916 & 0.1433 & 0.1955 & 0.2442 & 0.2896 & 0.3301 \\ 
+%%    \hline
+%% \end{tabular}
+%% \caption{Probability of a full block under a variety of parameter values --- $\gamma$ is the mean transaction rate and $\beta$ is a dispersion parameter so that the variance of transaction rates is $\gamma/\beta$.
+In the first row the transaction rate is constant at $\hat{\gamma}$, in the second through fourth rows the transaction rate is stochastic with fixed mean $\hat{\gamma}$, and in the final row the transaction rate is stochastic and unknown with mean $\hat{\gamma}$.}
+%% \label{tab:probsims}
+%% \end{table}
 

### 2017-01-17 model/overflowmcmcC.cpp
pushing some bug fixes from awhile ago

@@ -55,7 +55,7 @@ List overtrendmcmcCPP(int niter, arma::vec tobs, arma::vec D, arma::uvec M, arma
 			 double rwc, int H, int tune,
 			 arma::vec lowtarget, arma::vec hightarget, arma::mat tauchol,
 			 Function rtruncnorm, Function rktnb, Function rbetarejC){
-  Rprintf("Initializing 1");
+  //Rprintf("  Initializing 1");
   int nblocks = tau.size();
   arma::vec tau0full(nblocks + 1); tau0full(0) = 0; tau0full.subvec(1, nblocks) = tau;
   arma::vec tau0 = tau0full.subvec(0, nblocks - 1);
@@ -69,9 +69,9 @@ List overtrendmcmcCPP(int niter, arma::vec tobs, arma::vec D, arma::uvec M, arma
   arma::mat taurejs(niter, 1);
   arma::mat gamrejs(niter, 2);
   arma::mat alpharejs(niter, 1);
-  Rprintf(" 2");  
+  //Rprintf("      2");  
   double aomega = (vsigma + 1)/2;
-  double bomega = 1/(vsigma*pow(ssigma, 2.0));
+  double bomega = 1/(2*vsigma*pow(ssigma, 2.0));
   double asigma;
   double con2 = aalpha - 1;
   arma::uvec nds = diff(M);  //////////////////////
@@ -79,7 +79,7 @@ List overtrendmcmcCPP(int niter, arma::vec tobs, arma::vec D, arma::uvec M, arma
   arma::uvec nrej(nblocks);
   arma::vec rho(nblocks);
   arma::uvec nrejcumsum(nblocks);
-  Rprintf(" 3"); 
+  //Rprintf("      3"); 
   double bsigma;
   double omega;
   double sigma2;
@@ -94,7 +94,7 @@ List overtrendmcmcCPP(int niter, arma::vec tobs, arma::vec D, arma::uvec M, arma
   double gamprop;
   arma::vec gamtemp(2);
   double lanum;
-  Rprintf(" 4"); 
+  //Rprintf("      4"); 
   double ladenom;
   int Nover;
   double lowlim;
@@ -103,23 +103,23 @@ List overtrendmcmcCPP(int niter, arma::vec tobs, arma::vec D, arma::uvec M, arma
   int lower;
   int upper;
   arma::vec dk1(nblocks);
[…]

### 2017-08-30 status.tex
wrote a quick thing up on the status of the project

\documentclass[12pt]{article}

\setlength{\oddsidemargin}{-0.125in}
\setlength{\topmargin}{-0.5in} \setlength{\textwidth}{6.5in}
\setlength{\textheight}{9in}
\usepackage{rotating}
\setlength{\textheight}{9in} \setlength{\textwidth}{6.5in}
\setlength{\topmargin}{-40pt} \setlength{\oddsidemargin}{0pt}
\setlength{\evensidemargin}{0pt}
\usepackage{adjustbox,lipsum}
\setlength{\textheight}{8.5in} \setlength{\textwidth}{6.5in}
\setlength{\topmargin}{-36pt} \setlength{\oddsidemargin}{0pt}
\setlength{\evensidemargin}{0pt} \tolerance=500
\renewcommand{\baselinestretch}{1.5}

\usepackage{amsmath, amsfonts}
\usepackage{bm, bbm}
\usepackage{graphics, graphicx, color, epsf}
\usepackage[authoryear]{natbib}
\usepackage{xr, zref, hyperref}

\begin{document}
\section{Introduction}
This section describes the current status of the project. In particular, is states the two main models we have and discusses limitations and other problems of both.

\section{The Block Size Model}
The block size model is the original model we developed. I present a slightly different version of it here with a different measurement error model and a more general process model. Let $k=1,2,\dots,K$ denote blocks in order of arrival on the block chain. Then we model two key quantities: the arrival times of the each block in minutes, denoted by $\tau_k$, and the size of the block in MB, denoted by $d_k$. We directly observe $d_k$, but not $\tau_k$. Rather, we observe $t_k$, a noisy measurement of $\tau_k$ that is restricted to be larger than the median of $\bm{t}_{k-1:11} = (t_{k-11}, t_{k-10}, \dots, t_{k-1})$, which we denote by $m_k$. The measurement error is due to the fact that different computers mine different blocks, and they may not not have perfectly synced internal clocks. As a result, the block chain protocol rejects all blocks that have a reported arrival time that is too early.

In the arrival time measurement error model, we assume that the observed arrival times are normally distributed, centered on the true arrival time, but truncated from below at $m_k$. That is
\begin{align}
t_k | \bm{t}_{k-1:11}, \tau_k, \sigma^2 &\stackrel{ind}{\sim}\mathrm{N}(\tau_k, \sigma^2)1(t_k>m_k) && \mbox{(arrival time measurement model)} \label{eq:error}\\
\intertext{for $k=1,2,\dots,K$. Next, define the block inter-arrival times, $\delta_k = \tau_k - \tau_{k-1}$ for $k=1,2,\dots,K$ where $\tau_0\equiv 0$. The bitcoin protocol is designed so that the blocks follow a Poisson process with intensity $1/10$ per minute. This implies that the inter-arrival times are exponentially distribution with with mean $10$ minutes, i.e.}
\delta_k = \tau_k - \tau_{k-1} &\stackrel{iid}{\sim}\mathrm{Exp}(1/10) && \mbox{(inter-arrival time process model)}\label{eq:arrival}
\end{align}
for $k=1,2,\dots,K$.

Next, we need a model for the block sizes, i.e. the $d_k$s. The longer it takes for a block to arrive, the more transactions and therefore the more transaction data it should accumulate. Let $D(\tau)$ denote the total accumulated transaction data in the block chain. Then $D(\tau)$ we assume that $D(\tau)$ is a nondecreasing L\'{e}vy process, i.e. that $\{D(\tau): \tau > 0\}$ is continuous in $\tau$ with independent, stationary increments. There are many options here, but a convenient L\'{e}vy process is the gamma process, defined by
\begin{align*}
D(\tau + \delta) - D(\tau) &\sim \mathrm{G}(\Psi(\tau + \delta) - \Psi(\tau), \theta)
\end{align*}
for any time $\tau>0$ and increment $\delta > 0$ where $\Psi(\tau) = \int_0^\tau\psi(u)du$, and $\psi(u)>0$ is the intensity function of the  process. We use the shape/rate parameterization of the gamma distribution so that $\mathrm{E}[D(\tau + \delta) - D(\tau)] = [\Psi(\tau + \delta) - \Psi(\tau)]/\theta$. To fully specify this process model we need to specify the intensity function, but for now we will leave that aside. Given the gamma process for $D(\tau)$, we can define the block size process model. Let $D_k = D(\tau_k)$, $d_k = D_k - D_{k-1}$, and $\psi_k = \Psi(\tau_k)$ for $k=1,2,\dots,K$ where we set $D_0 = D(0) = 0$. Then we assume
\begin{align}
d_k | \psi_k, \theta &\stackrel{ind}{\sim}\mathrm{G}(\theta\psi_k, \theta) && \mbox{(block size process model)}\label{eq:size}
\end{align}
for $k=1,2,\dots,K$. We parameterize the intensity function and the gamma process so that $\mathrm{E}[d_k|\psi_k,\theta] = \psi_k$, so that the intensity function defines the mean of $d_k$'s gamma distribution.

Equations \eqref{eq:error} - \eqref{eq:size} define the block size model. To complete this model we must specify the intensity function, $\psi(u)$, and priors for all of the unknown parameters. In the original version of the paper we only considered a constant intensity $\psi(u) = \gamma > 0$, though we can also define an increasing intensity $\psi(u) = e^{\beta_0 + u\beta_1}$. In principle we can define any intensity function of the form $\psi(u) = e^{\bm{z}(u)'\bm{\beta}}$ where $\bm{z}(u)$ is a vector of basis functions of time, and $\bm{\beta}$ is a vector of regression coefficients. However, we need to be able to compute the integral $\int_{\tau}^{\tau + \delta}\psi(u)du$ in closed form, which restricts the form of $\psi(u)$. One possibility that I have not explored that may be more flexible is $\psi(u) = \sum_{p=1}^P\alpha_pu^{\beta_p - 1}$ with $\alpha_p, \beta_p>0$ for $p=1,2,\dots,P$.

\subsection{Problems and limitations}
The big problem with this model, which we identified after we got feedback on the first draft, is that the block size process model does not take into account that excess transactions that arrive during one block spill over into the next block. So all observed block sizes will be lower than the block size limit for that particular block. For example is every block has a 1 MB block size limit, we would never observe a block with more than 1 MB of transactions, even when the limit is routinely being reach. What $d_k$ represents is not the total amount of transaction data that was {\it submitted} to the block chain after block $k-1$ arrived but before block $k$, instead it represents the total amount of transaction data {\it accepted} by the block chain in block $k$.

There is information in the observed data about the spill over. When a block's size is close to the cap, then we know there's a decent chance that at least one transaction was pushed into the next block. There are a couple of problems with this, however: 1) diffwerent bitcoin miners implement different caps, and 2) transaction data comes in discrete chunks. As a consuquence of 2), a block's size can be strictly smaller than the cap and still have spill over transactions, so determining whether a block hit the cap is not as straightforward as seeing whether the block size is the same as the cap size. Further, because of 1) we do not know the cap except for a maximum value, e.g. 1 MB throughout much of bitcoin's history.

We tried to solve this problem with the second model we built. I won't recapitulate it here, but basically we replaced Equation \eqref{eq:size} with 1) a transaction arrival stochastic process, 2) a transaction size model, and 3) incorprated the block size constraints to allow for spillover. This model ended up being very messy and as a result difficult to fit --- I had to write my own MCMC algorithm to fit it. Plus, it seemed to drastically overestimate the probability that a given block has spillover. Because of this, I don't think we should pursue this model.

As an alternative, what we could do is stick with the model we have above and fit a separate model using block sizes (in MB) and block transaction counts in order to estimate the size of a typical transaction. Then we can compare the distribution of transaction sizes to the difference between the oberved block sizes and the cap in order to predict when the cap has been reached. We can do the same thing with the {\it predicted} block sizes, which we can obtain from the posterior predictive distribution of the block size model. This is substantially easier to do.
\end{document}

### 2017-08-30 status.tex
added some TODOs

@@
 \section{Introduction}
-This section describes the current status of the project. In particular, is states the two main models we have and discusses limitations and other problems of both.
+This note describes the current status of the project.
+In particular, is states the first model we used and discusses its limitations.
+Then it discusses what to do about those limitations, including whether we should use the second model.
 
@@
 for $k=1,2,\dots,K$.
-We parameterize the intensity function and the gamma process so that $\mathrm{E}[d_k|\psi_k,\theta] = \psi_k$, so that the intensity function defines the mean of $d_k$'s gamma distribution.
+We parameterize the intensity function and the gamma process so that $\mathrm{E}[d_k|\psi_k,\theta] = \psi_k$, so that the intensity function is $\theta\psi(u)$ and $\psi(u)$ defines the mean of $d_k$'s gamma distribution.
 
@@
 However, we need to be able to compute the integral $\int_{\tau}^{\tau + \delta}\psi(u)du$ in closed form, which restricts the form of $\psi(u)$.
-One possibility that I have not explored that may be more flexible is $\psi(u) = \sum_{p=1}^P\alpha_pu^{\beta_p - 1}$ with $\alpha_p, \beta_p>0$ for $p=1,2,\dots,P$.
+One possibility that I have not explored that may be more flexible is $\psi(u) = \sum_{p=1}^P\alpha_pu^{\beta_p - 1}$ with $\alpha_p, \beta_p\geq 0$ for $p=1,2,\dots,P$.
+In this case, then $\Psi(\tau) = \int_0^\tau\psi(u)du = \alpha_p\tau^{\beta_p}/\beta_p$ where $\tau^0/0$ is defined to be $\log(\tau)$.
+If $\beta_p < 1$ then the $p$th term of $\psi(u)$ decays in $u$, while if $\beta_p > 1$ it increases in $u$ and if $\beta_p = 1$ it is constant in $u$.
 
@@
 This is substantially easier to do.
+
+The transaction data model might look like this.
+Suppose $x_i$ denotes the amount of data for transaction $i$, then we assume $x_i \stackrel{iid}{\sim}\mathrm{G}(\alpha, \beta)$.
+But we do not observed the $x_i$s.
+Instead we observe the $d_k$, the amount of data in block $k$, and $n_k$, the number of transactions in block $k$.
+The model on the $x_i$s implies then that $d_k \stackrel{ind}{\sim} \mathrm{G}(\alpha n_k, \beta)$.
+So we can fit this model and get parameter estimates for $\alpha$ and $\beta$, and in particular compute $P(x_i > x)$ --- the probability that a given transaction is size $x$ or large for any $x$.
+When $x$ is the difference between the block size cap and an observed block size, then this is an estimate of the probability that there was block overflow.
+
+\section{What to do next}
+In short, this is what I want to do in order to get this paper into a form that we can submit:
+\begin{enumerate}
+\item Get updated data from the block chain. (Can we easily do this?)
+\item Fit transaction size models.
+The model I mentioned in the last paragraph of the previous section assumes that transactions {\it sizes} have been constant, on average, over time, and any differences in block sizes is due to the number of transactions.
+We can investigate this pretty easily.
+\item Fit several versions of the block size model --- basically with different choices of $\psi(u)$. 
+\item Compare predictions of the block size model to reality --- do some one step ahead predictions to see how well we predict block sizes.
+We can do this in different regions of the blockchain --- e.g.
+with the old data we have, and also with up-to-the-minute data.
+\item Use 1) and 2) in order to estimate the probability of transaction spillover in previous blocks, and new blocks. 
+\item Write everything up.
+\end{enumerate}
+
+Getting these models implemented and fit shouldn't be much work for me.
+Everything will be in Stan, so it'll mostly be a matter of pressing buttons.
+The most time consuming part will be writing everything up, including getting references to the relevant literature.
 \end{document}

### 2017-08-30 status.tex
tweaked a section header

@@
 
-\subsection{Problems and limitations}
+\subsection{Problems and solutions}
 The big problem with this model, which we identified after we got feedback on the first draft, is that the block size process model does not take into account that excess transactions that arrive during one block spill over into the next block.

### 2017-08-30 model/old/overflowmcmc.tex
moved old model stuff to an old folder

--- 2016-02-18 model/overflowmcmc.tex
@@
 The bulk of this material may end up in a technical appendix or as a section of the paper.
-Section \ref{sec:model} briefly describes the model and section \ref{sec:mcmc} describes the MCMC strategy for computing the posterior distribution.
+Section \ref{sec:model} briefly describes the model and Section \ref{sec:mcmc} describes the MCMC strategy for computing the posterior distribution.
 
@@
 We will assume that transactions arrive according to a non-homogenous Poisson process.
-Let $N(\tau)$ denote the number of transactions that have arrive by time $\tau$, and let $\lambda(\tau)$ denote some intensity function such that $\lambda(\tau) > 0$ for all times $\tau$.
+Let $N(\tau)$ denote the number of transactions that have arrived by time $\tau$, and let $\lambda(\tau)$ denote some intensity function such that $\lambda(\tau) > 0$ for all times $\tau$.
 Then any collection of non-overlapping increments $\left\{N(\tau_i + \delta_i) - N(\tau_i)\right\}$ where $\delta_i>0$ for all $i$ are independent with distribution
@@
 \begin{align*}
-P(\eta_k = \eta|\delta_{1:n_{b}}) \frac{\Gamma(\psi_k + \eta)}{\Gamma(\eta + 1)}\phi^{\psi_k}(1-\phi)^{\eta}
+P(\eta_k = \eta|\delta_{1:n_{b}}) = \frac{\Gamma(\psi_k + \eta)}{\Gamma(\psi_k)\Gamma(\eta + 1)}\phi^{\psi_k}(1-\phi)^{\eta}
 \end{align*}
@@
 \begin{align*}
-\sigma^2|\omega &\sim IG(v/2, \omega); && \omega \sim G(1/2, s^2/2).
+\sigma^2|\omega &\sim IG(v/2, \omega); && \omega \sim G(1/2, 1/\{2vs^2\}).
 \end{align*}
@@
 \begin{align*}
-\bm{\beta} \sim N(\bm{g}_\gamma, \bm{S}_\gamma)
+\bm{\gamma} \sim N(\bm{g}_\gamma, \bm{S}_\gamma)
 \end{align*}
@@
 \subsection{Priors on $\alpha$ and $\beta$}
-The conditionally conjugate prior for $beta$ is a gamma while $\alpha$ has no convenient conditionally conjugate form.
+The conditionally conjugate prior for $\beta$ is a gamma while $\alpha$ has no convenient conditionally conjugate form.
 We will assume that $\beta \sim G(a_\beta, b_\beta)$ and $\alpha \sim G(a_\alpha, b_\alpha)$.
@@
 We use data augmentation in two ways.
-First, in order to more effectively deal with the truncated distribution in the measurement error model for block arrival times we introduce $\tilde{\bm{t}}_k=(\tilde{t}_{1k},\dots,\tilde{t}_{\tilde{n}_kk}, t_k)$ as the original $t_k$ along with $\tilde{n}_k$ draws from the untruncated normal distribution for $t_k$ that occurred before $t_k$ was drawn from the restricted support.
-Second, the models for $\eta_{1:n_{b}}$ and $d_{1:n_d}$ induce a model for $(M_{1:n_{b}},D_{1:n_d})$ since the latter are a detrministic function of the former, but this model is not easy work with.
+First, in order to more effectively deal with the truncated distribution in the measurement error model for block arrival times we introduce $\widetilde{\bm{t}}_k=(\widetilde{t}_{1k},\dots,\widetilde{t}_{\widetilde{n}_kk}, t_k)$ as the original $t_k$ along with $\widetilde{n}_k$ draws from the untruncated normal distribution for $t_k$ that occurred before $t_k$ was drawn from the restricted support.
+Second, the models for $\eta_{1:n_{b}}$ and $d_{1:n_d}$ induce a model for $(M_{1:n_{b}},D_{1:n_d})$ since the latter are a deterministic function of the former, but this model is not easy work with.
 Instead, we explicitly use $\eta_{1:n_{b}}$ and $d_{1:n_d}$ in the Gibbs sampler as additional data augmentation.
 
-\subsection{Data augmentation for truncated distributions --- $\tilde{n}_k$ and $\tilde{t}_k$}
+\subsection{Data augmentation for truncated distributions --- $\widetilde{n}_k$ and $\widetilde{t}_k$}
 For block $k=1,2,n_{b}$ the measurement error model on block arrival times is
@@
 until it successfully obtains a $t_k > m_k$.
-Let $\tilde{n}_k$ be the number of failures before the observed success $t_k$.
-Then $\tilde{n}_k$ has a negative binomial distribution, specifically $\tilde{n}_k \sim NB(1, \rho_k)$ with density
+Let $\widetilde{n}_k$ be the number of failures before the observed success $t_k$.
+Then $\widetilde{n}_k$ has a negative binomial distribution, specifically $\widetilde{n}_k \sim NB(1, \rho_k)$ with density
 \begin{align*}
-p(\tilde{n}_k|\tau_k,\sigma^2) = \rho_k^{\tilde{n}_k}(1 - \rho_k)
+p(\widetilde{n}_k|\tau_k,\sigma^2) = \rho_k^{\widetilde{n}_k}(1 - \rho_k)
 \end{align*}
@@
 $\rho_k = \Phi([m_k - \tau_k]/\sigma)$.
-Conditional on $\tilde{n}_k$ the density of the $\tilde{n}_k$ failures, $\tilde{t}_{1k},\dots,\tilde{t}_{\tilde{n}_kk}$, is again truncated normal, but this time truncated to the opposite end of the parameter space.
-That is for $i = 1,\dots,\tilde{n}_k$
+Conditional on $\widetilde{n}_k$ the density of the $\widetilde{n}_k$ failures, $\widetilde{t}_{1k},\dots,\widetilde{t}_{\widetilde{n}_kk}$, is again truncated normal, but this time truncated to the opposite end of the parameter space.
+That is for $i = 1,\dots,\widetilde{n}_k$
 \begin{align*}
-\tilde{t}_{ik}\stackrel{iid}{\sim}N(\tau_k,\sigma^2)1(t_k < m_k).
+\widetilde{t}_{ik}\stackrel{iid}{\sim}N(\tau_k,\sigma^2)1(t_k < m_k).
 \end{align*}
@@
 \begin{enumerate}
-\item Draw $\tilde{n}_k \sim NB(1,\rho_k)$ where $\rho_k = \Phi([ m_k -  \tau_k]/\sigma)$.
-\item If $\tilde{n}_k > 0$, for $i=1,\dots,\tilde{n}_k$ draw $\tilde{t}_{ik} \sim N(\tau_k,\sigma^2)1(t_k < m_k)$ and form $\tilde{\bm{t}}_k = (\tilde{t}_{1k},\dots,\tilde{t}_{\tilde{n}_kk}, t_k)'$.
+\item Draw $\widetilde{n}_k \sim NB(1,\rho_k)$ where $\rho_k = \Phi([ m_k -  \tau_k]/\sigma)$.
+\item If $\widetilde{n}_k > 0$, for $i=1,\dots,\widetilde{n}_k$ draw $\widetilde{t}_{ik} \sim N(\tau_k,\sigma^2)1(t_k < m_k)$ and form $\widetilde{\bm{t}}_k = (\widetilde{t}_{1k},\dots,\widetilde{t}_{\widetilde{n}_kk}, t_k)'$.
 \end{enumerate}
-For simplicity in the other Gibbs steps define $\tilde{t}_{\tilde{n}_k + 1,k}\equiv t_k$.
+For simplicity in the other Gibbs steps define $\widetilde{t}_{\widetilde{n}_k + 1,k}\equiv t_k$.
 
-It is tempting to imbue $\tilde{n}_k$ and $\tilde{t_{ik}}$ with some interpretation relative to the model, e.g.
+It is tempting to imbue $\widetilde{n}_k$ and $\widetilde{t}_{ik}$ with some interpretation relative to the model, e.g.
 as in [CITE GELMAN PAPER].
-For example, $\tilde{n}_k$ may represent the number of blocks that were rejected by the Bitcoin protocol for having a timestamp earlier than the median of the previous elevent blocks.
+For example, $\widetilde{n}_k$ may represent the number of blocks that were rejected by the Bitcoin protocol for having a timestamp earlier than the median of the previous elevent blocks.
 While tempting this interpretation is not strictly speaking correct.
@@
 \subsection{Measurement error variance --- $\sigma^2$}
-The full conditional distribution of the measurement error variance, $\sigma^2$, depends only on $\tilde{\bm{t}}_{1:n_{b}}$ and $\tau_{1:n_{b}}$. It is
+The full conditional distribution of the measurement error variance, $\sigma^2$, depends only on $\widetilde{\bm{t}}_{1:n_{b}}$ and $\tau_{1:n_{b}}$. It is
 \begin{align*}
-p(\sigma^2|...) \propto \sigma^{-(n_{b} + \sum_{k=1}^{n_{b}}\tilde{n}_k)}\exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\tilde{t}_{ik} - \tau_k)^2\right]p(\sigma^2)
+p(\sigma^2|...) \propto \sigma^{-(n_{b} + \sum_{k=1}^{n_{b}}\widetilde{n}_k)}\exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\widetilde{n}_k}(\widetilde{t}_{ik} - \tau_k)^2\right]p(\sigma^2)
 \end{align*}
 where $p(\sigma^2)$ is the density of the prior distribution on $\sigma^2$.
-With the half-$t$ prior on $\sigma$ discussed above and letting $a_\sigma = (n_{b} + \sum_{k=1}^{n_b}\tilde{n}_k)/2$ and $b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\tilde{t}_{ik} - \tau_k)^2$ the full conditional distribution of $(\sigma^2,\omega)$ is
+With the half-$t$ prior on $\sigma$ discussed above and letting $a_\sigma = (n_{b} + \sum_{k=1}^{n_b}\widetilde{n}_k)$ and $b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\widetilde{n}_k}(\widetilde{t}_{ik} - \tau_k)^2$ the full conditional distribution of $(\sigma^2,\omega)$ is
 \begin{align*}
-p(\sigma^2,\omega|...) \propto (\sigma^2)^{-(a_\sigma+v)/2-1}\exp\left[-\frac{b_\sigma}{\sigma^2}\right]\exp\left[-\frac{\omega}{\sigma^2}\right]\omega^{(v + 1)/2 - 1}\exp\left[-\frac{\omega}{v s^2}\right].
+p(\sigma^2,\omega|...) \propto (\sigma^2)^{-(a_\sigma+v)/2-1}\exp\left[-\frac{b_\sigma}{\sigma^2}\right]\exp\left[-\frac{\omega}{\sigma^2}\right]\omega^{(v + 1)/2 - 1}\exp\left[-\frac{\omega}{2 v s^2}\right].
 \end{align*}
@@
 \begin{enumerate}
-\item Draw $\omega \sim G\left(\frac{v + 1}{2}, \frac{1}{\sigma^2} + \frac{1}{v s^2}\right)$ and form\footnote{The inconsistency in notation between the definition of $a_\sigma$ and $b_\sigma$ is a consequence of matching notation to the R code supplied for implementing this MCMC algorithm.}
+\item Draw $\omega \sim G\left(\frac{v + 1}{2}, \frac{1}{\sigma^2} + \frac{1}{v s^2}\right)$ and form
 \begin{align*}
-a_\sigma = \frac{n_{b} + \sum_{k=1}^{n_b}\tilde{n}_k + v}{2} &&\mbox{  and  }&& b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\tilde{t}_{ik} - \tau_k)^2;
+a_\sigma = v + n_{b} + \sum_{k=1}^{n_b}\widetilde{n}_k &&\mbox{  and  }&& b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\widetilde{n}_k}(\widetilde{t}_{ik} - \tau_k)^2;
 \end{align*}
-\item Draw $\sigma^2 \sim  IG\left(a_\sigma, b_\sigma/2 + \omega\right)$.
+\item Draw $\sigma^2 \sim  IG\left(a_\sigma/2, b_\sigma/2 + \omega\right)$.
 \end{enumerate}
@@
 \begin{align*}
-p(\bm{\tau}_{1:n_{b}}|...) \propto & \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \tau_k\right)^2 - \frac{1}{10}\tau_{n_{b}}\right]\\
+p(\bm{\tau}_{1:n_{b}}|...) \propto & \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\widetilde{n}_k+1}\left(\widetilde{t}_{ik} - \tau_k\right)^2 - \frac{1}{10}\tau_{n_{b}}\right]\\
 &\times \prod_{k=1}^{n_b}\Gamma[\psi_k(\tau_{k}, \tau_{k-1}) + \eta_k]\phi^{\psi_k(\tau_k,\tau_{k-1})}1(0\le \tau_1 \le \tau_2 \le \hdots \le \tau_{n_b})
@@
 \begin{align*}
-p(\bm{\delta}_{1:n_{b}}|...) \propto & \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \sum_{l=1}^k\delta_l\right)^2 - \frac{1}{10}\sum_{k=1}^{n_b}\delta_{k}\right]\\
+p(\bm{\delta}_{1:n_{b}}|...) \propto & \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\widetilde{n}_k+1}\left(\widetilde{t}_{ik} - \sum_{l=1}^k\delta_l\right)^2 - \frac{1}{10}\sum_{k=1}^{n_b}\delta_{k}\right]\\
 &\times \prod_{k=1}^{n_b}\Gamma(\psi_k + \eta_k)\phi^{\psi_k}1(0\le \delta_k).
@@
 \begin{align*}
-p(\bm{\zeta}_{1:n_{b}}|...) \propto & \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \sum_{l=1}^ke^{\zeta_l}\right)^2 - \frac{1}{10}\sum_{k=1}^{n_b}e^{\zeta_k} + \sum_{k=1}^{n_b}{\zeta_k}\right] \prod_{k=1}^{n_b}\Gamma(\psi_k + \eta_k)\phi^{\psi_k}.
+p(\bm{\zeta}_{1:n_{b}}|...) \propto & \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\widetilde{n}_k+1}\left(\widetilde{t}_{ik} - \sum_{l=1}^ke^{\zeta_l}\right)^2 - \frac{1}{10}\sum_{k=1}^{n_b}e^{\zeta_k} + \sum_{k=1}^{n_b}{\zeta_k}\right] \prod_{k=1}^{n_b}\Gamma(\psi_k + \eta_k)\phi^{\psi_k}.
 \end{align*}
@@
 \item Draw $u \sim U(0,1)$ and set $o_{1k} = F^{-1}(u*[1 - F((\bar{D}_{k-1} - D_{k-1})/D_k)])$ where $F$ is the cdf of the $Beta(\alpha, \alpha(M_k - M_{k-1} - 1))$ distribution.
-\item Draw $\tilde{\bm{o}}_{2:(M_k - M_{k-1})}\sim Dir(\alpha\bm{1}_{M_k - M_{k-1} - 1})$ and set $o_{ik} = \tilde{o}_i$ for $i = 2,3,\dots,M_k - M_{k-1}$.
+\item Draw $\widetilde{\bm{o}}_{2:(M_k - M_{k-1})}\sim Dir(\alpha\bm{1}_{M_k - M_{k-1} - 1})$ and set $o_{ik} = \widetilde{o}_i$ for $i = 2,3,\dots,M_k - M_{k-1}$.
 \end{enumerate}

### 2017-08-30 status.tex
added a caveat that the exponent should be greater than zero and changed notation

@@
 However, we need to be able to compute the integral $\int_{\tau}^{\tau + \delta}\psi(u)du$ in closed form, which restricts the form of $\psi(u)$.
-One possibility that I have not explored that may be more flexible is $\psi(u) = \sum_{p=1}^P\alpha_pu^{\beta_p - 1}$ with $\alpha_p, \beta_p\geq 0$ for $p=1,2,\dots,P$.
-In this case, then $\Psi(\tau) = \int_0^\tau\psi(u)du = \alpha_p\tau^{\beta_p}/\beta_p$ where $\tau^0/0$ is defined to be $\log(\tau)$.
-If $\beta_p < 1$ then the $p$th term of $\psi(u)$ decays in $u$, while if $\beta_p > 1$ it increases in $u$ and if $\beta_p = 1$ it is constant in $u$.
+One possibility that I have not explored that may be more flexible is $\psi(u) = \sum_{p=1}^P\beta_pu^{\alpha_p - 1}$ with $\alpha_p, \beta_p\geq 0$ for $p=1,2,\dots,P$.
+In this case, then $\Psi(\tau) = \int_0^\tau\psi(u)du = \beta_p\tau^{\alpha_p}/\beta_p$.
+If $\alpha_p < 1$ then the $p$th term of $\psi(u)$ decays in $u$, while if $\alpha_p > 1$ it increases in $u$ and if $\alpha_p = 1$ it is constant in $u$.
+Strictly speaking we require $\alpha_p > 0$ since if $\alpha_p = 0$ then $\Psi(0) = -\infty$, while if $\alpha_p > 0$ then $\Psi(0) = 0$, which is what we require.
 
